---
title: "Design: Adaptive Linking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design: Adaptive Linking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article describes the algorithm currently implemented by `pairwiseLLM` for placing items
from separately ranked sets onto one common scale. For a runnable, wrapper-first tutorial, begin
with [Guide: Adaptive Linking](adaptive-linking.html). The within-set algorithm that produces
the required inputs is documented separately in
[Design: Adaptive Pairing](within-set-adaptive-design.html).

The public entry point is `adaptive_rank()` with `run_mode = "link_one_spoke"` or
`run_mode = "link_multi_spoke"`. Linking has two phases:

1. **Phase A** runs or imports a completed within-set analysis for every set.
2. **Phase B** selects comparisons between a designated hub and one or more spokes, then estimates
   spoke items on the fixed hub scale.

The current linking contract uses anchored-joint estimation and globally shared judge parameters.
It also uses a hard-locked hub, concurrent spokes, and a fixed probe effort per refit. 

## Foundational concepts

Pairwise comparative judgment (CJ) asks a judge to choose between two items. A
Bradley--Terry--Luce (BTL) model represents each item by a latent location: the modeled probability
of choosing one item over another depends on their location difference. The scale is latent because
it has no external unit. Within-set comparisons identify relative locations inside one connected
set, but two separately fitted sets can have unrelated origins and therefore cannot be compared
directly.

**Linking** collects comparisons across sets to put their locations onto a common scale. The
implemented design is a hub-and-spoke system:

- the **hub** is the reference set whose Phase A locations define the common scale;
- a **spoke** is any other set whose items are estimated on that scale;
- a **hub anchor** is a hub item selected from a region of the current ranking for recurring
  cross-set comparisons; it is not a known standard or a fixed true score;
- an **active pair** is a hub--spoke pair selected to update the linking fit; and
- a **probe pair** is a reserved hub--spoke pair used to assess predictions and stopping.

The Phase A Bayesian fits use Markov chain Monte Carlo (MCMC) draws. An expected a posteriori (EAP)
estimate is the arithmetic mean of an item's retained posterior draws, and its posterior
standard deviation (SD) summarizes model-based uncertainty. Effective sample size (ESS) and
split rank-normalized \(\widehat R\), usually called R-hat, diagnose MCMC behavior. Phase B uses
maximum a posteriori (MAP) optimization for the spoke locations and a Laplace approximation based
on the local Hessian matrix for uncertainty. MCMC diagnostics therefore apply to Phase A fits;
Phase B has a separate deterministic-fit diagnostic contract.

## Glossary and units

| Term | Meaning |
|---|---|
| set | A collection ranked internally during Phase A and identified by `set_id`. |
| artifact | The canonical, validated Phase A record for one set. |
| hub scale | The numerical scale defined by the hub's fixed Phase A item locations. |
| spoke epoch | A period during which one spoke's linking contract and probe panel remain unchanged. |
| linking refit | A Phase B update using Phase A evidence and committed active cross-set evidence. |
| refit window | The comparisons allocated between two linking refits. |
| routing stratum | A rank region used to form anchor, long, mid, and local candidate domains. |
| Fisher information | Local curvature measuring how strongly a modeled observation informs free parameters. |
| D-optimal gain | Increase in the regularized log determinant of accumulated Fisher information. |
| frozen spoke | A spoke that passed the implemented rolling stop rule and receives no further pairs. |
| blocker | A named unmet condition recorded in `stop_blocker_codes`. |

The unit of selection is an unordered hub--spoke pair. The unit of judgment is its ordered
presentation as item \(A\) and item \(B\). Active comparison counts, probe comparison counts, and
attempted steps are distinct. An invalid judgment consumes an attempted step but is not committed
to either evidence stream.

## Algorithm overview

```text
validate the combined item table and linking configuration
  -> Phase A: import or run each required within-set analysis
  -> validate and freeze the canonical artifacts
  -> pool Phase A evidence to estimate global judge parameters
  -> initialize every ready spoke on the fixed hub scale
  -> while at least one spoke remains active:
       reserve or restore that spoke's probe panel
       allocate active and probe effort for the refit window
       select active pairs by routing domain and D-optimal gain
       collect no more than the fixed per-refit probe cap
       refit every eligible spoke, potentially in parallel
       compute diagnostics, calibration, stability, and blockers
       freeze each spoke whose rolling stop rule passes
  -> return common-scale item summaries and canonical logs
```

One controller advances all active spokes. "Concurrent" means that their work is interleaved and
their eligible refits can be dispatched together; a single judgment still compares one hub item
with one item from exactly one spoke. Spoke--spoke comparisons are not part of the public design.

## Phase A artifacts and validation

Phase A produces the within-set evidence needed before any cross-set comparison is scheduled. A
canonical artifact stores:

- `set_id`, `n_items`, `n_pairs_committed`, and the source refit, round, step, and phase scope;
- an item table containing `item_id`, `global_item_id`, `theta_raw_mean`, `theta_raw_sd`, and
  `rank_mu_raw`;
- the normalized BTL model identifier, its fit-contract surface, and `fit_config_hash`;
- within-set reliability, rank agreement, and diagnostic status;
- the exact committed within-set edge table and its content hash;
- posterior item draws when available; and
- judge-parameter mode and creation time.

`item_id` is the row identity used by the adaptive state. `global_item_id` is the stable identity
used to validate an item across separately produced inputs. Both mappings, the item count, all
required summaries, and the exact within-set edge domain must agree with the current run.

The fit-contract hash covers the normalized `model_variant`, judge-parameter mode, and relevant
Bayesian fitting settings. A `btl` artifact is therefore not silently converted to `btl_e_b`, and
an artifact produced under an incompatible configuration is rejected. This protects the likelihood
and prior assumptions used by the linking fit; it is stricter than checking that an `.rds` file can
be read.

### Run, import, and mixed behavior

`phase_a_mode` assigns a source to every required set:

| Mode | Source assignment | Failure behavior |
|---|---|---|
| `run` | Run Phase A for every set, in ascending `set_id` order. | Unfinished sets remain `pending_finalization`. |
| `import` | Import every set from `phase_a_artifacts`. | A missing or incompatible artifact fails before Phase B. |
| `mixed` | Import supplied sets; run the rest. | Imports are strict; unfinished runs remain pending. |

The hub must be ready before a spoke can enter Phase B. In a multi-spoke run, all required Phase A
work is completed before Phase B begins. `out$phase_a$set_status` records `source`, `status`, and
`validation_message`; `ready`, `pending_finalization`, and `failed` describe artifact readiness,
not the later linking result.

An imported artifact normally needs finite EAP reliability at or above
`phase_a_required_reliability_min`. `quality_gate_accepted = TRUE` is a trusted external-review
override for missing or low reliability. It does not change the evidence, repair MCMC diagnostics,
or make an incompatible model variant compatible.

## Common-scale identification

Within a BTL likelihood, adding the same constant to every item location leaves all pairwise
differences unchanged. Independently fitted sets therefore do not share an origin. The linking
model resolves this location indeterminacy by treating the hub's Phase A locations as fixed
constants and estimating spoke locations directly against them.

Let \(h\in\mathcal H\) index hub items and \(x\in\mathcal S_j\) index items in spoke \(j\). The
implementation stores the fixed hub values as `theta_hub_fixed` and the estimated spoke values as
`theta_spoke_global_mean`. For an ordered cross-set comparison \(k\), define

- \(A_k\) and \(B_k\) as the first- and second-presented items;
- \(Y_k=1\) when \(A_k\) is selected and \(Y_k=0\) when \(B_k\) is selected;
- \(\theta_{A_k}\) and \(\theta_{B_k}\) as their hub-scale locations;
- \(b\) as the globally shared first-position effect; and
- \(\epsilon\) as the globally shared lapse probability.

The observation model is

\[
  q_k=\operatorname{logit}^{-1}
  (\theta_{A_k}-\theta_{B_k}+b),
  \qquad
  p_k=(1-\epsilon)q_k+\frac{\epsilon}{2},
  \qquad
  Y_k\sim\operatorname{Bernoulli}(p_k).
\]

Here `plogis()` implements the inverse-logit function. Positive \(b\) raises the modeled
probability of choosing the first-presented item, conditional on the location difference.
The lapse mixture shrinks predictions toward one half. These are model parameters, not measures
of forward/reverse response consistency.

### Hard-locked hub and spoke priors

For each hub item, Phase B fixes

\[
  \theta_h=\widehat\theta^{A}_h,
\]

where \(\widehat\theta^{A}_h\) is `theta_raw_mean` in the hub artifact. Hub parameters are absent
from the free optimization block; they are not merely assigned narrow priors. Consequently Phase B
cannot move the hub to accommodate a spoke.

For spoke item \(x\), let \(m_x^A\) and \(s_x^A\) denote the Phase A `theta_raw_mean` and
`theta_raw_sd`. The Phase B prior is

\[
  \theta_x\sim\mathcal N(m_x^A,\tau_x^2),
  \qquad
  \tau_x=c\max(s_x^A,s_{\min}),
\]

where \(c\) is `anchored_joint_spoke_prior_scale` and \(s_{\min}\) is
`anchored_joint_sd_floor`. If \(s_x^A\) is non-finite, the implementation substitutes
`anchored_joint_spoke_prior_fallback_sd` and emits a warning. The prior carries Phase A location
and uncertainty into Phase B; active cross-set evidence and the spoke's exact Phase A within-set
edges update the spoke locations.

The implementation minimizes the negative log posterior over the vector
\(\boldsymbol\theta_j=(\theta_x:x\in\mathcal S_j)\) using Broyden--Fletcher--Goldfarb--Shanno
(BFGS) optimization. When optimization converges, it inverts the observed Hessian to obtain a
Laplace covariance approximation. `theta_link_eap` is the accepted common-scale location exposed
to users, although the deterministic Phase B value is a MAP estimate rather than an MCMC posterior
mean. `theta_link_sd` is the corresponding marginal Hessian-based SD. The established output name
is retained for schema compatibility.

### Globally shared judge parameters

Before Phase B, the runner pools the exact Phase A within-set evidence from every required set and
fits one BTL judge surface. The resulting position effect \(b\) and lapse probability \(\epsilon\)
are copied into every spoke's accepted state and held fixed during anchored-joint refits. Thus the
item locations are spoke-specific but the judge parameters are global across the hub and all
spokes. This avoids attributing a separate position or lapse process to each spoke when the same
judge and observation contract produced all comparisons.

## Routing ranks, anchors, and candidate domains

Phase B constructs a linking-global score map from the fixed hub locations and the latest accepted
spoke locations. Decreasing score, with item identifier breaking ties, supplies the global rank
used for routing. The same stratum construction as within-set ranking divides that order into a
fine top band and size-dependent base regions.

Hub anchors are selected from hub-only ranks. Non-anchor stages use the active hub domain; the
anchor stage may use the full hub domain. Every Phase B candidate has exactly one hub endpoint and
one endpoint from the currently routed spoke.

| Stage | Candidate domain | Design role |
|---|---|---|
| `anchor_link` (anchor link) | Hub anchor plus spoke item. | Revisit reference regions across the hub distribution. |
| `long_link` (long link) | Active endpoints far apart. | Maintain broad links and coarse ordering checks. |
| `mid_link` (mid link) | Active endpoints at intermediate distance. | Bridge broad and neighboring score regions. |
| `local_link` (local link) | Active endpoints in nearby strata. | Resolve common-scale ordering locally. |

The stage label narrows the eligible domain; it is not a separate statistical model. Reserved
probe keys are removed from every active domain. Duplicate, recent-degree, exposure, and
presentation-order rules are then applied. When a domain is larger than `C_max`, the controller
uses a seeded bounded sample, preserving reproducibility for a fixed state and seed.

Spoke coverage uses quantile bins of current linking-global spoke scores. A candidate involving an
under-covered bin receives `coverage_priority = 1`; if any such candidates exist, selection is
restricted to that subset before utility ordering. This is a discrete coverage rule, not a numeric
addition to D-optimal gain.

## D-optimal active selection

The active selector asks which eligible pair would produce the largest local increase in
information about the free spoke parameters. For a candidate hub item \(h\) and spoke item \(x\),
it averages the two presentation orientations,

\[
  \bar p_{hx}=\frac{1}{2}p(h\text{ in }A,x\text{ in }B)
              +\frac{1}{2}p(x\text{ in }A,h\text{ in }B),
\]

and uses the implemented information scale

\[
  w_{hx}=\bar p_{hx}(1-\bar p_{hx}).
\]

Let \(\mathbf e_x\) be the vector with one at spoke item \(x\)'s position in the free block and
zero elsewhere. The candidate's rank-one information contribution is

\[
  \mathbf I_{hx}=w_{hx}\mathbf e_x\mathbf e_x^\mathsf T.
\]

Let \(\mathbf I_t\) be the information accumulated from committed active pairs for the spoke in
the current refit window. With numerical ridge \(r=10^{-6}\), the stored D-optimal gain is

\[
  U_{hx}=\log\det(\mathbf I_t+\mathbf I_{hx}+r\mathbf I)
          -\log\det(\mathbf I_t+r\mathbf I),
\]

where \(\mathbf I\) is the identity matrix. The code uses the equivalent rank-one
`log1p()` calculation when possible and falls back to direct log determinants. Because each
cross-set observation has only one free spoke endpoint, the anchored-joint information state can
be stored as a diagonal vector. It is updated only after a valid active comparison; probes and
invalid attempts do not enter it.

Within the coverage-prioritized pool, candidates are ordered by decreasing `link_d_opt_gain`, then
lexicographically by `i` and `j`. An unavailable or entirely non-finite gain aborts selection. The
selector does not silently replace D-optimal ordering with TrueSkill ambiguity or random choice.

### Conceptual selection example

Suppose two eligible pairs address spoke items \(x_1\) and \(x_2\). Current accumulated diagonal
information is \((4,1)\), so \(x_2\) is less informed. If both pairs have \(\bar p=0.5\), then
\(w=0.25\). Ignoring the tiny ridge for illustration, their gains are approximately

\[
  \log(1+0.25/4)=0.061,
  \qquad
  \log(1+0.25/1)=0.223.
\]

The pair involving \(x_2\) ranks first because it increases the information determinant more. In
a real step, candidate strata, reserved probes, coverage priority, duplicate controls, and all
other filters are applied before this comparison of gains.

## Probes and calibration

At the beginning of a spoke epoch, the controller constructs a deterministic held-out panel across
hub and spoke rank bins. `probe_panel_edges` is a target, not a promise: the realized panel can be
smaller when the unique legal cross-product is too small. `probe_panel_shortfall` and
`probe_shortfall_reason` make that limitation visible.

Probe keys are reserved from active selection for the epoch. While a spoke is active, the runner
may collect at most `probe_pairs_per_refit_per_spoke` held-out pairs per refit window. The public
path uses `probe_acceleration_mode = "fixed_per_refit"`; a probe-only blocker does not increase
that cap. The active-floor logic protects a minimum amount of active work in a window, so probe
collection cannot consume the whole linking budget. A frozen spoke receives neither active nor
probe pairs.

For realized probe outcomes \(y_k\in\{0,1\}\) and predictions \(p_k\), the Brier score is

\[
  \operatorname{Brier}=\frac{1}{K}\sum_{k=1}^{K}(y_k-p_k)^2,
\]

where \(K\) is the number of valid realized probe edges. Lower values indicate better squared
probability error on this panel. Expected calibration error (ECE) partitions predictions into five
probability bins and computes the observation-weighted mean absolute difference between bin-average
outcomes and predictions. ECE is descriptive and depends on the realized panel and binning.

The runner also records:

- fractions of predictions near 0.5, in a midrange, and at extremes;
- unique hub and spoke endpoints represented;
- hub and spoke rank bins represented;
- the near-boundary Brier score;
- lagged probe-prediction root mean squared error (RMSE); and
- panel size, realization, reallocation, and shortfall fields.

These form `probe_quality_pass` and `probe_quality_blocker_codes`, a diagnostic assessment of panel
composition and calibration. In the current stop implementation, `probe_quality_pass` itself is
not included in `link_stop_pass`. The stopping conjunction instead uses the total Brier threshold,
minimum realized probe edges, and lagged probe-prediction RMSE described below. Users should still
inspect probe-quality blockers before interpreting a linked result.

## Refits, diagnostics, and identification

Each spoke has its own accepted item state, active-edge count, probe panel, accumulated information,
diagnostics, and stop history. At a Phase B refit, the anchored-joint fit uses the spoke's exact
Phase A within-set edges plus all valid active hub--spoke edges retained for that spoke. Hub within-set
evidence supports artifact validation and pooled judge estimation; fixed hub locations mean those
edges do not add free hub parameters to the per-spoke optimization.

The Phase B deterministic diagnostic gate requires all of the following:

- BFGS optimization reports convergence;
- all required fitted summaries are finite;
- the Hessian is positive definite; and
- finite non-negative marginal uncertainty is available.

`link_diagnostics_converged_pass`, `link_diagnostics_finite_summary_pass`, and
`link_diagnostics_uncertainty_pass` expose these checks. Divergence, R-hat, and ESS columns are
R missing values (`NA`) for MAP/Laplace Phase B fits because those are MCMC diagnostics. A failed
deterministic fit does not masquerade as a successful MCMC fit.

### Reliability and routing identification

For active hub and spoke items, let \(\widehat\mu_i\) be the accepted common-scale mean and
\(v_i\) its estimated marginal posterior variance. Define

\[
  V_\mu=\operatorname{Var}_i(\widehat\mu_i),
  \qquad
  \overline V_{\mathrm{post}}=\frac{1}{n}\sum_i v_i,
  \qquad
  R_{\mathrm{link}}=\frac{V_\mu}{V_\mu+\overline V_{\mathrm{post}}}.
\]

The implementation applies small denominator guards and returns an undefined value when the
decomposition is not numerically meaningful. `reliability_link_global` is conditional on the
model, prior, active item domain, and marginal approximation. It is not classical test-score
reliability and does not establish validity or independence of judgments.

`linking_identified` becomes true for a spoke when both

\[
  R_{\mathrm{link}}\geq
  \texttt{link_identified_reliability_min}
  \quad\text{and}\quad
  \rho_{\mathrm{rank}}(\mu^{\mathrm{TrueSkill}},\widehat\theta^{\mathrm{link}})
  \geq\texttt{link_rank_corr_min},
\]

where \(\rho_{\mathrm{rank}}\) is Spearman rank correlation and
\(\mu^{\mathrm{TrueSkill}}\) is the
current TrueSkill mean. This flag changes routing, including long-link tapering. It is not formal
proof of statistical identification.

## Budgets and concurrent spokes

The default active refit budget derives from the adaptive refit target. Stage floors are applied
first, the remainder is divided by the configured anchor/long/mid/local fractions, and deterministic
largest-remainder rules resolve integers. When `linking_identified` is true, the long target is
multiplied by `long_link_taper_multiplier` subject to `long_link_taper_floor`; released work is
redistributed according to current blocker weights.

Feasibility checks can shrink a stage target when its legal domain cannot fill the nominal quota.
The log distinguishes nominal budget, feasible budget, released work, and
`stage_budget_unfilled`. A shortfall does not authorize illegal duplicates or spoke--spoke pairs.

In `link_multi_spoke` mode, every active spoke first receives
`min_cross_set_pairs_per_spoke_per_refit` when capacity permits. Remaining active comparisons are
allocated in proportion to the sum of the top D-optimal gains for that spoke, using at most
`multi_spoke_budget_top_k` candidates. If all utility masses are zero, allocation uses equal
weights. Candidate capacity caps the allocation. This is concurrent utility-based scheduling, not
parallel judgments and not a guarantee of equal comparisons per spoke.

The Bayesian Phase A and Phase B refit dispatcher may use `future` workers when parallel refitting
is configured and available. Results are restored in deterministic spoke order. Parallel execution
does not change the evidence or sharing contract.

## Stopping, blockers, and freezing

Stopping is evaluated at linking refits, separately for each spoke. A row becomes
`link_stop_eligible` only when:

- a comparable lagged refit exists in the same spoke epoch;
- `min_refits_in_phase_b` has been reached;
- the deterministic link diagnostics pass;
- common-scale reliability is defined; and
- at least `probe_edges_min_for_stop` held-out probes have been realized.

An eligible row passes the instantaneous stopping conjunction when:

\[
  \begin{aligned}
  &R_{\mathrm{link}}\geq R_{\min},\\
  &\operatorname{Brier}\leq B_{\max},\\
  &\operatorname{RMSE}_{\mathrm{probe,lag}}\leq P_{\max},\\
  &\operatorname{RMSE}_{\theta,\mathrm{lag}}\leq T_{\max},\\
  &\text{hub anchored}=\texttt{TRUE}.
  \end{aligned}
\]

Here \(R_{\min}\), \(B_{\max}\), \(P_{\max}\), and \(T_{\max}\) are the logged
`*_used` thresholds. Probe RMSE compares current and lagged predicted probabilities on the realized
probe panel. Theta RMSE compares current and lagged common-scale locations over the configured
direct-evidence spoke scope. These stability statistics measure change between fitted states, not
agreement with an external truth.

Only eligible results enter a rolling window of `stability_window_refits`. `link_stop_pass` becomes
true when the window is full and at least `stability_passes_required` entries passed. The passes
need not be consecutive. A passing spoke is frozen once: `link_state_frozen = TRUE` and
`link_state_frozen_refit_id` record the transition. With `max_pairs_after_stop = 0`, no later pair
is scheduled for it. Other spokes continue until they independently freeze, starve, or the caller's
attempted-step budget ends.

`stop_blocker_codes` is a comma-separated audit summary. Current values include:

| Code | Meaning |
|---|---|
| `diagnostics_failed` | The applicable link-fit diagnostic contract did not pass. |
| `lag_not_eligible` | No comparable lagged refit exists in the current epoch. |
| `min_refits_not_met` | The spoke has not reached the minimum Phase B refit index. |
| `probe_edges_min_for_stop` | Too few held-out probe edges have been realized. |
| `reliability_link_global` | Reliability is undefined or below its stopping threshold. |
| `probe_brier` | Probe Brier score is unavailable or above its threshold. |
| `probe_pred_rmse_lagged` | Lagged probe prediction RMSE is unavailable or above threshold. |
| `theta_global_rmse_lagged` | Lagged common-scale theta RMSE is unavailable or above threshold. |
| `hub_not_anchored` | The required hub-anchor invariant is not satisfied. |
| `none` | No blocker in this stop conjunction is active. |

Reaching `n_steps` only exhausts the attempted-step allowance for that call. It does not set
`link_stop_pass`, invent an accepted fit, or imply that every spoke is ready. Invalid judgments can
consume this allowance without increasing active or probe evidence. Candidate starvation and
unfilled budgets remain visible in the step and link-stage logs.

## Code-derived defaults

Fixed contract values and size-scaled rules are resolved in the controller. The following tables
are generated from the implementation for a representative run with one 100-item hub and two
100-item spokes. Contract fields in the first table describe the current normative path and are not
additional public choices; documented public controls can override the numeric thresholds.

```{r linking-fixed-defaults, echo=FALSE}
link_defaults <- pairwiseLLM:::.adaptive_controller_defaults(300L)
link_defaults$run_mode <- "link_multi_spoke"
link_defaults <- pairwiseLLM:::.adaptive_controller_resolve_scaled_linking_defaults(
  controller = link_defaults,
  cfg_names = character(),
  set_ids = c(rep(1L, 100L), rep(2L, 100L), rep(3L, 100L))
)

fixed_default_names <- c(
  "link_estimation_mode", "judge_param_mode",
  "link_identified_reliability_min", "link_stop_reliability_min",
  "link_rank_corr_min", "stability_window_refits",
  "stability_passes_required", "min_refits_in_phase_b",
  "probe_brier_max", "probe_pred_rmse_max", "theta_global_rmse_max",
  "max_pairs_after_stop"
)
fixed_defaults <- data.frame(
  contract_or_control = fixed_default_names,
  default = vapply(
    link_defaults[fixed_default_names],
    function(x) paste(x, collapse = ", "),
    character(1L)
  ),
  row.names = NULL
)
knitr::kable(fixed_defaults)
```

```{r linking-scaled-defaults, echo=FALSE}
scaled_default_names <- c(
  "probe_pairs_per_refit_per_spoke", "probe_panel_edges",
  "probe_edges_min_for_stop"
)
scaled_defaults <- data.frame(
  resolved_control = scaled_default_names,
  value_for_100_item_max_spoke = vapply(
    link_defaults[scaled_default_names],
    as.integer,
    integer(1L)
  ),
  rule = c(
    "max(4, ceiling(0.0035 * maximum spoke size))",
    "max(160, ceiling(0.12 * maximum spoke size))",
    "max(120, ceiling(0.106 * maximum spoke size))"
  ),
  row.names = NULL
)
knitr::kable(scaled_defaults)
```

The resolved thresholds used for a particular run are repeated in `link_stage_log`. Audit those
`*_used` fields rather than assuming that a run used package defaults.

## Logs and user-visible outputs

The three canonical log surfaces have different units:

- `step_log` has one row per attempted judgment. Linking fields include set endpoints, spoke,
  stage, probe flags, predictive probability, D-optimal gain, filters, outcome validity, and token
  metadata where available.
- `round_log` has one row per Bayesian within-set refit. It remains the Phase A model and stopping
  audit surface.
- `link_stage_log` has one row per spoke per linking refit. It records the estimation contract,
  hub lock, judge sharing, active/probe budgets, information summaries, diagnostics, calibration,
  reliability, stability, blockers, stopping, and freezing.

`out$items` exposes within-set values in `theta_raw_eap`, `theta_raw_sd`, and `rank_raw`. Accepted
common-scale values appear in `theta_link_eap`, `theta_link_sd`, interval fields, and `rank_link`.
They remain `NA` until a Phase B state has been accepted for the item. `out$phase_a` exposes the
artifact manifest and per-set readiness, while `out$summary` and `out$logs$link_stage_log` expose
the current linking status.

The log schemas intentionally retain some transform-era columns for old-session compatibility.
On anchored-joint rows, transform-only fields are normalized to typed missing values. Their
presence in a persisted schema does not make them current configuration choices.

## Persistence and resume

When `session_dir` is supplied, the runner persists state, metadata, step, round, and link-stage
logs, the latest fit, Phase A artifacts, and optional item logs. Each file replacement is atomic;
the collection is checkpointed after completed refits, terminal events, the ordinary end of a
call, and the configured step cadence. A process failure can lose attempts after the last complete
checkpoint.

`validate_session_dir()` validates metadata and canonical log schemas. `load_adaptive_session()`
also rebuilds derived evidence indices and checks probe state. `resume = TRUE` then requires:

- identical item identifiers and row order;
- compatible schema and package state;
- the same linking mode, hub, model contract, and required Phase A evidence;
- unchanged artifact identities and hashes;
- consistent spoke epochs, accepted states, Fisher-information counts, probe panels, and realized
  probe indices; and
- consistent frozen and stopping state.

An incompatible or corrupt session aborts with the violated invariant. The runner does not silently
discard saved work, create a new probe panel, or restart Phase A. Use `resume = FALSE` with a new or
empty directory only when intentionally starting a new study.

Compatibility loaders can backfill fields needed to read older sessions and can normalize legacy
log names. Historical transform estimators, soft hub locks, free hub locks, alternative probe
acceleration schedules, and independent-spoke routing are not public options for new runs.

## Interpretation and evidence

The design intends to preserve a stable reference scale, direct active effort toward locally
informative spoke parameters, and reserve independent comparisons for calibration. Those goals
follow from the routing and estimation rules; they are not empirical findings about every judge,
item pool, or budget.

The package computes conditional model diagnostics and reproducible audit fields. Passing them
does not guarantee an unbiased judge, correct model specification, representative probe panel,
external validity, or minimum ranking error. Non-passing or unavailable diagnostics identify a
reason not to accept the automated stop; they do not diagnose the substantive cause on their own.
Evaluate content coverage, subgroup behavior, presentation effects, invalid-response patterns, and
external criteria alongside the package logs.

## Foundational references

Bradley, R. A., & Terry, M. E. (1952). Rank analysis of incomplete block designs: I. The method of
paired comparisons. *Biometrika, 39*(3/4), 324--345.

Brier, G. W. (1950). Verification of forecasts expressed in terms of probability. *Monthly Weather
Review, 78*(1), 1--3.

Fedorov, V. V. (1972). *Theory of Optimal Experiments*. Academic Press.

Luce, R. D. (1959). *Individual Choice Behavior: A Theoretical Analysis*. Wiley.

> Mercer, S. H. (2026). *Adaptive-linking design* [R package vignette]. In *pairwiseLLM: Pairwise
> comparison tools for large language model-based writing evaluation* (Version 1.3.1).
