---
title: "Guide: Adaptive Linking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide: Adaptive Linking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article shows how to place rankings from separately evaluated sets onto one common scale.
It uses [adaptive_rank()] as the main entry point. The executable examples use a deterministic
local judge so that rendering does not require credentials or make network requests. A detailed
live-LLM workflow example appears later.

Adaptive linking has two phases:

1. **Phase A** ranks each set internally and produces a canonical within-set artifact.
2. **Phase B** collects selected hub--spoke comparisons and estimates spoke items on the hub scale.

These phases are distinct from the within-set selection stages described in
[Guide: Adaptive Pairing](adaptive-pairing.html). For the algorithmic rationale,
see [Design: Adaptive Linking](adaptive-linking-design.html).

## Public contract at a glance

| User task | `adaptive_rank()` input or output | Current behavior |
|---|---|---|
| Identify sets | `data$set_id` | Positive integer-like values; `hub_id` must name one observed set. |
| Identify items across sources | `data$global_item_id` | Non-empty and unique across the combined input. |
| Link one spoke | `run_mode = "link_one_spoke"` | The input must contain exactly one non-hub set. |
| Link several spokes | `run_mode = "link_multi_spoke"` | Spokes progress concurrently. |
| Obtain Phase A data | `phase_a_mode = "run"` | Run within-set work before any Phase B comparison. |
| Reuse Phase A data | `phase_a_mode = "import"` | Require compatible artifacts for every set and fail fast otherwise. |
| Combine both approaches | `phase_a_mode = "mixed"` | Import supplied sets and run all remaining sets. |
| Inspect Phase A | `out$phase_a` | Manifest, paths, and per-set source/status/validation message. |
| Inspect linked scores | `out$items` | `theta_link_eap`, its uncertainty summaries, and `rank_link`. |
| Diagnose Phase B | `out$logs$link_stage_log` | Per-spoke budgets, probes, gates, and blockers. |
| Continue safely | `session_dir`, `resume = TRUE` | Load compatible state with identical IDs and order. |

Phase B has one public estimation contract: anchored-joint estimation with hub item parameters
hard-locked to their Phase A values. Judge parameters are shared globally, including across
spokes. Candidate selection uses the implemented D-optimal utility. Held-out probes are collected
at the configured fixed per-refit cap while a spoke is active.

## Prerequisites

Bayesian Phase A refits and Phase B linking require `cmdstanr`, a working C++ toolchain, and
CmdStan. They do not require an LLM credential when a custom local judge is supplied.

```{r cmdstan-check}
cmdstan_available <- requireNamespace("cmdstanr", quietly = TRUE) &&
  tryCatch({
    cmdstanr::cmdstan_version()
    TRUE
  }, error = function(e) FALSE)

cmdstan_available
```

If this prints `FALSE`, install and verify CmdStan before running the remaining executable chunks:

```{r cmdstan-install, eval=FALSE}
install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
)
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

## Construct a hub and spokes

This is a **runnable offline example**. Each source system uses short local IDs such as `01` and
`02`, so those values repeat between sets. `ID` is the row identifier consumed by
`adaptive_rank()` and must be unique in the combined table. `global_item_id` is the stable
cross-source identity used by linking and must also be globally unique.

```{r linking-data}
library(pairwiseLLM)

make_set <- function(set_id, prefix, scores) {
  local_id <- sprintf("%02d", seq_along(scores))
  data.frame(
    ID = paste0(prefix, "_", local_id),
    text = paste("Writing sample", prefix, local_id),
    quality_score = as.double(scores),
    set_id = as.integer(set_id),
    global_item_id = paste0("study_2026_", prefix, "_", local_id),
    stringsAsFactors = FALSE
  )
}

hub_samples <- make_set(1L, "hub", c(6.0, 5.0, 4.0, 3.0, 2.0, 1.0))
spoke_2_samples <- make_set(2L, "school_b", c(5.5, 4.5, 3.5, 2.5, 1.5, 0.5))
spoke_3_samples <- make_set(3L, "school_c", c(5.2, 4.2, 3.2, 2.2, 1.2, 0.2))
linking_samples <- rbind(hub_samples, spoke_2_samples, spoke_3_samples)

stopifnot(!anyDuplicated(linking_samples$ID))
stopifnot(!anyDuplicated(linking_samples$global_item_id))

deterministic_judge <- function(A, B, state, ...) {
  list(
    is_valid = TRUE,
    Y = as.integer(A$quality_score[[1L]] >= B$quality_score[[1L]]),
    invalid_reason = NA_character_
  )
}
```

The simulated score exists only to make the judge reproducible. It is not passed to the Bayesian
model as an item score. A valid custom judge returns `Y = 1` when displayed item A wins and
`Y = 0` when B wins.

The following settings keep this vignette quick. Very low refit and probe thresholds are useful
for exercising the workflow, not for a substantive study. Production runs should normally retain
the package defaults unless a study design justifies changing them.

```{r tutorial-settings}
tutorial_btl <- list(
  refit_pairs_target = 3L,
  ess_bulk_min = 1,
  ess_bulk_min_near_stop = 1,
  max_rhat = 5,
  divergences_max = 100L
)

tutorial_link_controls <- list(
  phase_a_required_reliability_min = 0,
  probe_panel_edges = 8L,
  probe_pairs_per_refit_per_spoke = 1L,
  probe_edges_min_for_stop = 2L,
  min_refits_in_phase_b = 1L,
  stability_window_refits = 1L,
  stability_passes_required = 1L
)
```

## Phase A from raw samples

A practical workflow often ranks the hub and each spoke in separate jobs. Each call below starts
from raw samples and writes a reusable canonical artifact under its session directory.

```{r phase-a-runs, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
phase_a_root <- tempfile("pairwisellm-phase-a-")

hub_run <- adaptive_rank(
  data = hub_samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 6L,
  btl_config = tutorial_btl,
  session_dir = file.path(phase_a_root, "hub"),
  persist_item_log = TRUE,
  resume = FALSE,
  seed = 101L,
  progress = "none"
)

spoke_2_run <- adaptive_rank(
  data = spoke_2_samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 6L,
  btl_config = tutorial_btl,
  session_dir = file.path(phase_a_root, "spoke-2"),
  persist_item_log = TRUE,
  resume = FALSE,
  seed = 102L,
  progress = "none"
)

hub_run$phase_a$set_status
spoke_2_run$phase_a$set_status
hub_run$phase_a$artifact_paths
```

The status is `ready` only after the latest canonical artifact has committed within-set evidence,
passing MCMC diagnostics, and EAP reliability at or above
`phase_a_required_reliability_min`. A small call can instead return `pending_finalization`; resume
that same session with more steps until its quality criteria pass. Phase B never begins merely
because an artifact file exists.

```{r phase-a-inspection, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
hub_artifact <- hub_run$phase_a$manifest[["1"]]

names(hub_artifact)
hub_artifact[c("set_id", "n_items", "n_pairs_committed", "fit_model_id")]
hub_artifact$diagnostics
head(hub_artifact$items)
head(hub_artifact$phase_a_within_set_evidence)
```

`out$phase_a`, `out$phase_a$manifest`, an artifact `.rds` file, a
`phase_a_artifacts/` directory, or a saved session directory can be supplied to a later wrapper
call. The wrapper normalizes these surfaces before exact runtime validation.

## Link one spoke

For a production import, use only artifacts whose status is `ready`. This deterministic example
uses extremely small budgets, so the code below explicitly accepts its known simulated Phase A
artifacts as a **trusted override**. Setting `quality_gate_accepted = TRUE` tells the package to
trust external quality review; it does not repair weak evidence or diagnostics. Do not use it to
silence a failed real study.

```{r trusted-simulation-artifacts, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
phase_a_artifacts <- list(
  `1` = hub_run$phase_a$manifest[["1"]],
  `2` = spoke_2_run$phase_a$manifest[["2"]]
)

phase_a_artifacts <- lapply(phase_a_artifacts, function(artifact) {
  artifact$quality_gate_accepted <- TRUE
  artifact
})
```

Now combine the hub and one spoke and request import mode. Phase B compares hub items only with
spoke items; it does not schedule spoke--spoke pairs.

```{r one-spoke-link, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
one_spoke_samples <- linking_samples[linking_samples$set_id %in% c(1L, 2L), ]
one_spoke_session <- tempfile("pairwisellm-one-spoke-")

one_spoke <- adaptive_rank(
  data = one_spoke_samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 10L,
  adaptive_config = c(
    list(
      run_mode = "link_one_spoke",
      hub_id = 1L,
      phase_a_mode = "import",
      phase_a_artifacts = phase_a_artifacts
    ),
    tutorial_link_controls
  ),
  btl_config = tutorial_btl,
  session_dir = one_spoke_session,
  persist_item_log = TRUE,
  resume = FALSE,
  seed = 201L,
  progress = "none"
)
```

### Common-scale results

`theta_raw_eap` and `rank_raw` describe the latest within-set scale.
`theta_link_eap` and `rank_link` describe the common hub scale after an accepted Phase B refit.
Lower rank numbers are better; the latent score has no external unit.

```{r one-spoke-results, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
one_spoke$summary

one_spoke$items[, c(
  "item_id", "set_id",
  "theta_raw_eap", "rank_raw", "theta_link_eap", "theta_link_sd", "rank_link"
)]
```

Linked values remain `NA` before the first accepted Phase B fit. Reaching `n_steps` is only reaching
the attempted-step budget for that call; it is not evidence that the spoke is identified or ready
to report.

### Logs, probes, and blockers

```{r one-spoke-logs, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
names(one_spoke$logs)

cross_set_steps <- one_spoke$logs$step_log[
  one_spoke$logs$step_log$is_cross_set %in% TRUE,
  c(
    "step_id", "pair_id", "A_id", "B_id", "Y", "status",
    "link_spoke_id", "link_stage", "is_probe_step", "judge_invalid_reason"
  )
]
head(cross_set_steps)

link_status <- one_spoke$logs$link_stage_log[, c(
  "refit_id", "spoke_id", "link_estimation_mode", "hub_anchored",
  "reliability_link_global", "linking_identified", "link_stop_eligible",
  "link_stop_pass", "link_state_frozen", "stop_blocker_codes"
)]
tail(link_status)

probe_status <- one_spoke$logs$link_stage_log[, c(
  "refit_id", "spoke_id", "probe_edges_planned", "probe_edges_realized",
  "probe_panel_shortfall", "probe_shortfall_reason", "probe_quality_pass",
  "probe_quality_blocker_codes"
)]
tail(probe_status)

budget_status <- one_spoke$logs$link_stage_log[, c(
  "refit_id", "spoke_id", "B_spoke_refit_budget",
  "n_cross_edges_active_since_last_refit", "n_cross_edges_probe_since_last_refit",
  "stage_budget_unfilled"
)]
tail(budget_status)
```

The link-stage log is the authoritative diagnostic surface. `linking_identified` indicates that
the implemented identification gate passed; `link_stop_pass` means all stopping gates passed on
that row. Once a spoke stops, `link_state_frozen` is one-way for that run. A non-empty
`stop_blocker_codes` value identifies unmet gates such as diagnostics, reliability, coverage,
stability, or probe requirements.

Probes are held-out hub--spoke comparisons used for calibration and stopping checks. The planned
panel can be smaller than its target when the eligible pair domain is too small. The runner uses
the fixed `probe_pairs_per_refit_per_spoke` cap while a spoke is active; probes do not accelerate
just because they are the last blocker.

## Import and mixed Phase A modes

Import mode is fail-fast: every required set needs a compatible canonical artifact. Compatibility
checks include the set ID, item identities, item count, schema/configuration fingerprints, exact
within-set committed-edge evidence, and normalized BTL model variant. For example, an artifact
created with `model_variant = "btl"` cannot be imported into a `"btl_e_b"` run. Reuse it with the
matching variant or rerun Phase A; there is no cross-variant conversion.

Mixed mode imports the sets you supply and runs Phase A for every remaining set. Here the hub is
imported and spoke 3 is run locally:

```{r mixed-mode, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
mixed_samples <- linking_samples[linking_samples$set_id %in% c(1L, 3L), ]

mixed <- adaptive_rank(
  data = mixed_samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 6L,
  adaptive_config = c(
    list(
      run_mode = "link_one_spoke",
      hub_id = 1L,
      phase_a_mode = "mixed",
      phase_a_artifacts = list(`1` = phase_a_artifacts[["1"]])
    ),
    tutorial_link_controls
  ),
  btl_config = tutorial_btl,
  session_dir = tempfile("pairwisellm-mixed-"),
  resume = FALSE,
  seed = 301L,
  progress = "none"
)

mixed$phase_a$set_status
```

The expected sources are `import` for set 1 and `run` for set 3. If the run budget ends before set
3 passes its gate, its status remains `pending_finalization` and no Phase B work is scheduled.

Common import errors name the failed invariant: a missing artifact, set/item mismatch, absent
within-set evidence, different fit configuration, insufficient reliability, or incompatible model
variant. Treat the full error as a data provenance problem; do not catch it and silently start a
new link.

Useful message fragments map directly to corrective action:

| Message fragment | What to check |
|---|---|
| `configured for import but no artifact was provided` | Supply every required set or use mixed mode. |
| `global_item_id mapping mismatch` | Reconcile the artifact with the combined input identities. |
| `item-count metadata mismatch` | Use the exact Phase A item set, without adding or dropping rows. |
| `item_id mapping mismatch` | Restore the original row IDs and order for that set. |
| `missing reliability_EAP_within` | Rerun/finalize Phase A or perform a documented external review. |
| `reliability gate failed` | Continue Phase A or justify a trusted external quality decision. |
| `evidence-domain availability failure` | Recreate the artifact with canonical committed-edge history. |

Configuration- or model-variant mismatches similarly require the matching Phase A configuration;
the error reports the expected and observed surfaces.

## Link multiple spokes concurrently

First produce the third canonical artifact. As above, the trusted flag is appropriate here only
because the fixture and its deterministic truth are under our control.

```{r third-artifact, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
spoke_3_run <- adaptive_rank(
  data = spoke_3_samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 6L,
  btl_config = tutorial_btl,
  session_dir = file.path(phase_a_root, "spoke-3"),
  persist_item_log = TRUE,
  resume = FALSE,
  seed = 103L,
  progress = "none"
)

spoke_3_artifact <- spoke_3_run$phase_a$manifest[["3"]]
spoke_3_artifact$quality_gate_accepted <- TRUE
multi_artifacts <- c(phase_a_artifacts, list(`3` = spoke_3_artifact))
```

`link_multi_spoke` is the only multi-spoke run mode. All active spokes share the judge parameters,
but each has its own evidence counts, allocation, diagnostics, blockers, and frozen state.

```{r multi-spoke, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
multi_spoke <- adaptive_rank(
  data = linking_samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 16L,
  adaptive_config = c(
    list(
      run_mode = "link_multi_spoke",
      hub_id = 1L,
      phase_a_mode = "import",
      phase_a_artifacts = multi_artifacts,
      min_cross_set_pairs_per_spoke_per_refit = 1L
    ),
    tutorial_link_controls
  ),
  btl_config = tutorial_btl,
  session_dir = tempfile("pairwisellm-multi-spoke-"),
  resume = FALSE,
  seed = 401L,
  progress = "none"
)

multi_cross <- multi_spoke$logs$step_log[
  multi_spoke$logs$step_log$is_cross_set %in% TRUE &
    !is.na(multi_spoke$logs$step_log$pair_id),
  c("step_id", "set_i", "set_j", "link_spoke_id", "link_stage")
]
head(multi_cross)

latest_by_spoke <- multi_spoke$logs$link_stage_log[
  !duplicated(multi_spoke$logs$link_stage_log$spoke_id, fromLast = TRUE),
  c(
    "spoke_id", "B_spoke_refit_budget", "linking_identified",
    "link_stop_pass", "link_state_frozen", "stop_blocker_codes"
  )
]
latest_by_spoke
```

Every committed cross-set row has exactly one hub endpoint. Spokes run concurrently in the sense
that one controller advances all active spokes and allocates work among them; a single comparison
still involves only one spoke. One spoke can freeze while another continues.

## Persistence, interruption, and resume

With `session_dir`, the wrapper writes state, step, round, link-stage, metadata, optional BTL fit,
Phase A artifacts, and optional per-refit item logs. Writes are atomic at the individual-file
level. Completed refits, terminal events, the end of an ordinary call, and configured checkpoints
are saved. An abrupt process termination can lose attempts since the most recent checkpoint.

```{r linking-persistence, eval=isTRUE(get0("cmdstan_available", ifnotfound=FALSE))}
validate_session_dir(one_spoke_session)[c(
  "schema_version", "package_version", "n_items"
)]

saved_state <- load_adaptive_session(one_spoke_session)
summarize_adaptive(saved_state)

before_steps <- nrow(one_spoke$logs$step_log)
resumed_one_spoke <- adaptive_rank(
  data = one_spoke_samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 2L,
  adaptive_config = c(
    list(
      run_mode = "link_one_spoke",
      hub_id = 1L,
      phase_a_mode = "import",
      phase_a_artifacts = phase_a_artifacts
    ),
    tutorial_link_controls
  ),
  btl_config = tutorial_btl,
  session_dir = one_spoke_session,
  persist_item_log = TRUE,
  resume = TRUE,
  progress = "none"
)

c(
  before = before_steps,
  after = nrow(resumed_one_spoke$logs$step_log)
)
```

Resume is strict. The supplied IDs and order must exactly match the saved state. Stale or corrupt
schemas, incompatible artifacts, changed link configuration, or inconsistent probe/frozen state
abort with an actionable error rather than silently starting over. Use `resume = FALSE` only with
a new or empty directory when you intentionally want a new study. For expensive LLM runs, set
`checkpoint_every_steps` according to how many completed calls you can afford to repeat.

## A detailed live-LLM workflow

The following is a **live-API example and is not evaluated**. It uses the OpenAI configuration
recorded as tested with package 1.3.1 on 2026-09-03. A tested configuration is evidence about that
dated request shape, not a promise that the provider still offers the model. Check
[Backends and Tested Model Configurations](model-compatibility.html) and the provider catalog
before a long run.

Set `OPENAI_API_KEY` outside the script. Replace `real_linking_samples` with a data frame containing
unique `ID`, text, integer-like `set_id`, and globally unique `global_item_id` columns. Keep a copy
of that input in its exact row order for resume.

```{r live-llm-one-spoke, eval=FALSE}
library(pairwiseLLM)

stopifnot(nzchar(Sys.getenv("OPENAI_API_KEY")))

real_linking_samples <- utils::read.csv(
  "writing-samples-to-link.csv",
  stringsAsFactors = FALSE
)
stopifnot(!anyDuplicated(real_linking_samples$ID))
stopifnot(!anyDuplicated(real_linking_samples$global_item_id))
stopifnot(length(unique(real_linking_samples$set_id)) == 2L)

live_link <- adaptive_rank(
  data = real_linking_samples,
  id_col = "ID",
  text_col = "text",
  backend = "openai",
  model = "gpt-5.6-luna",
  endpoint = "responses",
  trait_name = "Overall writing quality",
  trait_description = paste(
    "Prefer the response that is clearer, better organized, better supported,",
    "and more effective for its intended audience."
  ),
  judge_args = list(
    reasoning = "none"
  ),
  n_steps = 200L,
  adaptive_config = list(
    run_mode = "link_one_spoke",
    hub_id = 1L,
    phase_a_mode = "run",
    max_pairs_after_stop = 0L
  ),
  session_dir = "adaptive-link-live",
  persist_item_log = TRUE,
  checkpoint_every_steps = 10L,
  resume = TRUE,
  seed = 20260904L,
  progress = "refits",
  save_outputs = TRUE
)
```

This single high-level call builds the LLM judge, performs or continues Phase A, enters Phase B only
when all required Phase A sets are ready, persists state, and returns reporting views. Do not add
tutorial thresholds from the offline fixture. The defaults provide the package's current quality
and stopping policy.

Inspect both progress and quality before reporting a linked ranking:

```{r live-llm-inspection, eval=FALSE}
live_link$phase_a$set_status
live_link$summary

live_link$items[, c(
  "item_id", "set_id", "theta_link_eap", "theta_link_sd", "rank_link"
)]

live_link$logs$step_log[, c(
  "step_id", "status", "judge_backend", "judge_model", "judge_endpoint",
  "judge_invalid_reason", "llm_status_code", "llm_error_message",
  "prompt_tokens", "completion_tokens", "total_tokens"
)] |>
  tail()

live_link$logs$link_stage_log[, c(
  "refit_id", "spoke_id", "linking_identified", "link_stop_pass",
  "link_state_frozen", "probe_edges_realized", "probe_quality_pass",
  "stop_blocker_codes"
)] |>
  tail()
```

`include_raw = TRUE` can preserve serialized raw provider responses for deeper auditing, but those
payloads may contain submitted text and increase storage. Enable it only under an appropriate data
retention policy. Invalid API, refusal, or parse responses consume attempted steps but do not
commit comparisons or update either ranking model. Inspect `status`, `judge_invalid_reason`, and
provider error fields before adding more budget.

Continue the exact study with the same input and configuration:

```{r live-llm-resume, eval=FALSE}
live_link <- adaptive_rank(
  data = real_linking_samples,
  id_col = "ID",
  text_col = "text",
  backend = "openai",
  model = "gpt-5.6-luna",
  endpoint = "responses",
  trait_name = "Overall writing quality",
  trait_description = paste(
    "Prefer the response that is clearer, better organized, better supported,",
    "and more effective for its intended audience."
  ),
  judge_args = list(reasoning = "none"),
  n_steps = 100L,
  adaptive_config = list(
    run_mode = "link_one_spoke",
    hub_id = 1L,
    phase_a_mode = "run",
    max_pairs_after_stop = 0L
  ),
  session_dir = "adaptive-link-live",
  persist_item_log = TRUE,
  checkpoint_every_steps = 10L,
  resume = TRUE,
  progress = "refits"
)
```

For independently completed Phase A jobs, change `phase_a_mode` to `"import"` and set
`phase_a_artifacts` to their wrapper outputs or session directories. For partial reuse, use
`"mixed"` and supply only the completed sets. For more than one spoke, change `run_mode` to
`"link_multi_spoke"`; no other mode switch is required.

Live comparisons usually dominate monetary cost and elapsed time. Phase A reuse avoids repeating
within-set judgments when the same compatible artifacts are linked again. Phase B cost grows with
active spokes, selected cross-set comparisons, held-out probes, invalid responses, and any extra
steps needed to clear diagnostics or stopping blockers. Bayesian refits add local computation.

## Troubleshooting a spoke that did not link

Use this order; it separates input failures from an ordinary incomplete run.

1. Check `out$phase_a$set_status`. Every required set must be `ready`; otherwise resume the named
   `run` source or replace the failed imported artifact.
2. Confirm `set_id`, `global_item_id`, the input IDs, and row order match the artifacts and any
   resumed session exactly.
3. Read the import error. Model-variant, config-hash, evidence-domain, and item mismatches require
   a matching artifact or a new Phase A run, not coercion.
4. Filter committed cross-set rows in `step_log`. No rows can indicate Phase A gating, candidate
   starvation, insufficient eligible anchors, or exhausted pair domains.
5. Inspect the latest `link_stage_log` row for the spoke. Read `stop_blocker_codes`, diagnostic and
   reliability fields, active/probe edge counts, coverage, shortfalls, and allocated budget.
6. If probes are short, inspect `probe_panel_shortfall` and `probe_shortfall_reason`; a tiny hub,
   spoke, or eligible-anchor pool can make the planned panel infeasible.
7. Treat `n_steps` exhaustion as an incomplete call. Resume with more budget only after ruling out
   repeated invalid judgments, persistent non-identification, or structural coverage problems.
8. If the session fails validation, preserve it for audit. Start a new directory only after the
   incompatibility is understood; never merge serialized files by hand.

Linked ranks are conditional on the observed judgments, the anchored-joint BTL model, its priors,
and successful diagnostics. Adaptive selection directs comparisons under the implemented utility;
it does not guarantee identification, a stable ranking, or a particular cost reduction.
