---
title: "Guide: Adaptive Warm Start"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Guide: Adaptive Warm Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(pairwiseLLM)
```

## What warm-start prediction does

A **warm-start prediction** estimates relative writing quality before comparisons
are collected. A **task-specific sub-model** learns from one assessment's texts or
features and its Bradley–Terry–Luce (BTL) scores. An **ensemble** averages calibrated
predictions from two or more separately trained task models. **Prior calibration**
uses out-of-fold (OOF) predictions to map model output to standardized quality;
`make_warm_start_prior()` then centers those values for predictive initialization.
The prior SD is a separate, explicitly chosen quantity, defaulting to 0.5.

Each sub-model standardizes its own outcome using its training mean and sample SD.
Independent BT/BTL scales do not need linking: raw unlinked scores are never pooled.
Model compatibility cannot establish that training assessments are independent or
that predictions are valid in a new population. Validate the intended domain.

Predictive information can initialize Bayesian BTL priors, TrueSkill locations,
or both. It does not change the initial observed pair schedule. Without predictive
input, existing cold-start behavior is preserved. **No default predictive model is bundled.** Real
bundled models remain deferred; the examples below use synthetic user models.

## Choose the predictive destination

`warm_start_mode` chooses which model receives the predictive information:

| Mode | BTL predictive prior | TrueSkill predictive `mu` | Connected bootstrap |
| --- | --- | --- | --- |
| `cold` | no | no | seeded shuffled chain |
| `btl_only` | yes | no | same seeded shuffled chain |
| `trueskill_only` | no | yes | same seeded shuffled chain |
| `both` | yes | yes | same seeded shuffled chain |

With omitted/`NULL` mode, **no predictive input defaults to `cold`** and
**predictive input defaults to `btl_only`**. This preserves existing callers.
Request `warm_start_mode = "both"` explicitly to initialize both models;
`trueskill_only` initializes TrueSkill while retaining cold BTL priors.

The following example uses numeric locations only, without Python, glmnet, model
training, or CmdStan. The names establish exact item-ID alignment.

```{r destination-modes}
locations <- c(a = -1, b = -0.25, c = 0.25, d = 1)
numeric_prior <- make_warm_start_prior(locations)
historical <- adaptive_rank_start(names(locations), seed = 17L,
  warm_start_prior = numeric_prior)
warmed <- adaptive_rank_start(names(locations), seed = 17L,
  warm_start_prior = numeric_prior, warm_start_mode = "both")
historical$meta$warm_start_mode
warmed$trueskill_state$items[, c("item_id", "mu", "sigma")]
```

In `trueskill_only` and `both`, the mapping is

```text
mu_i = mu0 + sigma0 * centered_predictive_location_i
mu0 = 25; sigma0 = 25/3; fixed scale multiplier = 1
```

The centered location is `prior$prior_mean`, aligned by item ID before any location
changes. TrueSkill sigma is unchanged by predictive initialization; its ordinary
default is `25/3`. This does not prevent sigma from evolving after observed judgments.
BTL `prior_sd` controls the Normal prior for raw theta. Ensemble disagreement,
calibration error, and RMSE are predictive diagnostics; none determines TrueSkill
sigma. The mapping never restores a training task's raw score units.

Explicit `cold` with predictive input errors, as does a non-cold mode without input.
`warm_start_model` and `warm_start_prior` are mutually exclusive. The model-only
`warm_start_prior_sd` override is a BTL control and is rejected in `trueskill_only`;
an already resolved prior retains its stored SDs.

Every mode keeps the same **connected shuffled bootstrap**: a seeded spanning path
of `N - 1` valid committed comparisons. Predictive initialization does not replace
observed connectivity or choose the first pair. With the same items and seed, the
initial unordered pairs, presentation balancing, and invalid-result retries are
common; a fixed judge yields the same initial outcomes. TrueSkill can evolve
differently from those outcomes because its initial locations differ.

After the bootstrap, `adaptive_config$pairing_strategy` selects `hybrid` (default),
`random`, `trueskill_p50`, or `trueskill_pollitt`. The direct strategies currently
require ordinary within-set mode. See [Guide: Adaptive Pairing](adaptive-pairing.html)
for their exact targets and restrictions.

## Optional extraction environment and frozen features

Schema inspection is available without Python:

```{r schema}
schema <- warm_start_feature_schema()
schema[c("position", "feature", "family")]
system.file("python", "README.md", package = "pairwiseLLM")
```

The installed README gives explicit user-run setup and CMUdict resource verification
instructions. There is no public installer. Package loading, examples and prediction
never create environments or download software. A Python installation needs `venv`
and `pip`/`ensurepip` support; maintainer bootstrap instructions are separate from
runtime extraction.

The tested stack is Python 3.12.3, TextDescriptives 2.8.4, textstat 0.7.13,
spaCy 3.7.5 and `en_core_web_lg` 3.7.1. The full lock and resource provenance are
installed under `python/`. Runtime version/resource checks reject incompatible
stacks with an actionable message. Linux testing does not establish Windows/macOS
compatibility.

Select an existing interpreter explicitly before initialization. A conflicting
`RETICULATE_PYTHON` setting or already initialized interpreter requires correcting
the configuration or restarting R. A status call can initialize Python and load
the large language model; it is not a prerequisite for precomputed prediction.
These commands are shown without execution because they require your environment:

```{r extraction, eval=FALSE}
python <- "/path/to/venv/bin/python"
status <- warm_start_python_status(python = python)
status$problems
features <- extract_warm_start_features(
  ids = c("a", "b"), texts = c("First response.", "Another response."), python = python
)
```

The returned table has character `item_id`, 20 numeric features in frozen order,
and `attr(features, "warm_start_schema") == "writing_features_v1"`. Missing or
duplicate IDs, missing required columns, and incompatible schemas fail explicitly.
Use RDS to preserve attributes. If a cache format drops attributes, restore the
known schema explicitly after verifying its origin; the attribute alone is not
verified extraction provenance. Required features are never replaced with zero.

The schema is curated from constructs and upstream definitions, not training outcomes.
See `warm_start_feature_schema()` for each feature's definition and missingness rules.
In particular, `upstream_entropy_per_token` divides upstream probability-weighted
entropy by **all** spaCy tokens, including punctuation/whitespace. It is neither
per-word perplexity nor conventional Shannon entropy, and does not use filtered
`n_tokens` as its denominator. Tokenization, zero-vector coherence and undefined
values follow the frozen upstream contract; upgrades cannot silently redefine v1.

## Develop one model per assessment

The following deterministic data are fabricated to illustrate the public interface.
They are not extracted student features.
Chunks needing glmnet/withr are skipped when those optional packages are unavailable.

```{r train, eval=requireNamespace("glmnet", quietly=TRUE) && requireNamespace("withr", quietly=TRUE)}
# Synthetic features illustrate the interface, not predictive validity.
example_features <- function(seed) {
  withr::local_seed(seed)
  fields <- warm_start_feature_schema()$feature
  x <- as.data.frame(matrix(runif(15 * length(fields)), nrow = 15))
  names(x) <- fields
  x$n_tokens <- 11:25
  x$token_length_mean <- 2 + 10 * x$token_length_mean
  x$token_length_std <- 0.2 + x$token_length_std
  x$dale_chall_readability_score <- 5 + 20 * x$dale_chall_readability_score
  x <- data.frame(item_id = as.character(1:15), x)
  attr(x, "warm_start_schema") <- "writing_features_v1"
  x
}
features <- example_features(3103)
theta <- 10 + 0.4 * features$n_tokens - 2 * features$token_length_mean
# A small alpha grid keeps this example fast; the default has 41 values.
model <- fit_warm_start_model(features$item_id, theta, "synthetic-a",
  features = features, alpha_grid = c(0, 1))

summary(model)
```

For real data, replace the synthetic table and `theta` with aligned item IDs,
precomputed features, and finite BT/BTL scores from **one** assessment. Alternatively,
pass `texts` and `python` instead of `features`; extraction runs once before fitting.
The `task_id` labels assessment provenance, not a registry name or grouping column.

Default validation is nested five outer by five inner folds, with seed 1 and
41 alpha candidates `seq(0, 1, by = 0.025)` spanning ridge through lasso. The example
uses two alphas solely to shorten execution. Folds are shared across alpha candidates.
Insufficient data produces an error; fold counts are not silently reduced.

Every applicable training fold learns missingness filtering (>20% missing or
all missing), median imputation, constant/near-zero-variance removal, and sample-SD
scaling. Near-zero variance requires unique fraction <=10% and frequency ratio >19.
Outcome scaling uses the corresponding outer training set throughout its inner
fits and held-out scoring. No PCA or univariate screening is introduced when p > n;
this example has 20 candidate features and 15 observations. See
`fit_warm_start_model()` for exact reference-path, weighted-loss, SE and tie rules.
The default lambda is the largest within one SE of the selected alpha's minimum;
`lambda_rule = "lambda.min"` is an explicit expert override.

OOF calibration learns an intercept/slope from predictions whose coefficient fits
excluded those rows. These folds also select hyperparameters, so calibration-fit
statistics are not independent validation. Calibration for an outer holdout uses
only its outer training data. Final deployment uses full-data OOF calibration and
an all-row coefficient refit. Degenerate calibration errors; finite negative slopes
are allowed, and undefined validation diagnostics have NA values with reasons.

```{r validation, eval=requireNamespace("glmnet", quietly=TRUE) && requireNamespace("withr", quietly=TRUE)}
model$validation$metrics
head(model$validation$predictions)
model$training[c("alpha", "lambda", "n_nonzero")]
```

Pearson r, squared Pearson r, Spearman rho, RMSE, MAE and diagnostic calibration
intercept/slope come from outer held-out predictions. Final refit predictions are
for deployment, not a replacement for those validation results. Outer fold records
retain their own selected alpha, lambda and nonzero counts.

## Save, register and combine task models

Model use from precomputed features needs neither glmnet nor Python. Store the
portable object, not a training-engine fit. Explicit saving and registration are
separate operations; ordinary save/load preserves the complete audit.

```{r storage, eval=requireNamespace("glmnet", quietly=TRUE) && requireNamespace("withr", quietly=TRUE)}
path <- tempfile(fileext = ".rds")
save_warm_start_model(model, path = path)
restored <- load_warm_start_model(path = path)
stopifnot(identical(predict(restored, features), predict(model, features)))
unlink(path)

local({
  # Keep this executable example out of the real user registry.
  withr::local_envvar(c(R_USER_DATA_DIR = withr::local_tempdir()))
  register_warm_start_model(model, name = "example")
  list_warm_start_models(source = "user")
  registered <- load_warm_start_model(name = "example", source = "user")
  remove_warm_start_model("example")
})
```

Real registrations use `tools::R_user_dir("pairwiseLLM", "data")` under `models`.
Names normalize to lowercase hyphen-separated identifiers; collisions require
explicit overwrite. Loading positionally always means a path. Same-name user and
bundled entries require an explicit source. List/remove operations manage obsolete
entries; no automatic backup history accumulates. Bundles are read-only.

Fit another independently generated synthetic assessment on its own scale, then
combine the models with stable component names. Actual training assessments must
be independent; relabeling the same training data does not establish independence.

```{r ensemble, eval=requireNamespace("glmnet", quietly=TRUE) && requireNamespace("withr", quietly=TRUE)}
features_b <- example_features(3104)
theta_b <- 30 + features_b$n_tokens - 3 * features_b$token_length_mean
model_b <- fit_warm_start_model(features_b$item_id, theta_b, "synthetic-b",
  features = features_b, alpha_grid = c(0, 1))
ensemble <- ensemble_warm_start_models(assessment_a = model, assessment_b = model_b)

predictions <- predict(ensemble, features)
predictions
summary(predictions)
```

## Inspect calibrated standardized coefficients

Use `warm_start_coefficients()` to inspect the fitted model weights without
reconstructing a glmnet fit. Individual models return one row per frozen feature in
schema order:

```text
feature | retained | calibrated_std_coefficient
```

```{r coefficients, eval=requireNamespace("glmnet", quietly=TRUE) && requireNamespace("withr", quietly=TRUE)}
coef_tbl <- warm_start_coefficients(model)
coef_tbl

coef_ensemble <- warm_start_coefficients(ensemble)
coef_ensemble
```

The predictors were centered and divided by their training-sample SDs, and the
fitted target was within-task standardized BT/BTL quality. For a retained feature,
the reported coefficient is the stored elastic-net coefficient multiplied by the
learned OOF calibration slope. Holding the other included predictors fixed, it is
therefore the change in calibrated within-task standardized prediction for a
one-training-SD increase in that feature.

`retained = FALSE` with coefficient `NA` means preprocessing removed the feature,
so no fitted standardized coefficient exists. `retained = TRUE` with coefficient
`0` means the feature survived preprocessing but elastic net assigned it zero
calibrated weight at the selected alpha and lambda. Positive and negative signs
describe fitted conditional direction. Correlated predictors can redistribute
weight, so coefficient magnitude is not unique predictive importance, causal
influence, or a share of explained variance.

Ensemble output has one ordered column per component:

```text
feature | assessment_a_std_coefficient | assessment_b_std_coefficient | ...
```

Each component used its own training distribution to standardize predictors; the
columns do not share one raw-feature SD. Side-by-side values show fitted direction,
magnitude, and stability across independently trained task models. `NA` and `0`
retain their component-specific meanings. The table neither estimates an aggregate
ensemble coefficient nor changes equal-weight prediction averaging. Once a
portable model exists, this inspection needs neither Python nor glmnet.

For a named list of models, use `do.call(ensemble_warm_start_models, models)`.
Components must share a frozen schema and standardized outcome definition, and have
learned calibration. Different preprocessing and full/reduced audit formats are
supported. Uncalibrated components, nested ensembles and failed components error.
No automatic stacking or silent component removal occurs.

| Quantity | Meaning |
| --- | --- |
| `raw_prediction` | Linear prediction on the standardized target scale |
| `calibrated_prediction` | Single-model output after its stored OOF calibration |
| `component_<name>` | Calibrated output from that ensemble component |
| `ensemble_mean` | Equal-weight mean of calibrated component outputs |
| `ensemble_sd` | Between-model sample SD, denominator k - 1; diagnostic only |
| `prior_mean` | Centered predictive location used by the selected destination mode |
| `prior_sd` | Chosen Normal SD for raw BTL theta, default 0.5; not TrueSkill sigma |

Complete component prediction tables remain in attributes, including raw predictions.
Their attributes do not automatically subset with tibble rows. Use explicit ID
alignment at prior boundaries; do not assume generic slicing updates nested metadata.
For text input, ensemble prediction extracts once using `texts`, `ids`, and `python`.
Single-model `predict()` accepts precomputed features only.

```{r bundles}
# Empty in this release. Future reviewed bundles use the same public prediction path.
list_warm_start_models(source = "bundled")
```

When bundles become available, load an explicit advertised name with
`load_warm_start_model(name = ..., source = "bundled")`. Named lookup verifies
manifest inventory, containment, size, checksum and metadata agreement. An MD5
checksum detects changed bytes; it does not authenticate a publisher.

## Convert predictions and resume

```{r prior, eval=requireNamespace("glmnet", quietly=TRUE) && requireNamespace("withr", quietly=TRUE)}
prior <- make_warm_start_prior(predictions, ids = features$item_id)
head(data.frame(item_id = prior$item_id, mean = prior$prior_mean, sd = prior$prior_sd))
state <- adaptive_rank_start(features$item_id, seed = 1, warm_start_prior = prior)
```

Conversion uses calibrated single predictions or ensemble means and centers them in R.
Do not inverse-transform to an original training scale or calibrate twice. Expert
numeric scores require names or explicit IDs. IDs must match exactly; scalar SDs
recycle, named SDs align by ID, and unnamed vectors follow input order. Invalid
scores/SDs or missing items error.

The SD is for `theta_raw`. Stan centers raw theta, inducing dependence and changing
centered-theta marginal SDs. All four active variants (`btl`, `btl_e`, `btl_b`,
`btl_e_b`) support supplied priors. Adaptive BTL consumes them only in `btl_only`
and `both`; `cold` and `trueskill_only` retain raw means zero and SDs one.
Default BTL-warm adaptive refits subset saved scores to the active fitted IDs before centering,
including fitted items without comparisons. Downstream transform, anchored-joint
and pooled judge refits retain their existing priors; evidence is not injected twice.

Standalone sampling requires your installed CmdStan toolchain and is not run here:

```{r sampling, eval=FALSE}
results <- build_btl_results_data(data.frame(ID1 = "1", ID2 = "2", better_id = "1"))
fit <- fit_bayes_btl_mcmc(results, ids = features$item_id, warm_start_prior = prior)
```

Instead of supplying a prior, initialize with `warm_start_model = ensemble` and
`warm_start_features = features`. Model objects, paths and explicit loader reference
lists are accepted. Without precomputed features, supply assessment texts and select
`warm_start_python`. Prediction runs once; initialization and resume never train.
Model-specific feature/Python/SD arguments cannot accompany an already resolved prior.
Custom BTL fit functions must check `state$meta$warm_start_mode` and consume
`state$predictive_prior` only in `btl_only` or `both`. The saved prior also exists
in `trueskill_only`, so presence alone does not mean BTL is warm. Custom-fit
signatures and the standalone `fit_bayes_btl_mcmc(warm_start_prior = ...)` interface
are unchanged; standalone BTL has no adaptive four-mode argument.

```{r resume, eval=requireNamespace("glmnet", quietly=TRUE) && requireNamespace("withr", quietly=TRUE)}
local({
  directory <- withr::local_tempdir()
  artifact <- file.path(directory, "ensemble.rds")
  save_warm_start_model(ensemble, artifact)
  session <- file.path(directory, "session")
  state <- adaptive_rank_start(features$item_id, session_dir = session,
    warm_start_model = artifact, warm_start_features = features)
  unlink(artifact)
  resumed <- adaptive_rank_resume(session)
  stopifnot(identical(resumed$predictive_prior, state$predictive_prior))
})
```

Sessions store numeric priors and compact provenance, not trained components or
nested-CV audits. Resume uses these saved values even if the original artifact is
removed or replaced, with no Python/glmnet requirement. On resumed `adaptive_rank()`
calls, omit **all** warm-start arguments, including mode. Saved mode, pairing
strategy, current TrueSkill values, bootstrap progress, and round state remain
authoritative. Older sessions with no mode migrate to `cold` without a prior and
`btl_only` with one; missing strategy becomes `hybrid`. Migration never warms
TrueSkill retroactively. Omit strategy on resume or supply its saved value;
changing it requires a new session.

Run-required linking Phase A supports predictive initialization and the common
per-set bootstrap. Imported artifacts retain their generation identity and are
not rerun merely because predictive input is supplied. Direct strategies cannot
run inside linking Phase A, although compatible artifacts from direct within-set
runs can be imported. Phase B selection and prior rules are unchanged: predictive
evidence is not injected into transforms, anchored-joint or pooled-judge priors,
D-optimal selection, or probes.
See [Guide: Adaptive Pairing](adaptive-pairing.html) for judging and continued runs,
and [Standalone Bayesian BTL with CmdStan](bayesian-btl.html) for sampling diagnostics.

## Model artifacts and the maintainer workflow

A portable model stores schema/version and original feature order, fitted
preprocessing, named coefficients/intercept, outcome mean/sample SD, calibration,
training metadata and validation evidence. Prediction validates the schema before
applying these parameters; no glmnet object is required.

Full-audit model format 1 retains IDs, outcomes, tuning traces and fold predictions.
`prepare_warm_start_model(model, omit_audit = TRUE)` creates model format 2 with
summary-only evidence and unchanged numeric predictions. Summary metrics cannot be
recomputed from reduced artifacts. Ensemble format 1 is separate from its component
formats, R serialization version, package version and manifest version.

Reduction is not anonymization. Task labels, notes, provenance and diagnostic prose
still need review. Preparation time is not training time, and supplied provenance is
not verified extraction history. Load only trusted RDS files.

In the source repository, `data-raw/warm-start/README.md` describes the explicit
maintainer build/review/promotion workflow. It uses the public APIs to train separate
assessments, keep full audits privately, reduce deployment components, and compare
predictions before/after storage. Text builds capture extraction versions and lock
hashes; cached provenance remains supplied or unavailable. Private recursive review
checks provenance, distribution rights, validation, retained sensitive material and
compressed size before promotion. Heuristics do not guarantee privacy. Review reports
can themselves contain restricted material and are not installed.

Only reviewed deployment artifacts and their manifest belong in installed `models`.
Training texts, full audits, staged files, environments and third-party model binaries
stay outside the package. No general predictive-performance or size threshold is
inferred from the synthetic examples or component validation metrics.

## Citation

> Mercer, S. H. (2026). *Guide: Adaptive Warm Start* [R package vignette]. Comprehensive
> R Archive Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
