---
title: "Guide: Adaptive Pairing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide: Adaptive Pairing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

cmdstan_available <- identical(Sys.getenv("PAIRWISELLM_RUN_CMDSTAN_VIGNETTES"), "true") &&
  requireNamespace("cmdstanr", quietly = TRUE) &&
  tryCatch(
    !is.null(cmdstanr::cmdstan_version()),
    error = function(e) FALSE
  )

knitr::opts_chunk$set(eval = cmdstan_available)
```

## Scope

This article shows how to rank **one set of items** with [adaptive_rank()]. The wrapper reads and
validates the items, starts or resumes a run, requests comparisons from a judge, performs Bayesian
refits, and returns the ranking and audit logs. Most users should start with this wrapper rather
than assembling the lower-level state and runner functions themselves. Executable CmdStan chunks
are disabled during ordinary package builds. Set `PAIRWISELLM_RUN_CMDSTAN_VIGNETTES=true` to opt in
when rendering this source locally.

## Choose a comparison workflow

Use exhaustive pairing when the set is small enough to evaluate every one of the
`choose(N, 2)` unordered pairs and complete coverage is important. Use a random sample when you
want a simple fixed-budget design or a baseline for comparison. Use adaptive selection when each
judgment is costly and you want the next comparison to respond to evidence already collected.

The runtime accepts `N >= 2` unique items. A range such as 30--2,000 items is operating guidance,
not an enforced limit. Very small sets may exhaust all eligible candidates before a Bayesian
refit, while large studies should be planned around provider cost, refit time, and storage.

## What happens during a run?

Within-set ranking has three phases.

1. **Connected shuffled bootstrap.** Every predictive mode uses the same seeded chain of
   `N - 1` valid comparisons to connect the observed graph. Predictive initialization occurs
   before these observations and does not change their schedule.
2. **Post-bootstrap pairing.** TrueSkill updates after every valid comparison. The selected
   pairing strategy chooses subsequent pairs; the default hybrid balances anchor, long, mid,
   and local comparisons, exposure, repeated-pair limits, and fallbacks.
3. **Phase 3: near stop.** After a Bayesian refit passes diagnostics and its EAP reliability is
   within 0.05 of the stopping threshold, later refits use the stricter near-stop ESS requirement.
   Pair-selection behavior does not otherwise change merely because Phase 3 was entered.

TrueSkill is the fast, step-by-step selection model. Bayesian Bradley--Terry--Luce (BTL) refits are
slower and intermittent. They provide posterior item estimates and uncertainty, convergence
diagnostics, EAP reliability, stability checks, and stopping decisions. Hybrid live ranks,
strata, rolling anchors, pair probabilities, base utility, and the long-link gate use TrueSkill
throughout. BTL still contributes to `global_identified`, which can change later hybrid tapering
and routing; selection is not wholly independent of BTL.

```text
items -> predictive initialization -> connected shuffled bootstrap
  -> select one pair -> judge -> commit valid outcome and update TrueSkill
           ^                          |
           |                          v
           +-- continue <- periodic BTL refit -> stop check
```

Adaptive selection is intended to direct effort toward useful comparisons, but it does not
guarantee a particular ranking, reliability, cost reduction, or improvement over random pairing.
Those outcomes depend on the items, judge, budget, and model assumptions.

## Choose a post-bootstrap pairing strategy

Set `adaptive_config = list(pairing_strategy = "trueskill_p50")`, for example.
The four choices apply after the common bootstrap:

| Strategy | Partner rule |
| --- | --- |
| `hybrid` (default) | Staged anchor/long/mid/local selection with TrueSkill ambiguity utility. |
| `random` | Uniform seeded choice among legal partners. |
| `trueskill_p50` | Minimize `abs(p_ts(i > j) - 0.50)`. |
| `trueskill_pollitt` | Minimize `min(abs(p_ts(i > j) - 1/3), abs(p_ts(i > j) - 2/3))`. |

Each direct strategy first chooses a focal item uniformly from sorted IDs at the
minimum current committed degree, using the run seed and committed count. It then
selects a legal partner by the rule above; target-distance ties break by partner ID.
Pollitt is **Pollitt-inspired**: it uses TrueSkill probabilities, whereas the earlier
article used BTL probabilities.

Direct strategies use no hybrid stage quotas, coverage overrides, or star-cap
fallbacks. They allow at most two observations per unordered pair, preserve normal
presentation balancing and reversal on repeat, and retry the same policy draw after
invalid results. They stop if the chosen focal item has no legal partner, even if
other pairs remain. These strategies currently support **ordinary within-set runs
only**; linking runs require `hybrid`. Phase B linking remains unchanged.

In direct step logs, `round_stage` and `pair_type` are `direct_pairing`,
`pairing_strategy` identifies the policy, and `i_id` identifies the focal item.
`p_ij` is the pre-judgment TrueSkill probability for presented A over B;
`target_distance` is symmetric under reversal and is missing for random pairing.
BTL refit cadence and stopping still apply to every strategy.

## Install and check CmdStan

Bayesian refits require the suggested `cmdstanr` package, a working C++ toolchain, and CmdStan.
This is a one-time machine setup; it does not require an LLM credential.

```{r cmdstan-install, eval=FALSE}
install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
)
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

Before starting a study, confirm that CmdStan can be found:

```{r cmdstan-check, eval=TRUE}
cmdstan_available

if (cmdstan_available) {
  cmdstanr::cmdstan_version()
}
```

The executable ranking chunks below run when CmdStan is available. The deterministic judge is
local, so the example never makes a network request or incurs provider charges.

## A complete offline run with `adaptive_rank()`

Start from a clean R session and load the package data. The `ID` column already contains unique,
non-missing identifiers. We retain `quality_score` only to simulate a judge.

```{r example-setup, eval=TRUE}
library(pairwiseLLM)

data("example_writing_samples", package = "pairwiseLLM")
samples <- example_writing_samples[, c("ID", "text", "quality_score")]

trait <- trait_description("overall_quality")
prompt_template <- set_prompt_template()

deterministic_judge <- function(A, B, state, ...) {
  a_wins <- A$quality_score[[1]] >= B$quality_score[[1]]
  list(
    is_valid = TRUE,
    Y = as.integer(a_wins),
    invalid_reason = NA_character_
  )
}
```

A custom judge receives one-row item tables `A` and `B` plus the current state. A valid response
must contain `is_valid = TRUE` and `Y` equal to `1` when A wins or `0` when B wins. The simulated
judge uses a fixture score so its choices are deterministic. Because it bypasses an LLM, the trait
and prompt are not consulted until the live example later in this article.

The minimum useful wrapper call supplies data, its ID and text columns, a judge, and a step budget.
We also choose a session directory so progress is persisted. With 20 items, the implemented
default `refit_pairs_target` is 20 committed comparisons: `ceiling(N / 2)` clamped to
`[20, 5000]`. A budget of 22 attempted steps covers the connected bootstrap and produces one refit
under this deterministic fixture.

```{r adaptive-run}
session_dir <- tempfile("pairwisellm-adaptive-")

out <- adaptive_rank(
  data = samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 22L,
  session_dir = session_dir,
  persist_item_log = TRUE,
  resume = FALSE,
  seed = 42L,
  progress = "none"
)
```

`n_steps` is a maximum number of attempted steps for that call, not a promise that every attempt
will commit and not a global stopping criterion. A run can return earlier after Bayesian stopping
or candidate starvation.

### Inspect the ranking and stopping state

The wrapper returns the final state and common reporting views together. `out$items` is sorted by
the current within-set rank; lower rank values are better.

```{r adaptive-results}
out$summary

out$items[, c(
  "item_id", "theta_raw_eap", "theta_raw_sd", "rank_raw", "degree"
)] |>
  head()

out$refits[, c(
  "refit_id", "total_pairs_done", "diagnostics_pass",
  "reliability_EAP", "stop_decision", "stop_reason"
)]
```

The latent `theta_raw_eap` scale is relative within this set. Its sign and absolute magnitude are
not an external score scale. Posterior standard deviations describe model uncertainty under the
fitted BTL model; they do not include every possible source of judge or sampling error.

`out$summary$last_stop_reason` remains missing when the wrapper merely reaches its requested step
budget. For within-set runs, terminal reasons include `"btl_converged"` and
`"candidate_starvation"`. Inspect the last refit row as well as the run summary: no single
diagnostic should be interpreted as proof that the ranking is correct.

### Inspect comparisons and audit logs

```{r adaptive-logs}
names(out$logs)

step_log <- adaptive_step_log(out$state)
step_log[, c(
  "step_id", "pair_id", "A_id", "B_id", "Y", "status",
  "round_stage", "fallback_used", "starvation_reason"
)] |>
  head()

summarize_adaptive(out$state)
summarize_refits(out$state, last_n = 1L, include_optional = FALSE)
summarize_items(out$state, top_n = 5L)

result_history <- adaptive_results_history(out$state)
head(result_history)
```

The step log has one row per attempt. A non-missing `pair_id` identifies a committed result. The
round log has one row per completed Bayesian refit, and the item log stores posterior item
summaries for each refit. `adaptive_results_history()` converts committed outcomes to the
three-column format accepted by `build_bt_data()`.

`adaptive_get_logs()` returns the step, round, link-stage, and item-step views together. Use
`adaptive_step_log()` to audit calls and transactional validity, `adaptive_round_log()` for refit
diagnostics and stopping, and `adaptive_item_log(stack = TRUE)` for change across refits. Empty
typed tables are expected before the corresponding event occurs; they are not evidence that the
accessor failed. `summarize_adaptive()` is a compact view, not a diagnostic recomputation.

## Lower-level lifecycle and offline judges

Most users should use `adaptive_rank()`. The lower-level lifecycle is useful when an application
needs to hold the state itself: create it with `adaptive_rank_start()`, advance it with
`adaptive_rank_run_live()`, persist it, and reconstruct it with `adaptive_rank_resume()` or
`load_adaptive_session()`. Do not modify state lists or canonical log columns by hand.

```{r lower-level-lifecycle}
low_level_items <- samples[1:5, ]
low_level_items$item_id <- as.character(low_level_items$ID)

low_level_state <- adaptive_rank_start(low_level_items, seed = 17L)
low_level_state <- adaptive_rank_run_live(
  state = low_level_state,
  judge = deterministic_judge,
  n_steps = 3L,
  btl_config = list(refit_pairs_target = 5000L),
  progress = "none"
)

summarize_adaptive(low_level_state)
```

A custom judge has the contract `judge(A, B, state, ...)`. `A` and `B` are one-row item tables.
Return `list(is_valid = TRUE, Y = 1L)` when A wins or `Y = 0L` when B wins. To simulate refusal,
timeout, or parsing failure, return `is_valid = FALSE`, `Y = NA_integer_`, and a stable
`invalid_reason`. Use a local score column, a fixed lookup table, or a separately seeded random
generator to make simulations reproducible. The package's internal scenario harness is test
infrastructure, not a public workflow function.

`make_adaptive_judge_llm()` builds the same judge interface around `llm_compare_pair()`. It is a
lower-level alternative to supplying backend/model arguments to `adaptive_rank()`, and its calls
are live, billable, and subject to the selected provider's failure and privacy behavior.

### Replay frozen directed judgments

`validate_adaptive_replay()` checks a table with character `A_id`, character `B_id`,
and binary `Y` (1 means presented A wins). By default it requires every one of the
`N * (N - 1)` ordered pairs for the panel. Self-pairs, foreign or missing IDs,
duplicate ordered keys, malformed outcomes, and missing orientations error.
`make_adaptive_judge_replay()` returns the stored result for the exact ordered key;
it never infers a reverse result from a forward judgment and never makes network calls.

This synthetic example runs both bootstrap and direct steps without a Bayesian refit:

```{r directed-replay, eval=TRUE}
replay_ids <- c("a", "b", "c", "d")
frozen <- expand.grid(A_id = replay_ids, B_id = replay_ids, stringsAsFactors = FALSE)
frozen <- frozen[frozen$A_id != frozen$B_id, ]
# Fabricated ordered outcomes, used only to demonstrate the interface.
frozen$Y <- as.integer(frozen$A_id < frozen$B_id)
frozen <- validate_adaptive_replay(frozen, item_ids = replay_ids)
replay_judge <- make_adaptive_judge_replay(frozen, item_ids = replay_ids)
replay_state <- adaptive_rank_start(replay_ids, seed = 17L,
  adaptive_config = list(pairing_strategy = "trueskill_p50", dup_max_obs_relaxed = 2L))
replay_state <- adaptive_rank_run_live(replay_state, replay_judge,
  n_steps = 5L, progress = "none")
adaptive_step_log(replay_state)[, c("A_id", "B_id", "Y", "pairing_strategy")]
```

Keep `strict_use = TRUE` and `complete = TRUE` for independent-evidence replay.
Set `adaptive_config$dup_max_obs_relaxed = 2L` so hybrid cannot request a third
observation from the two available orientations. Its general default remains 3;
direct strategies already cap at two. Presentation balancing and repeat reversal
remain active.

Create a fresh judge for each independent replicate. Strict use tracks successful
lookups in the judge closure and checks the supplied state's committed history.
The judge closure and matrix are not saved in the session: retain the matrix and
provenance separately. To resume, recreate a judge from the identical matrix and
pass the loaded state to `adaptive_rank_run_live()`. Missing or reused directed
judgments raise errors instead of supplying new evidence.

## Persistence, interruption, and resume

Supplying `session_dir` makes the wrapper persist the initial state, every completed refit,
terminal stops, the end of an ordinary call, and intermediate checkpoints. The default checkpoint
cadence is every 100 attempted steps; change it with `checkpoint_every_steps` when losing that much
work would be costly. Frequent checkpoints increase disk I/O.

An abrupt process termination can lose attempts since the most recent completed save. After an
interrupt, validate what is on disk rather than assuming the in-memory state was written.

```{r adaptive-persistence}
session_metadata <- validate_session_dir(session_dir)
session_metadata[c("schema_version", "package_version", "n_items",
  "warm_start_mode", "pairing_strategy")]

loaded_state <- load_adaptive_session(session_dir)
summarize_adaptive(loaded_state)

resumed <- adaptive_rank(
  data = samples,
  id_col = "ID",
  text_col = "text",
  judge = deterministic_judge,
  n_steps = 2L,
  session_dir = session_dir,
  persist_item_log = TRUE,
  resume = TRUE,
  progress = "none"
)

c(
  before = nrow(out$logs$step_log),
  after = nrow(resumed$logs$step_log)
)
```

Resume is intentionally strict. If saved artifacts are present but invalid, `adaptive_rank()`
aborts instead of silently creating a new run. The input IDs and their order must exactly match
the saved session. Set `resume = FALSE` only when you deliberately want a new state, and choose a
new or empty session directory to avoid mixing studies.

Resume preserves mode, strategy, predictive prior/provenance, current TrueSkill
state, committed history, bootstrap progress, and round state. Omit all predictive
warm-start arguments; omit strategy or supply the saved strategy. A legacy session
without mode becomes `cold` if no prior is saved and `btl_only` if one is saved;
missing strategy becomes `hybrid`. Migration never warms TrueSkill retroactively.
Within-set continuation also reuses saved `btl_config` when it is omitted.

The wrapper handles ordinary saving. For an explicit snapshot outside a wrapper call, use the
lower-level persistence helpers:

```{r adaptive-manual-save}
snapshot_dir <- tempfile("pairwisellm-snapshot-")
save_adaptive_session(resumed$state, snapshot_dir, overwrite = TRUE)
validate_session_dir(snapshot_dir)
snapshot <- load_adaptive_session(snapshot_dir)
```

Setting `persist_item_log = TRUE` writes a separate item-log file for each refit. The canonical
state, step log, round log, and metadata are persisted regardless; the option is useful when
per-refit item histories must be inspected independently.

## Invalid judgments and failed calls

Invalid judge output is transactional: it creates an attempted step with no committed `pair_id`,
does not update TrueSkill or comparison history, and does not count toward the BTL refit cadence.
It still consumes one unit of `n_steps`. The LLM judge created internally by `adaptive_rank()`
maps provider errors to invalid results and preserves available status, error, token, and raw
response fields in the step log.

This small example deliberately rejects its first attempt, then accepts the next one:

```{r invalid-judgment, eval=TRUE}
invalid_once_judge <- function(A, B, state, ...) {
  if (nrow(state$step_log) == 0L) {
    return(list(
      is_valid = FALSE,
      Y = NA_integer_,
      invalid_reason = "simulated_parse_failure"
    ))
  }
  deterministic_judge(A, B, state)
}

invalid_demo <- adaptive_rank(
  data = samples[1:4, ],
  id_col = "ID",
  text_col = "text",
  judge = invalid_once_judge,
  n_steps = 2L,
  resume = FALSE,
  seed = 7L,
  progress = "none"
)

adaptive_step_log(invalid_demo$state)[, c(
  "step_id", "pair_id", "status", "judge_valid", "judge_invalid_reason"
)]
```

Investigate repeated invalid rows before increasing the budget. Common causes include a thrown
judge error, a response that does not name either presented ID, or a custom judge that returns a
missing or non-binary `Y`.

## A detailed live-LLM workflow

The following is a **live-API example and is not evaluated**. It requires an OpenAI key, network
access, and incurs provider charges. The request shape shown here was recorded as tested with
pairwiseLLM 1.3.1 on 2026-09-05. That is dated evidence about this configuration, not a promise
that the provider still offers the model. Check
[Backends and Tested Model Configurations](https://shmercer.github.io/pairwiseLLM/articles/model-compatibility.html)
and the provider catalog
before a long run.

Set `OPENAI_API_KEY` outside the script. Replace `real_samples` with a data frame or file containing
one row per item, a unique ID, and the text to compare. Preserve that input in its exact row order
for resume.

```{r live-openai, eval=FALSE, purl=TRUE}
library(pairwiseLLM)

stopifnot(nzchar(Sys.getenv("OPENAI_API_KEY")))

real_samples <- utils::read.csv(
  "writing-samples.csv",
  stringsAsFactors = FALSE
)
stopifnot(nrow(real_samples) >= 2L)
stopifnot(all(c("ID", "text") %in% names(real_samples)))
stopifnot(!anyNA(real_samples$ID), !anyDuplicated(real_samples$ID))
stopifnot(!anyNA(real_samples$text), all(nzchar(real_samples$text)))

live_trait <- trait_description("overall_quality")
live_prompt <- set_prompt_template()
live_session <- "adaptive-live-session"

live <- adaptive_rank(
  data = real_samples,
  id_col = "ID",
  text_col = "text",
  backend = "openai",
  model = "gpt-5.6-luna",
  endpoint = "responses",
  trait_name = live_trait$name,
  trait_description = live_trait$description,
  prompt_template = live_prompt,
  judge_args = list(reasoning = "none"),
  n_steps = 200L,
  session_dir = live_session,
  checkpoint_every_steps = 10L,
  persist_item_log = TRUE,
  resume = FALSE,
  seed = 20260904L,
  progress = "refits",
  save_outputs = TRUE
)
```

This one call validates and normalizes the data, constructs the LLM judge, initializes the
within-set controller, attempts at most 200 judgments, performs Bayesian refits when due, persists
the session, and returns reporting views. `n_steps` is a per-call ceiling, not a target number of
valid comparisons or a guarantee that the stopping criteria will pass. Use a new or empty
`session_dir` with `resume = FALSE`.

The call intentionally retains the package's inference and stopping defaults. Do not copy reduced
diagnostic thresholds from a test fixture into a real study merely to make it finish sooner.

### Inspect the live run

Inspect the ranking, diagnostics, and stopping state before reporting results:

```{r live-openai-inspection, eval=FALSE, purl=TRUE}
live$summary

live$items[, c(
  "item_id", "theta_raw_eap", "theta_raw_sd", "rank_raw", "degree"
)] |>
  head(10L)

live$refits[, c(
  "refit_id", "total_pairs_done", "diagnostics_pass",
  "reliability_EAP", "stop_decision", "stop_reason"
)] |>
  tail()
```

An empty `live$items` means no successful BTL refit is available yet. Reaching the requested step
budget does not establish convergence. Read the latest refit diagnostics and stop reason together.

The step log records each attempted provider call, including provenance, validity, errors, and
available token counts:

```{r live-openai-audit, eval=FALSE, purl=TRUE}
live_steps <- live$logs$step_log[, c(
  "step_id", "pair_id", "A_id", "B_id", "Y", "status",
  "judge_backend", "judge_model", "judge_endpoint", "judge_invalid_reason",
  "llm_status_code", "llm_error_message",
  "prompt_tokens", "completion_tokens", "total_tokens"
)]

tail(live_steps)

colSums(live_steps[, c(
  "prompt_tokens", "completion_tokens", "total_tokens"
)], na.rm = TRUE)

live_steps[live_steps$status == "invalid", ]
```

Invalid API, refusal, or parse responses consume attempted steps but do not commit comparisons,
update TrueSkill, or advance the BTL refit cadence. Investigate repeated invalid rows before adding
budget. `include_raw = TRUE` on `adaptive_rank()` also stores serialized raw responses for deeper
auditing, but those payloads can contain submitted text and increase storage. Enable it only under
an appropriate retention policy.

### Continue the same study

Resume with the same input IDs, row order, trait, prompt, model settings, and session directory.
Only `n_steps`, progress display, and checkpoint cadence should normally change between calls.

```{r live-openai-resume, eval=FALSE, purl=TRUE}
live <- adaptive_rank(
  data = real_samples,
  id_col = "ID",
  text_col = "text",
  backend = "openai",
  model = "gpt-5.6-luna",
  endpoint = "responses",
  trait_name = live_trait$name,
  trait_description = live_trait$description,
  prompt_template = live_prompt,
  judge_args = list(reasoning = "none"),
  n_steps = 100L,
  session_dir = live_session,
  checkpoint_every_steps = 10L,
  persist_item_log = TRUE,
  resume = TRUE,
  progress = "refits",
  save_outputs = TRUE
)
```

Resume validates saved schemas and aborts on incompatible artifacts or mismatched input IDs instead
of silently starting a new run. A sudden process termination can lose attempts since the last
completed refit or checkpoint, so choose `checkpoint_every_steps` according to how many paid calls
you can afford to repeat.

Each live comparison is usually the dominant monetary and wall-clock cost. Bayesian refits are the
dominant local compute cost and become more expensive as the item set and posterior model grow.
Cost also rises with invalid calls and additional steps needed to clear diagnostics or coverage
gaps. Use bounded increments, inspect the audit and refit logs between calls, and do not add budget
when candidate starvation or a persistent structural problem requires a design change instead.

## Troubleshooting

**CmdStan is missing or compilation fails.** Run `cmdstanr::check_cmdstan_toolchain()` and
`cmdstanr::cmdstan_version()`. Install or repair the compiler toolchain before retrying. An LLM
credential cannot substitute for CmdStan.

**IDs are missing or duplicated.** `id_col` must identify a non-missing, unique column. For the
shipped data it is `id_col = "ID"`. Resume also requires the same IDs in the same order as the
saved state.

**The judge output is malformed.** A custom judge must return `is_valid`, `Y`, and preferably
`invalid_reason`. For a valid result, `Y` must be exactly `0` or `1`. Review
`judge_invalid_reason`, `llm_status_code`, and `llm_error_message` in `adaptive_step_log()`.

**The run reports candidate starvation.** The hybrid selector applies its implemented within-stage
fallbacks before declaring a stage starved. A terminal `"candidate_starvation"` means no eligible
pair remained after those fallbacks. This can occur with very small or heavily repeated designs;
inspect `fallback_path`, `starvation_reason`, and committed pair counts rather than treating it as
Bayesian convergence.
Direct strategies instead stop when the chosen minimum-degree focal item has no
legal partner; they do not search other focal items or invoke hybrid fallbacks.

**There is no item summary yet.** `out$items` is empty until the first successful BTL refit. Under
the default cadence, at least 20 new committed comparisons are needed. Invalid and starved
attempts do not advance that count.

## Related documentation

See [adaptive_rank()], [adaptive_step_log()], [adaptive_round_log()], [adaptive_item_log()],
[save_adaptive_session()], and [validate_session_dir()] for complete argument and schema details.
For the within-set algorithm and statistical rationale, see
[Design: Adaptive Pairing](https://shmercer.github.io/pairwiseLLM/articles/within-set-adaptive-design.html).

## Predictive warm-start priors

At initialization, `adaptive_rank()` accepts either `warm_start_model` or a resolved
`warm_start_prior`. With model input, precomputed `warm_start_features` avoids Python;
otherwise `warm_start_python` selects an existing extraction environment. Prediction
runs once and the session stores numeric priors. On resume, omit all warm-start
arguments: the saved prior remains usable after the original model is removed.
`warm_start_mode` selects `cold`, `btl_only`, `trueskill_only`, or `both`.
Omitted/`NULL` mode defaults to `cold` without predictive input and `btl_only`
with it. Request `both` explicitly to initialize both models. TrueSkill-warm modes
use `mu_i = 25 + (25/3) * prior_mean_i`, aligned by item ID with fixed multiplier 1
and unchanged sigma. BTL prior SD and ensemble diagnostics do not control sigma.
Explicit `cold` with predictive input, or a non-cold mode without input, errors.
All modes retain the same seeded connected shuffled bootstrap; subsequent
TrueSkill-based selections can differ. See
[Guide: Adaptive Warm Start](adaptive-warm-start.html) for training and prior conversion.

## Citation

> Mercer, S. H. (2026). *Guide: Adaptive pairing* [R package vignette]. Comprehensive R Archive
> Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
