---
title: "Adaptive comparative judgment: estimation, reliability, and evaluation"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Adaptive comparative judgment: estimation, reliability, and evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(pairwiseLLM)
has_withr <- requireNamespace("withr", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_withr)
```

## Five related questions

Comparative judgment (CJ) uses decisions about pairs of items to estimate their
relative strengths. An adaptive run chooses later pairs using earlier results.
Good pair selection, precise estimates, and accurate predictions are related,
but each needs its own assessment:

**Pair selection → parameter estimation → internal reliability → empirical score
recovery → held-out predictive performance.**

| Step | Question | What to inspect |
| --- | --- | --- |
| Pair selection | Which comparison should be collected next? | Strategy, allowed pairs, presentation order, and selection logs. |
| Parameter estimation | What strengths explain the collected outcomes? | Estimator, likelihood, adjustments, estimates, uncertainty, and convergence. |
| Internal reliability | How large is estimated item spread relative to estimated uncertainty? | SSR or Bayesian EAP reliability, with their definitions and provenance. |
| Empirical score recovery | How closely do scores reproduce an external reference? | Correlation, ranks, spread, and score differences against truth or a dense estimate. |
| Held-out predictive performance | How well do probabilities predict new judgments? | Log loss, Brier score, and pairwise accuracy on outcomes withheld from selection and fitting. |

The arrows organize the workflow; they do not imply that passing one step proves
the next. A dense score estimate is itself uncertain, and high internal
reliability does not establish agreement with it, judge validity, or predictive
calibration. Calibration means that predicted chances agree with observed win
frequencies over comparable predictions.

This guide uses synthetic outcomes only. It makes no provider or network calls.
The executable examples require the suggested `withr` package for local seeds;
if it is absent, the code is displayed without execution. The conventional and
Firth examples have additional optional-engine guards. Bayesian sampling is
shown separately and never runs during document builds. Example sizes, penalties,
and replicate counts illustrate interfaces, not recommended study settings.

## Start with comparisons chosen in advance

Here `theta` is relative strength on a log-odds scale. For the simple BT model,
the probability that A wins is `plogis(theta[A] - theta[B])`. Zero is a reference
location, not a passing score. Separately fitted panels do not automatically
share an origin or scale.

We choose all ordered pairs before drawing any winners. A fixed training set,
a denser reference set, and an evaluation set use separate random streams.
They share items and pair identities but contain independently generated
judgments. No evaluation outcome is used to choose pairs, fit scores, or select
settings. This assesses new judgments on known items, not generalization to
unseen items or judges.

```{r synthetic-data}
ids <- letters[1:6]
truth <- setNames(seq(-1, 1, length.out = length(ids)), ids)
pairs <- expand.grid(object1 = ids, object2 = ids, stringsAsFactors = FALSE)
pairs <- pairs[pairs$object1 != pairs$object2, ]
simulate_rows <- function(repeats, seed) {
  dat <- pairs[rep(seq_len(nrow(pairs)), each = repeats), ]
  p <- plogis(truth[dat$object1] - truth[dat$object2])
  dat$result <- withr::with_seed(seed, rbinom(nrow(dat), 1, p))
  rownames(dat) <- NULL
  dat
}
fixed_data <- simulate_rows(2L, 30601L)
reference_data <- simulate_rows(40L, 30602L)
heldout_data <- simulate_rows(20L, 30603L)
example_design <- data.frame(
  dataset = c("fixed training", "dense reference", "held-out evaluation"),
  outcome_seed = c(30601L, 30602L, 30603L),
  comparisons = c(nrow(fixed_data), nrow(reference_data), nrow(heldout_data)),
  selection = "all ordered pairs chosen before outcomes"
)
example_design
```

`fit_bt_model()` accepts these three columns: `object1`, `object2`, and binary
`result`, where 1 means the first item wins. For recorded winners in
`ID1`/`ID2`/`better_id` or `A_id`/`B_id`/`better_id` columns, first use
`build_bt_data()`. The [getting-started guide](getting-started.html) demonstrates
that conversion with bundled public data.

### Conventional fitting and estimator provenance

The default `engine = "auto"` tries sirt, then BradleyTerry2 if sirt is missing
or its execution fails. Explicit engines never switch to another estimator.
Validation failures such as disconnected comparisons are not repaired by that
automatic fallback. Use an explicit engine when an analysis requires a particular
method, and retain the returned `provenance`, including effective settings,
identification, convergence, engine version, and any fallback reason.

```{r conventional-fit, eval=requireNamespace("withr", quietly=TRUE) && requireNamespace("sirt", quietly=TRUE)}
conventional_fit <- fit_bt_model(
  fixed_data, engine = "sirt", sirt_eps = 0.3, verbose = FALSE
)
summarize_bt_fit(conventional_fit)
conventional_fit$ssr
conventional_fit$provenance[c(
  "requested_engine", "engine", "engine_version", "package_version",
  "adjustment", "convergence", "fallback_reason"
)]
```

The explicit `sirt_eps` above controls sirt's conventional epsilon adjustment;
omitting it preserves the installed engine's default. It is not a lapse
probability. sirt reports iterations rather than an explicit convergence flag;
inspect its recorded stopping status as well as reliability.

### What the adjustments change

| Estimator | Point-estimate adjustment | Reported uncertainty |
| --- | --- | --- |
| `engine = "sirt"` | Conventional epsilon adjustment depends on observed win proportions. | Engine SEs and independently checked SSR; retain resolved sirt settings. |
| `engine = "brglm2"` | Firth / mean bias reduction adds half the log determinant of expected information to the log likelihood (`AS_mean`). | Inverse original-data expected information at the adjusted estimate, transformed to centered item coordinates. |
| `engine = "alpha"` | Hamilton-style adjustment adds equal pseudo-wins in both directions for every possible pair, including unobserved pairs. | Inverse original-data expected information at the adjusted estimate, transformed to centered item coordinates. |

Firth is an explicit option for random or nonadaptive schedules, including
connected data with undefeated or winless items. Alpha adjustment is motivated
by adaptive scheduling. The package does not infer a schedule type from outcomes,
choose alpha, or establish that either adjustment is universally superior.
Neither adjustment selects pairs.

For \(N\) items, alpha adds \(c=\alpha/(N-1)\) pseudo-wins in each direction to
every unordered pair. With \(p_{ij}=\operatorname{logit}^{-1}(\theta_i-\theta_j)\),
the objective is

\[
\ell_\alpha(\theta)=\ell(\theta)+
\frac{\alpha}{N-1}\sum_{i<j}\log\{p_{ij}(1-p_{ij})\}.
\]

The Firth and alpha covariance matrices use the **original observed comparisons**,
not the inverse penalized Hessian. Their SEs are conditional on the realized
comparison graph: adjusting point estimates does not automatically account for
the variability of an outcome-dependent schedule. See `fit_bt_model()` for
the exact objectives, numerical checks, and the Hamilton and Tawn reference.

```{r firth-fit, eval=requireNamespace("withr", quietly=TRUE) && requireNamespace("brglm2", quietly=TRUE)}
firth_fit <- fit_bt_model(fixed_data, engine = "brglm2", verbose = FALSE)
summarize_bt_fit(firth_fit)
firth_fit$provenance[c("engine_version", "adjustment", "identification")]
```

```{r alpha-fit}
fixed_fit <- fit_bt_model(fixed_data, engine = "alpha", alpha = 0.5, verbose = FALSE)
summarize_bt_fit(fixed_fit)
fixed_fit$provenance[c("engine_version", "package_version", "adjustment")]
```

This alpha fit provides a common example that needs no additional fitting engine.
The penalty is chosen in advance for demonstration; it is not selected using the
reference or held-out outcomes.

## Internal reliability: SSR and Bayesian EAP

### Scale Separation Reliability (SSR)

Let \(\hat\theta_i\) and \(SE_i\) be estimates and standard errors on the same
scale for the same \(N\) items. Conventional SSR uses

\[
V_{\mathrm{observed}}=\frac{1}{N-1}\sum_i(\hat\theta_i-\bar{\hat\theta})^2,
\qquad V_{\mathrm{error}}=\frac{1}{N}\sum_i SE_i^2,
\]
\[
\widehat V_{\mathrm{true}}=V_{\mathrm{observed}}-V_{\mathrm{error}},
\qquad SSR=\frac{\widehat V_{\mathrm{true}}}{V_{\mathrm{observed}}}
=1-\frac{V_{\mathrm{error}}}{V_{\mathrm{observed}}}.
\]

In words, SSR subtracts average squared uncertainty from observed score spread,
then expresses the remainder as a proportion of that spread. “Estimated
true-score variance” is the name of this subtraction; it does not use known truth
or prove that the remainder equals empirical recoverable variance.

```{r ssr-components}
ssr_components <- scale_separation_reliability(fixed_fit$theta$theta, fixed_fit$theta$se)
ssr_components
```

The helper uses sample variance across items and the mean of **squared SEs**,
not the square of their mean. It does not align names for you. Negative SSR and
negative estimated true-score variance are retained: estimated uncertainty then
exceeds observed spread. Invalid inputs are not repaired by dropping items.

| Fit or calculation | SSR availability |
| --- | --- |
| sirt | Independently computed and checked against the engine's `mle.rel`; missing/invalid SEs or undefined arithmetic cause an error. |
| Firth or alpha | Available with valid uncertainty and positive score variance; a valid equal-score fit is retained with `zero_score_variance` and missing reliability. |
| BradleyTerry2 | Unavailable (`unavailable_se_convention`): its reference-based SEs are not used for centered-score SSR by this wrapper. |
| Experimental lapse model | Always unavailable (`prototype_ssr_unavailable`), for both interior and boundary fits. |
| Standalone SSR helper | Requires at least two finite scores with positive variance and matching finite, nonnegative SEs; otherwise errors. |

Do not combine adjusted scores with another estimator's SEs, or substitute
bootstrap SDs to manufacture a “corrected SSR.” Report the components, estimator,
SE convention, item set, and status along with the coefficient.

### Bayesian BTL and EAP reliability

An expected a posteriori (EAP) score is an item's posterior mean. If

\[
m_i=E(\theta_i\mid\text{data}),\qquad
v_i=\operatorname{Var}(\theta_i\mid\text{data}),
\]

the implemented Bayesian coefficient is

\[
R_{\mathrm{EAP}}=
\frac{\operatorname{var}_i(m_i)}
{\operatorname{var}_i(m_i)+N^{-1}\sum_i v_i}.
\]

It describes posterior mean spread relative to that spread plus average
posterior uncertainty. Unlike SSR, it adds uncertainty to the denominator
instead of subtracting it from observed variance. The variances across items
and across posterior draws use R's sample-variance convention. Invalid draws,
fewer than two items or draws, or nonpositive between-item variance give missing
reliability; otherwise the result is clamped to \([0,1]\).

The posterior model and priors affect both means and variances. Inspect sampling
diagnostics before interpreting this coefficient. It does not measure empirical
score recovery, held-out accuracy, or all sources of judgment error.

The following optional example fits the same fixed training outcomes with the
simple `btl` likelihood. It requires installed CmdStan, `cmdstanr`, `parallelly`,
and a compiler; it is **not evaluated during builds**. The seed and short sampler
settings demonstrate reproducibility, not an adequate sampling budget.

```{r bayesian-fit, eval=FALSE}
bayes_input <- build_btl_results_data(data.frame(
  ID1 = fixed_data$object1, ID2 = fixed_data$object2,
  better_id = ifelse(fixed_data$result == 1, fixed_data$object1, fixed_data$object2)
), backend = "synthetic", model = "fixed_binary_bt")
sampler_settings <- list(
  chains = 2L, parallel_chains = 1L, iter_warmup = 250L,
  iter_sampling = 250L, seed = 30604L
)
bayes_fit <- fit_bayes_btl_mcmc(
  bayes_input, ids = ids, model_variant = "btl", cmdstan = sampler_settings
)
list(model_variant = "btl", selection = example_design$selection[1L],
     sampler = sampler_settings, package_version = as.character(packageVersion("pairwiseLLM")))
summarize_refits(bayes_fit)[, c(
  "round_id", "total_pairs", "diagnostics_pass", "reliability_EAP",
  "max_rhat", "min_ess_bulk", "divergences"
)]
```

The [Bayesian BTL guide](bayesian-btl.html) explains setup, other likelihood
variants, item summaries, and diagnostics. The
[adaptive design guide](within-set-adaptive-design.html#diagnostics-reliability-and-stopping)
explains how EAP reliability participates in stopping. Passing that operational
rule is not independent validation of the scores.

## Collect an adaptive schedule

Every strategy begins with seeded comparisons connecting the items. Later
selection uses the current state:

| `pairing_strategy` | Later comparisons |
| --- | --- |
| `random` | Choose a minimum-degree focal item, then a random allowed partner. |
| `trueskill_p50` | Choose a minimum-degree focal item, then the partner closest to a TrueSkill win probability of one half (maximum ambiguity). |
| `trueskill_pollitt` | Choose a minimum-degree focal item, then the partner closest to a TrueSkill win probability of one third or two thirds. This is Pollitt-inspired, not the original BTL-probability procedure. |
| `hybrid` (default) | Mix anchor, long, mid, and local comparisons with exploration, coverage, and exposure constraints. |

Degree means the number of completed comparisons involving an item. Direct
strategies (the first three rows) balance presentation and reverse a repeated
pair; they allow at most two judgments per pair. They can stop when the chosen
focal item has no legal partner. Random here is therefore a specific constrained
policy, not uniform sampling from all pairs. A sparse replay reservoir adds its
own restrictions; this example uses ordinary within-set selection.

TrueSkill provides fast step-by-step selection estimates. Bayesian refits supply
posterior scores, diagnostics, reliability, and stopping checks; their
`global_identified` signal also affects hybrid routing. A later frequentist fit
does not replace the scheduling model or retroactively alter collected pairs.

### Hybrid recovery stays within hybrid

When an attempt cannot select a pair, the canonical ladder is:

| Attempt | What changes |
| --- | --- |
| `base` | Use normal stage bounds and duplicate policy. |
| `expand_locality` | Widen mid/local bounds, retaining long-link semantics. |
| `uncertainty_pool` | Keep base bounds and increase exploration. |
| `dup_relax` | Allow the conditional third judgment when the configured ceiling permits it; ambiguity and utility conditions still apply. |
| `global_safe` | Use the broadest stage-compatible domain with the ordinary two-judgment limit. |

Stage identity and hard constraints remain in force. If exploration misses every
pair but a previous attempt had eligible candidates, the usual exploitation rule
can select from the last viable pool. These are recovery attempts **within
hybrid**, not substitution of random, p50, or Pollitt strategies. Exhaustion can
stop collection before the requested budget. Inspect `fallback_used`,
`fallback_path`, and `candidate_starved` in `adaptive_step_log()` and the optional
report from `summarize_adaptive(state, include_starvation = TRUE)`. Direct
strategies do not invoke this ladder. See the
[design guide](within-set-adaptive-design.html#fallback-ladder) for the exact rules.

### A small provider-free p50 run

Save the pristine initial state before collecting outcomes. The synthetic judge
below supplies fresh winners. Its function name does not make a provider call:
`adaptive_rank_run_live()` also accepts a local judge like this one.

```{r adaptive-collection}
initial <- adaptive_rank_start(ids, seed = 17L,
  adaptive_config = list(pairing_strategy = "trueskill_p50"))
judge <- function(A, B, state, ...) {
  p <- plogis(truth[[A$item_id]] - truth[[B$item_id]])
  list(is_valid = TRUE, Y = as.integer(runif(1) < p))
}
schedule_config <- list(refit_pairs_target = 100L)
observed <- withr::with_seed(7304L, adaptive_rank_run_live(
  initial, judge, n_steps = 12L, btl_config = schedule_config, progress = "none"
))
adaptive_data <- adaptive_results_history(observed)
adaptive_fit <- fit_bt_model(adaptive_data, engine = "alpha", alpha = 0.5, verbose = FALSE)
selection_log <- adaptive_step_log(observed)
list(strategy = initial$controller$pairing_strategy, selector_seed = 17L,
     outcome_seed = 7304L, scheduling = schedule_config,
     committed_comparisons = nrow(adaptive_data), estimator = adaptive_fit$provenance$adjustment)
selection_log[, c("A_id", "B_id", "Y", "pairing_strategy", "fallback_used")]
```

`adaptive_results_history()` already returns the three columns needed by
`fit_bt_model()` and defaults to committed comparisons only. It needs no further
`build_bt_data()` conversion. This demonstration specifies a refit interval
larger than its entire budget, so it has no Bayesian refit or EAP coefficient.
It does not demonstrate statistical stopping. In a design with scheduled refits,
preserve those refits, including their optional CmdStan requirements. The
[practical adaptive guide](adaptive-pairing.html) covers that full lifecycle.

## Empirical recovery and scale dispersion

In a simulation we know the generating scores. With real data, a dense or
high-information fit can instead serve as a reference, but it is not known truth.
Its uncertainty, estimator, overlap with training outcomes, and model assumptions
affect apparent recovery. Here the dense reference uses independent judgments.
Align IDs and center scores before comparing differences; do not rescale each
vector to unit SD if scale expansion is the question.

```{r score-recovery}
reference_fit <- fit_bt_model(reference_data, engine = "alpha", alpha = 0.5, verbose = FALSE)
theta_adaptive <- setNames(adaptive_fit$theta$theta, adaptive_fit$theta$ID)[ids]
theta_reference <- setNames(reference_fit$theta$theta, reference_fit$theta$ID)[ids]
score_recovery <- function(estimate, reference) {
  data.frame(
    pearson_r_squared = cor(estimate, reference)^2,
    rank_correlation = cor(estimate, reference, method = "spearman"),
    score_sd = sd(estimate), reference_sd = sd(reference),
    sd_ratio = sd(estimate) / sd(reference),
    centered_rmse = sqrt(mean(((estimate - mean(estimate)) -
      (reference - mean(reference)))^2))
  )
}
recovery <- rbind(
  against_truth = score_recovery(theta_adaptive, truth[ids]),
  against_dense_estimate = score_recovery(theta_adaptive, theta_reference)
)
recovery
reference_fit$provenance$adjustment
```

Correlation describes association, not equality. Multiplying every score by a
positive constant leaves ranks and Pearson correlation unchanged, so it also
leaves **squared Pearson correlation** unchanged. This is the meaning of
“R-squared” here; a prediction-error statistic such as \(1-SSE/SST\) need not be
invariant. The score SD and variance do change, and BT differences become larger
log odds. Thus an adaptive schedule could stretch the scale while preserving
strong rank recovery. Examine spread separately instead of using correlation
as a test for the absence of scale distortion.

```{r scale-expansion}
scale_comparison <- rbind(
  reference = score_recovery(theta_reference, theta_reference),
  doubled = score_recovery(2 * theta_reference, theta_reference)
)
scale_comparison
```

Doubling here is an algebraic illustration, not a new estimator. This small
example is not a benchmark establishing which selector or estimator performs best.
Neither recovery table estimates calibration of SSR or EAP reliability; assessing
that requires an appropriate repeated-sampling design and reference target.

## Held-out predictive evaluation

Score recovery compares item scores. Predictive evaluation compares probabilities
with new binary judgments. For first-item outcomes \(y_k\) and probabilities \(p_k\),
use the following summaries (lower loss is better, higher accuracy is better):

\[
\mathrm{log\ loss}=-\frac1M\sum_k[y_k\log p_k+(1-y_k)\log(1-p_k)],
\qquad \mathrm{Brier}=\frac1M\sum_k(p_k-y_k)^2.
\]

Pairwise accuracy counts correct winners at a stated probability threshold.
It discards the strength of the probability: 0.51 and 0.99 choose the same winner
but express very different confidence. Log loss particularly penalizes confident
mistakes. Brier score also uses probabilities. These are overall predictive
scores, not complete calibration diagnostics.

```{r heldout-prediction}
p_adaptive <- predict(adaptive_fit, newdata = heldout_data[c("object1", "object2")])
p_reference <- predict(reference_fit, newdata = heldout_data[c("object1", "object2")])
p_expanded <- plogis(2 * qlogis(p_reference))
prediction_metrics <- function(p, y) {
  data.frame(
    log_loss = -mean(ifelse(y == 1, log(p), log1p(-p))),
    brier = mean((p - y)^2),
    accuracy = mean((p >= 0.5) == y)
  )
}
predictive_results <- rbind(
  adaptive = prediction_metrics(p_adaptive, heldout_data$result),
  dense_reference = prediction_metrics(p_reference, heldout_data$result),
  doubled_reference = prediction_metrics(p_expanded, heldout_data$result)
)
predictive_results
```

The example classifies exactly 0.5 as an A win. It does not clip probabilities:
a probability-zero realized event gives infinite log loss. Predictions here are
plug-in probabilities from the alpha fits. The doubled reference retains the
same predicted winners while moving probabilities farther from one half.
Held-out losses can expose overconfidence that rank correlation or accuracy
misses; no single finite test set guarantees this ordering.

Keep evaluation outcomes out of adaptive state updates, reference fitting,
penalty selection, and stopping decisions. If you tune a method using evaluation
results, reserve another independent test set for the final assessment. Sharing
item identities is intentional here, but copying the same judgment into training
and evaluation would not be independent evaluation.

## Schedule-aware parametric bootstrap

`bootstrap_bt_model()` simulates new binary BT outcomes from the initial fitted
scores, refits the chosen estimator, aligns item IDs, and centers scores. For
each item it computes

\[
\widehat{\mathrm{bias}}_i=\overline{\theta_i^*}-\hat\theta_i,
\qquad \hat\theta_{i,\mathrm{corrected}}=\hat\theta_i-\widehat{\mathrm{bias}}_i.
\]

This estimates point-estimate bias under the fitted generating model and the
specified design. Corrections can change ranks and need not improve recovery.

### Fixed versus adaptive scheduling

Fixed mode retains an outcome-independent schedule's ordered rows and repeats,
and draws fresh outcomes. It is appropriate for the training pairs chosen in
advance above:

```{r fixed-bootstrap}
fixed_boot <- bootstrap_bt_model(fixed_fit, mode = "fixed", n_rep = 20L,
                                 seed = 1304L, keep = "theta")
fixed_boot$theta
fixed_boot$provenance[c("mode", "seed", "pairing_strategy", "refit", "versions")]
```

An adaptive schedule depends on earlier outcomes. Its bootstrap must repeat the
selection decisions after simulating those outcomes. Freezing the observed
adaptive pairs conditions on a realized schedule and answers a different question.
The package uses the canonical runner, not a separate approximation to p50,
Pollitt-inspired, random, or hybrid selection:

```{r adaptive-bootstrap}
adaptive_boot <- bootstrap_bt_model(adaptive_fit, mode = "adaptive", n_rep = 4L,
  seed = 2304L, initial_state = initial, budget = 12L,
  btl_config = schedule_config, keep = "theta")
adaptive_boot$theta
adaptive_boot$replicates[, c(
  "replicate", "selector", "outcome", "scheduling_refit", "estimator",
  "success", "n_comparisons", "schedule_digest"
)]
adaptive_boot$provenance[c("mode", "seed", "pairing_strategy", "refit", "versions")]
```

The pristine initial connected pairs and presentations remain fixed; subsequent
selector randomness and simulated outcomes use deterministic per-replicate
seeds. Separate seeds cover selection, outcomes, scheduling refits, and the final
estimator. Compatible software and deterministic callbacks are needed for exact
reproduction. Serial and optional parallel execution use the same replicate
seeds and aggregation order.

Keep the original scheduling configuration. This example deliberately has no
scheduled refits within its budget; a design that uses Bayesian refits must rerun
them. In hybrid, those refits can affect later routing and tapering. The final
bootstrap estimator is separate from the scheduling fitter. The bootstrap budget
counts committed comparisons including initialization; statistical stopping does
not shorten that fixed-budget experiment, but terminal selection exhaustion
fails a replicate. Hybrid recovery remains hybrid, including inside bootstrap.

### What the Monte Carlo summaries mean

| Output | Meaning |
| --- | --- |
| `bias`, `theta_corrected` | Estimated bias and corrected point estimate under the fitted model and specified schedule semantics. |
| `bootstrap_sd` | Dispersion of the successfully refitted replicate estimates. |
| `mcse_bias` | Simulation error in the estimated bias, calculated as `bootstrap_sd / sqrt(n_success)`. |
| `n_success`, `n_failed`, `replicates` | Whether the correction has sufficient successful simulations and what happened in each replicate. |

Neither bootstrap SD nor Monte Carlo error is a corrected-score standard error.
The API does not provide corrected-score confidence intervals or corrected SSR.
These small replicate counts keep the example fast; their Monte Carlo error can
be substantial and they are not study recommendations. The default requires all
replicates to succeed. A deliberately lower `min_success` excludes failures and
therefore changes the summary to successful refits; inspect every failure. An
unmet threshold raises an error carrying `condition$result` diagnostics and
missing corrected scores.

The built-in final estimators are alpha and Firth; other supported estimators
require an explicit callback. The generating model is simple binary BT, without
lapse, position, or ties. Lapse fits are rejected. Keep full fit/bootstrap objects
to retain settings and provenance. The [bootstrap guide](bt-bootstrap.html)
explains callback requirements, optional parallel workers, and artifact retention.

## Experimental lapse and positional-bias model matching

`fit_bt_model(..., engine = "lapse")` uses the Bayesian `btl_e_b` likelihood
with frequentist unpenalized maximum likelihood:

\[
P(A\text{ wins})=(1-\epsilon)
\operatorname{logit}^{-1}(\theta_A-\theta_B+\beta)+\epsilon/2.
\]

The item strengths sum to zero; positive positional beta favors the first
presented item; lapse epsilon is the probability of guessing. Unlike conventional
sirt epsilon, lapse epsilon is an estimated likelihood parameter. The estimator
jointly fits all three parameter groups without Bayesian priors. It helps
separate differences in probability models from differences in estimation;
it is experimental, not the default BT estimator or a Bayesian replacement.

The following separate four-item example uses both presentation directions and
many inexpensive simulated binary rows to identify the extra parameters. The
counts are demonstration choices, not guidance on the data needed in practice.

```{r lapse-fits}
lapse_ids <- letters[1:4]
lapse_truth <- setNames(seq(-2, 2, length.out = 4), lapse_ids)
lapse_pairs <- expand.grid(object1 = lapse_ids, object2 = lapse_ids, stringsAsFactors = FALSE)
lapse_pairs <- lapse_pairs[lapse_pairs$object1 != lapse_pairs$object2, ]
lapse_pairs <- lapse_pairs[order(lapse_pairs$object1, lapse_pairs$object2), ]
fit_lapse_example <- function(beta, epsilon, seed) {
  p <- (1 - epsilon) * plogis(lapse_truth[lapse_pairs$object1] -
    lapse_truth[lapse_pairs$object2] + beta) + epsilon / 2
  wins <- withr::with_seed(seed, rbinom(nrow(lapse_pairs), 2000L, p))
  dat <- lapse_pairs[rep(seq_len(nrow(lapse_pairs)), each = 2000L), ]
  dat$result <- unlist(lapply(wins, function(w) c(rep(1L, w), rep(0L, 2000L - w))))
  fit_bt_model(dat, engine = "lapse", verbose = FALSE)
}
lapse_interior <- fit_lapse_example(beta = 0.3, epsilon = 0.2, seed = 305L)
lapse_boundary <- fit_lapse_example(beta = 0, epsilon = 0, seed = 30501L)
lapse_examples <- list(interior = lapse_interior, boundary = lapse_boundary)
lapply(lapse_examples, function(fit) {
  list(engine = fit$engine, beta = fit$beta, epsilon = fit$epsilon,
       convergence = fit$provenance$convergence$status,
       uncertainty = fit$provenance$uncertainty, ssr = fit$ssr)
})
list(selection = "all ordered pairs chosen before outcomes",
     outcome_seeds = c(interior = 305L, boundary = 30501L),
     model_variant = lapse_interior$model_variant,
     adjustment = lapse_interior$provenance$adjustment,
     package_version = lapse_interior$provenance$package_version)
predict(lapse_boundary, newdata = lapse_pairs)
```

A valid **interior MLE** has \(0<\epsilon<1\) and passes identification,
stationarity, and information/Hessian checks. Its joint covariance includes
estimation of beta and epsilon; item SEs are local model-based approximations
conditional on the realized graph, not a claim of repeated-sample calibration.

A valid **epsilon-zero boundary MLE** needs no guessing component. It retains
theta, beta, likelihoods, and predictions, with convergence status
`converged_boundary`. Ordinary joint interior uncertainty is unavailable:
`theta$se` is missing, `vcov` and `parameter_vcov` are `NULL`, and uncertainty
status is `nonregular_boundary`. A small positive epsilon alone does not
establish a boundary optimum. Epsilon one leaves strengths and beta unidentified
and is rejected. Connectedness alone is insufficient for this model's
identification, and genuine numerical/identification failures still error.

Conventional SSR is unavailable for **both** interior and boundary lapse fits.
Do not compute it manually from the interior SEs, or feed a lapse fit into the
simple-BT bootstrap. See `fit_bt_model()` and `predict.pairwiseLLM_bt_lapse()`
for the full experimental contract and failure diagnostics.

## Reporting the complete workflow

Report which strategy collected the comparisons, which estimator fit them, its
settings and uncertainty convention, and which internal reliability formula was
used. Then report recovery against the stated reference separately from held-out
predictive performance. Include score spread, evaluation-data provenance, seeds,
and software versions. Retain selection logs and complete fit/bootstrap objects.

The examples illustrate those distinctions. They do not establish universal
estimator or strategy superiority, universal reliability calibration, suitability
for a particular study, or equivalence of internal reliability and empirical
recovery.

## Citation

> Mercer, S. H. (2026). *Adaptive comparative judgment: estimation, reliability, and evaluation* [R package vignette]. Comprehensive R Archive Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
