---
title: "Guide: Adaptive Warm Start"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Guide: Adaptive Warm Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(pairwiseLLM)
```

## Before you start

Warm start is optional; skip it for your first adaptive run. Return here when
you have defensible prior scores or a prediction model developed from earlier
assessments. Start with [Adaptive Pairing](adaptive-pairing.html) for the ranking
workflow itself.

**You get:** starting scores for Bayesian Bradley–Terry–Luce (BTL), TrueSkill,
or both. They give the models initial information about relative writing quality.
Observed comparisons still drive the assessment; a warm start does not guarantee
better rankings or fewer comparisons.

The first example uses numeric scores and the optional `withr` package. The longer example uses fabricated
texts, already extracted features, and invented outcomes. It runs locally when
the optional `glmnet`, `pls`, `e1071`, and `withr` packages are installed. Otherwise,
its training and dependent chunks are skipped. Python is needed only to repeat
text extraction; CmdStan is needed for later Bayesian fitting, not initialization.
No example installs software or contacts a comparison provider.

**No pretrained predictive model is bundled.** The example data demonstrate the
workflow, not the accuracy of a model for students or a new assessment.

## Choose where starting scores are used

`make_warm_start_prior()` converts predictions into ID-aligned starting information.
It centers the scores by subtracting their mean. The **prior standard deviation
(SD)** is a separate choice: smaller values put more confidence in the starting
scores. The default of 0.5 controls the raw BTL prior; it is not the uncertainty of the
final centered scores. It is not learned from model accuracy or disagreement.

`warm_start_mode` chooses which model receives this information:

| Mode | BTL predictive prior | TrueSkill predictive `mu` | Default connected bootstrap |
| --- | --- | --- | --- |
| `cold` | no | no | seeded shuffled chain |
| `btl_only` | yes | no | same seeded shuffled chain |
| `trueskill_only` | no | yes | same seeded shuffled chain |
| `both` | yes | yes | same seeded shuffled chain |

The bootstrap is the initial connected set of observed comparisons. It is the
same for all modes under the default `bootstrap_policy = "shuffled_connected"`
when the items and seed are the same. Later pair choices can
differ as TrueSkill updates from its starting values and the observed judgments.

With no predictive input, the default mode is `cold`. With predictive input,
the default is `btl_only`. Request `both` explicitly to initialize both models.
The names below connect each starting score to its item:

```{r destination-modes, eval=requireNamespace("withr", quietly=TRUE)}
locations <- c(a = -1, b = -0.25, c = 0.25, d = 1)
numeric_prior <- make_warm_start_prior(locations)
historical <- adaptive_rank_start(names(locations), seed = 17L,
  warm_start_prior = numeric_prior)
warmed <- adaptive_rank_start(names(locations), seed = 17L,
  warm_start_prior = numeric_prior, warm_start_mode = "both")
historical$meta$warm_start_mode
warmed$trueskill_state$items[, c("item_id", "mu", "sigma")]
```


The output shows the saved mode and TrueSkill's initial locations (`mu`) and
uncertainties (`sigma`). TrueSkill maps centered scores to its own scale as
`25 + (25/3) * prior_mean`. By default, its initial sigma remains `25/3`
(or the constructor-supplied value); BTL prior SD does not set TrueSkill uncertainty.
Both models can subsequently learn from judgments.


### Opt in to predictive uncertainty

Use `warm_start_trueskill = "predictive_distribution"` with `trueskill_only` or
`both` to initialize the entire TrueSkill distribution. For each matching item ID,
`mu = 25 + (25/3) * prior_mean` and `sigma = (25/3) * prior_sd`, with `beta = 25/6`.
The mapping overrides constructor means and sigmas and never clips SD.

```{r predictive-distribution}
# Illustrative synthetic SD values; this example does not establish calibration.
distribution_prior <- make_warm_start_prior(c(a = -1, b = 0, c = 1),
  prior_sd = c(c = 0.8, a = 0.2, b = 0.5))
distribution_state <- adaptive_rank_start(c("c", "a", "b"), seed = 11,
  warm_start_prior = distribution_prior, warm_start_mode = "both",
  warm_start_trueskill = "predictive_distribution")
distribution_state$trueskill_state$items[, c("item_id", "mu", "sigma")]
distribution_state$meta$trueskill_mapping
```

With `warm_start_model`, this opt-in requires explicit `warm_start_prior_sd`;
it never silently substitutes the historical 0.5 default. Named vectors align by
exact item ID; unnamed vectors follow prediction-input order. Prebuilt priors use
their stored SD. A scalar remains available for sensitivity analyses but is not
essay-specific calibration. Legacy modes retain their existing 0.5 fallback.

The upstream predictive workflow must establish uncertainty calibration using
training data only. Explicit provision alone does not prove calibration, and
raw ensemble disagreement is not calibrated predictive uncertainty. Saved metadata
records the SD source, scalar/per-item rule, mapping version and parameters,
and integrity digests; calibration remains marked as upstream, unverified.

This convention assumes transformed BTL prior SD and TrueSkill uncertainty are
on comparable latent scales. BTL places the prior on `theta_raw`, then centers
effects for identifiability; centering induces dependence. The mapped sigma is
therefore not a marginal SD of centered BTL effects, and the two models' posteriors
are not equivalent. BTL prior and identifiability rules remain unchanged.
`trueskill_only` still leaves BTL cold; `both` also warms BTL. The mapping does not
depend on the graph or later selection strategy. Changing the distribution alone
retains the default shuffled bootstrap.

Resume retains the saved mapping and evolved ratings without initializing again.
Omit `warm_start_trueskill` along with all other warm-start arguments on resume.

## Choose the initial graph independently

`bootstrap_policy = "shuffled_connected"` preserves historical initialization.
Use `"predictive_connected"` to choose the initial connected tree from the frozen
predictive TrueSkill distribution. This requires `warm_start_trueskill =
"predictive_distribution"`, mode `trueskill_only` or `both`, a replay reservoir,
ordinary `within_set` mode, and explicit `pairing_strategy = "trueskill_pollitt"`.
Linking and Phase B are unsupported for this graph policy.

Create the reservoir from **selectable primary observations only**. Exclude
held-out edges and separate reversal audits before calling
`make_adaptive_replay_reservoir()`. The package validates the supplied manifest;
it cannot infer which observations your study designated as held out.

The tree builder receives only allowed A/B endpoints, frozen initial means/SDs,
and a seed. It favors probabilities near 1/3 or 2/3, starts with degree cap two,
and doubles that cap only when needed to connect the graph. Exact-score ties
use a local seeded permutation. It builds once before judging, preserves
recorded A/B orientation, and never reads outcomes or evolving ratings.
After `N - 1` successful bootstrap comparisons, the existing Pollitt selector
continues using current ratings and its existing degree rules.

These five configurations separate estimation, selection state, and graph effects:

| Arm | `warm_start_mode` | Predictive prior | `warm_start_trueskill` | `bootstrap_policy` |
| --- | --- | --- | --- | --- |
| Cold | `cold` | omitted | `NULL` | `shuffled_connected` |
| Estimation warm | `btl_only` | supplied | `NULL` | `shuffled_connected` |
| Legacy-graph coherent | `both` | supplied | `predictive_distribution` | `shuffled_connected` |
| Selection warm | `trueskill_only` | supplied | `predictive_distribution` | `predictive_connected` |
| Fully coherent | `both` | supplied | `predictive_distribution` | `predictive_connected` |

All five use the same selectable reservoir, seed, and Pollitt selector.
Legacy-graph coherent and fully coherent have identical predictive BTL priors
and initial TrueSkill means **and SDs**; only graph policy changes. The word
"legacy" here describes the graph, not historical fixed-sigma initialization.

This small example uses invented scores, SDs, and outcomes. They demonstrate the
API and do not establish predictive calibration or improved ranking accuracy.

```{r five-arm-bootstrap, eval=requireNamespace("withr", quietly=TRUE)}
bootstrap_ids <- letters[1:6]
bootstrap_observations <- data.frame(
  A_id = c("b", "c", "d", "e", "f", "a", "d", "b", "f"),
  B_id = c("a", "b", "c", "d", "e", "f", "a", "e", "c"),
  Y = c(1L, 1L, 0L, 0L, 1L, 0L, 1L, 0L, 1L))
bootstrap_reservoir <- make_adaptive_replay_reservoir(
  bootstrap_observations, bootstrap_ids)
bootstrap_prior <- make_warm_start_prior(
  setNames(c(-1, -0.6, -0.2, 0.2, 0.6, 1), bootstrap_ids),
  prior_sd = setNames(c(0.2, 0.7, 0.3, 0.8, 0.4, 0.6), bootstrap_ids))
bootstrap_modes <- c(cold = "cold", estimation = "btl_only",
  legacy_graph = "both", selection = "trueskill_only", full = "both")
bootstrap_states <- lapply(names(bootstrap_modes), function(arm) {
  adaptive_rank_start(bootstrap_ids, seed = 87L,
    replay_reservoir = bootstrap_reservoir,
    warm_start_mode = bootstrap_modes[[arm]],
    warm_start_prior = if (arm == "cold") NULL else bootstrap_prior,
    warm_start_trueskill = if (arm %in% c("cold", "estimation")) NULL else
      "predictive_distribution",
    bootstrap_policy = if (arm %in% c("selection", "full"))
      "predictive_connected" else "shuffled_connected",
    adaptive_config = list(pairing_strategy = "trueskill_pollitt"))
})
names(bootstrap_states) <- names(bootstrap_modes)
stopifnot(identical(bootstrap_states$selection$warm_start_pairs,
  bootstrap_states$full$warm_start_pairs))
stopifnot(identical(bootstrap_states$legacy_graph$trueskill_state,
  bootstrap_states$full$trueskill_state))
bootstrap_judge <- make_adaptive_judge_replay(bootstrap_reservoir)
bootstrap_partial <- adaptive_rank_run_live(bootstrap_states$full,
  bootstrap_judge, n_steps = 2L, progress = "none")
summarize_adaptive(bootstrap_partial, include_bootstrap = TRUE)$bootstrap[[1]]
```

An invalid judgment leaves queue progress and reservoir consumption unchanged;
retry uses the same ordered edge. `n_steps` counts attempts. When comparing arms
at equal budgets, count committed comparisons with
`summarize_adaptive(state)$committed_pairs`, not attempted requests.

Predictive sessions save their initial distribution, mapping identity, tree
policy/version, seed, ordered queue, diagnostics, and SHA-256 digest. Session
metadata repeats the policy/version and digest. Resume validates these frozen
artifacts without predicting, initializing ratings, or rebuilding the tree.
Keep the original reservoir to recreate its judge; changing its membership,
orientation, or outcomes is rejected.

```{r predictive-bootstrap-resume, eval=requireNamespace("withr", quietly=TRUE)}
bootstrap_session <- tempfile("predictive-bootstrap-")
save_adaptive_session(bootstrap_partial, bootstrap_session)
bootstrap_restored <- adaptive_rank_resume(bootstrap_session)
stopifnot(identical(bootstrap_restored$bootstrap, bootstrap_partial$bootstrap))
bootstrap_continued <- adaptive_rank_run_live(bootstrap_restored,
  make_adaptive_judge_replay(bootstrap_reservoir), n_steps = 1L, progress = "none")
unlink(bootstrap_session, recursive = TRUE)
```

On `adaptive_rank()` wrapper resume, omit initialization arguments. An explicit
bootstrap policy must match the saved policy; an explicit predictive seed must
also match. Warm-start mapping/model/prior overrides are rejected. Older saved
sessions without bootstrap policy metadata retain their shuffled queue, current
ratings, and progress. Default summary columns remain unchanged; use
`include_bootstrap = TRUE` for the additional audit list-column.

## A complete example with three algorithms

### Load example texts and inspect the features

A **feature** is a measured property of a text, such as its length or vocabulary.
`writing_features_v1` has 20 features and remains the default. The expanded
`writing_features_v2` keeps those 20 first and adds 26, using the same extraction
software. More features do not by themselves establish better prediction.

The package includes features extracted from fabricated English texts, so you
can run the rest of this example without Python. The first 40 texts are used for
training; eight different texts are reserved for prediction. The example's
`theta` values are invented, not estimates from a BTL analysis.

```{r example-data}
example <- readRDS(system.file("extdata", "warm-start-example.rds", package = "pairwiseLLM"))
training <- example$training
new_items <- example$new_items
features <- example$training_features
new_features <- example$new_features
schema <- "writing_features_v2"
head(training[c("item_id", "theta")])
head(warm_start_feature_schema(schema)[c("position", "feature", "family")])
```

In your own assessment, use one row per item and completed BT/BTL scores from
that assessment. Keep IDs, texts, and scores aligned. Do not pool raw scores from
separately ranked assessments. For separate writing traits, develop a model from
the corresponding trait's scores.

### Extract features when you have Python configured

These calls reproduce the example feature tables. They are shown without
execution in an ordinary render because they require your existing Python
environment. Substitute your own aligned texts and IDs when using real data.

```{r extraction, eval=FALSE}
python <- "/path/to/venv/bin/python"
status <- warm_start_python_status(python = python, schema = schema)
status$problems
features <- extract_warm_start_features(
  ids = training$item_id, texts = training$text, schema = schema, python = python)
new_features <- extract_warm_start_features(
  ids = new_items$item_id, texts = new_items$text, schema = schema, python = python)
```

Each table contains `item_id` and 46 numeric features in a fixed order. Undefined
values remain missing; the training procedure handles permitted missing values.
Use `saveRDS()` to cache features without losing the `warm_start_schema` attribute.
The schema attribute identifies the expected columns; it does not prove where
cached features came from. Saved v1 models still require v1 features.

The installed setup instructions are available with:

```{r python-instructions}
system.file("python", "README.md", package = "pairwiseLLM")
```

Extraction uses the pinned Python 3.12.3, spaCy 3.7.5, TextDescriptives 2.8.4,
textstat 0.7.13, and English model/resources described there. No environment or
resource is installed automatically. A status check loads the large language
model and can take time. If R has already initialized another interpreter,
restart R before selecting this one. The pinned environment has been tested on
Linux; Windows/macOS setup paths are not validation of those Python environments.

### Share the same training and validation splits

**Cross-validation (CV)** temporarily holds out some samples, trains on the
remaining samples, and predicts the held-out ones. Nested CV has two levels:
inner splits choose model settings; outer splits assess predictions for samples
kept outside that training and tuning process.

Make one plan and reuse it for all three algorithms. This gives each algorithm
the same training and held-out samples, making their validation results comparable.
The plan checks the task label, ordered IDs, and outcomes; it cannot be reused
with different scores or a different order of items.

```{r shared-plan, eval=requireNamespace("withr", quietly=TRUE)}
plan <- make_warm_start_cv_plan(training$item_id, training$theta,
  task_id = "example-assessment", seed = 259L, outer_folds = 5L, inner_folds = 5L)
```

During fitting, missing-value handling and feature scaling are learned separately
from each training split. Held-out samples do not determine that preprocessing.
Scores are standardized using the relevant training set, so held-out predictions
and observations are compared on the same scale. Too few suitable samples gives
an error; the requested fold counts are not silently reduced.

### Fit the three algorithms

| Algorithm | What it allows | Optional R package |
| --- | --- | --- |
| Elastic net | A weighted combination of features, with shrinkage to limit overfitting | `glmnet` |
| Partial least squares (PLS) | A few combinations of correlated features | `pls` |
| Radial-basis support-vector regression (RBF-SVR) | Nonlinear relationships between features and scores | `e1071` |

The examples use smaller explicit tuning grids to keep execution quick. These
are demonstration settings, not recommendations selected from the results.
The default grids and deterministic selection rules are documented in
`fit_warm_start_model()`.

```{r train, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
model <- fit_warm_start_model(training$item_id, training$theta, "example-assessment",
  features = features, schema = schema, cv_plan = plan, alpha_grid = c(0, 0.5))
pls_model <- fit_warm_start_model(training$item_id, training$theta, "example-assessment",
  features = features, schema = schema, cv_plan = plan, engine = "pls",
  engine_control = list(ncomp = 1:2))
svr_model <- fit_warm_start_model(training$item_id, training$theta, "example-assessment",
  features = features, schema = schema, cv_plan = plan, engine = "svr_rbf",
  engine_control = list(cost = c(0.5, 2), gamma_multiplier = c(0.5, 1)))
```

Each model also learns a calibration step from **out-of-fold (OOF)** predictions:
training-side predictions made without fitting on the corresponding rows.
Calibration puts the model output onto the standardized quality scale. It is
learned within the outer training set before predicting its held-out samples.
The final model uses all training samples only after outer validation is complete.

### Inspect validation before combining models

```{r validation, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
models <- list(elastic_net = model, pls = pls_model, svr_rbf = svr_model)
validation <- do.call(rbind, lapply(models, function(x) {
  unlist(x$validation$metrics[c("rmse", "mae", "pearson_r", "spearman_rho")])
}))
knitr::kable(validation, digits = 3)
```

These metrics describe the outer held-out predictions. Smaller RMSE and MAE mean
smaller prediction errors; larger Pearson or Spearman correlations mean stronger
agreement in scores or ordering. Squared Pearson correlation is the square of
that correlation, not a separate measure of unbiased prediction. Calibration
intercept/slope describe the validation predictions; they do not apply another
calibration to the deployed model. Undefined metrics are reported with reasons.

Do not replace these results with predictions on the training samples from the
final fitted model. The invented example outcomes make the table useful for
learning the interface only. It does not select an algorithm automatically or
establish performance on a new population.

### Combine algorithms trained on this assessment

A **same-task algorithm ensemble** averages the calibrated predictions of models
trained on the same assessment. Supply named, full model objects with matching
IDs, outcomes, feature schema, and CV plan:

```{r algorithm-ensemble, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
algorithm_ensemble <- ensemble_warm_start_algorithms(
  elastic_net = model, pls = pls_model, svr_rbf = svr_model)
summary(algorithm_ensemble)$validation$metrics
```

Each component receives equal weight after its own calibration. There are no
learned weights or additional ensemble calibration. The ensemble's validation
averages the aligned outer held-out predictions first, then compares those means
with the corresponding held-out observations. It does not average the three
models' error metrics. Legacy or reduced models lack the full shared evidence
needed to construct a new same-task ensemble.

### Predict new items and initialize an assessment

```{r algorithm-prediction, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
algorithm_predictions <- predict(algorithm_ensemble, new_features)
algorithm_predictions
algorithm_prior <- make_warm_start_prior(algorithm_predictions,
  ids = new_items$item_id, prior_sd = 0.5)
head(data.frame(item_id = algorithm_prior$item_id,
  mean = algorithm_prior$prior_mean, sd = algorithm_prior$prior_sd))
```

`ensemble_mean` is the mean calibrated prediction; `ensemble_sd` describes how
much the algorithms disagree. That disagreement does **not** set the prior SD.
Here we explicitly choose 0.5. The prior constructor centers the predictions for
these new items; it does not restore the old assessment's raw BTL scale.

Initialize any of the three predictive modes using the same resolved prior:

```{r algorithm-initialization, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
btl_state <- adaptive_rank_start(new_items$item_id, seed = 17L,
  warm_start_prior = algorithm_prior, warm_start_mode = "btl_only")
trueskill_state <- adaptive_rank_start(new_items$item_id, seed = 17L,
  warm_start_prior = algorithm_prior, warm_start_mode = "trueskill_only")
both_state <- adaptive_rank_start(new_items$item_id, seed = 17L,
  warm_start_prior = algorithm_prior, warm_start_mode = "both")
both_state$trueskill_state$items[, c("item_id", "mu", "sigma")]
```

These calls initialize state without collecting judgments or running Bayesian
sampling. Continue with the judging and running steps in
[Adaptive Pairing](adaptive-pairing.html). All four Bayesian BTL variants support
predictive priors. In `trueskill_only`, BTL keeps its cold prior.

## Save a model and resume a session

Prediction from precomputed features needs neither Python nor any fitting engine.
Ordinary save/load retains the full model and its validation evidence:

```{r storage, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
local({
  directory <- withr::local_tempdir()
  path <- file.path(directory, "algorithm-ensemble.rds")
  save_warm_start_model(algorithm_ensemble, path)
  restored <- load_warm_start_model(path)
  stopifnot(identical(predict(restored, new_features), algorithm_predictions))

  # This example uses a temporary registry, leaving your own registry alone.
  withr::local_envvar(c(R_USER_DATA_DIR = directory))
  register_warm_start_model(restored, "example")
  list_warm_start_models(source = "user")
})
```

Registered names are normalized to lowercase with hyphens. If a user model and
a bundled model share a name, select `source` explicitly. No pretrained models
are currently listed by `list_warm_start_models(source = "bundled")`.

To save an adaptive session, set `session_dir` at initialization. Resume uses the
saved numeric prior even if the original model file is no longer available:

```{r resume, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
local({
  directory <- withr::local_tempdir()
  state <- adaptive_rank_start(new_items$item_id, session_dir = directory,
    warm_start_prior = algorithm_prior, warm_start_mode = "both")
  resumed <- adaptive_rank_resume(directory)
  stopifnot(identical(resumed$predictive_prior, state$predictive_prior))
})
```

On resume, omit all warm-start arguments, including mode. The session retains its
prior, mode, TrueSkill state, and progress. To change the starting information,
start a new session. Model input and prior input are alternatives: do not supply
both. `warm_start_prior_sd` is available with model input for BTL warming and
required for predictive-distribution initialization; a resolved prior already
contains its chosen SDs.

## Combining models from separate assessments

A **cross-task ensemble** has a different purpose: combining calibrated models
from separately developed assessments. Each model standardizes its own outcome;
raw unlinked BT/BTL scores are never pooled. Use `ensemble_warm_start_models()`.
It retains component validation summaries, not a common ensemble validation score.

The following second assessment is entirely fabricated, including its feature
values. It illustrates the separate API without suggesting that relabeling the
first assessment would create independent training data.

```{r ensemble, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
features_b <- local({
  withr::local_seed(3104L)
  x <- features
  # Reuse valid feature ranges, but independently shuffle each column.
  for (field in warm_start_feature_schema(schema)$feature) x[[field]] <- sample(x[[field]])
  x$item_id <- paste0("assessment-b-", seq_len(nrow(x)))
  x
})
theta_b <- withr::with_seed(3105L,
  0.1 * features_b$n_tokens + rnorm(nrow(features_b)))
model_b <- fit_warm_start_model(features_b$item_id, theta_b, "synthetic-b",
  features = features_b, schema = schema, alpha_grid = c(0, 0.5))
ensemble <- ensemble_warm_start_models(assessment_a = model, assessment_b = model_b)
summary(ensemble)
```

Components must share a schema and standardized outcome definition and have
learned calibration. Their preprocessing and CV plans may differ. Repeated models
and task labels remain accepted for backward compatibility, but do not demonstrate
independent evidence. Nested ensembles are rejected. Assess whether the training
assessments and target population justify using their predictions together.

## Inspect calibrated standardized coefficients

Use `warm_start_coefficients()` to inspect the fitted model weights without
loading the original fitting engine. This applies to elastic net and PLS.
RBF-SVR has no equivalent linear feature coefficients; requesting them produces
an informative error, including when it belongs to an ensemble. Individual models return one row per frozen feature in
schema order:

```text
feature | retained | calibrated_std_coefficient
```

```{r coefficients, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
coef_tbl <- warm_start_coefficients(model)
coef_tbl

coef_ensemble <- warm_start_coefficients(ensemble)
coef_ensemble
```

The predictors were centered and divided by their training-sample SDs, and the
fitted target was within-task standardized BT/BTL quality. For a retained feature,
the reported coefficient is the stored elastic-net or PLS coefficient multiplied by the
learned OOF calibration slope. Holding the other included predictors fixed, it is
therefore the change in calibrated within-task standardized prediction for a
one-training-SD increase in that feature.

`retained = FALSE` with coefficient `NA` means preprocessing removed the feature,
so no fitted standardized coefficient exists. `retained = TRUE` with coefficient
`0` means the feature survived preprocessing but elastic net assigned it zero
calibrated weight at the selected alpha and lambda. Positive and negative signs
describe fitted conditional direction. Correlated predictors can redistribute
weight, so coefficient magnitude is not unique predictive importance, causal
influence, or a share of explained variance.

Ensemble output has one ordered column per component:

```text
feature | assessment_a_std_coefficient | assessment_b_std_coefficient | ...
```

Each component used its own training distribution to standardize predictors; the
columns do not share one raw-feature SD. Side-by-side values show fitted direction,
magnitude, and stability across independently trained task models. `NA` and `0`
retain their component-specific meanings. The table neither estimates an aggregate
ensemble coefficient nor changes equal-weight prediction averaging. Once a
portable model exists, this inspection needs neither Python nor a fitting backend.

## Understand saved models

Full models store their feature schema, preprocessing, calibration, fitting
settings, and validation evidence. Elastic net and PLS store linear coefficients;
RBF-SVR stores numeric support vectors and kernel parameters. Prediction uses
these saved numbers without loading the original fitting engine. A nonlinear
model does not provide linear feature coefficients.

`prepare_warm_start_model(x, omit_audit = TRUE)` makes a smaller deployment
artifact while retaining the same predictions:

```{r reduction, eval=all(vapply(c("glmnet", "pls", "e1071", "withr"), requireNamespace, logical(1), quietly=TRUE))}
deployment <- prepare_warm_start_model(algorithm_ensemble, omit_audit = TRUE)
stopifnot(identical(predict(deployment, new_features)$ensemble_mean,
  algorithm_predictions$ensemble_mean))
```

Reduction removes detailed training and validation rows. The retained summaries
can no longer be independently recalculated from that reduced object. Keep the
full evidence separately when an audit is needed. Reduction is not anonymization:
metadata and, for SVR, support-vector values still need review before sharing.
Load RDS files only from trusted sources.

Current single models use format 3 in both full and summary-only form. Legacy
model formats 1 and 2 retain their original meanings. Same-task algorithm
ensembles have their own format 1 and `artifact_type = "algorithm_ensemble"`;
cross-task ensembles use `artifact_type = "ensemble"`. These format numbers are
separate from the package version.

For precise model fields, tuning grids, weighted error/selection rules, and
calibration conventions, see `fit_warm_start_model()` and
`pairwiseLLM_warm_model`. Feature definitions are in `warm_start_feature_schema()`;
the installed Python README records extraction setup and schema hashes. In the
source repository, `data-raw/warm-start/README.md` describes model publication and
private audit review. None of these examples fits or publishes a production model.

Warm starts do not change Phase B linking or its prior, selection, and stopping
rules. See [Adaptive Linking](adaptive-linking.html) for that separate workflow.

## Citation

> Mercer, S. H. (2026). *Guide: Adaptive Warm Start* [R package vignette]. Comprehensive
> R Archive Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
