---
title: "Advanced: Submitting and Polling Multiple Batches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: Submitting and Polling Multiple Batches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# 1. Overview

This vignette demonstrates how to use **pairwiseLLM** for more complex **batch workflows**, including:

- Submitting many batches at once (e.g., across templates, providers, models, and “thinking” settings)
- Polling batches until they complete
- Downloading and parsing batch results
- Writing a **batch registry CSV** so you can safely **resume** after interruptions

The examples are based on the A/B template testing dev scripts for:

- OpenAI (via `run_openai_batch_pipeline()` + `openai_get_batch()`)
- Anthropic (via `run_anthropic_batch_pipeline()` + `anthropic_get_batch()`)
- Gemini (via low-level helpers such as `build_gemini_batch_requests()` and `gemini_create_batch()`)

> Note: All heavy API calls in this vignette are set to `eval = FALSE` so that the vignette remains CRAN-safe. You can enable them in your own project.

For basic function usage, see the companion vignette:

* [`vignette("getting-started")`](https://shmercer.github.io/pairwiseLLM/articles/getting-started.html)

For prompt evaluation and positional-bias diagnostics, see the companion vignette:

* [`vignette("prompt-template-positional-bias")`](https://shmercer.github.io/pairwiseLLM/articles/prompt-template-positional-bias.html)

# 2. Setup and API Keys

```{r setup, message=FALSE}
library(pairwiseLLM)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(stringr)
```

Required environment variables:

| Provider | Environment Variable |
|----------|----------------------|
| OpenAI   | `OPENAI_API_KEY`     |
| Anthropic | `ANTHROPIC_API_KEY` |
| Gemini    | `GEMINI_API_KEY`    |

Check which are set:

```{r}
check_llm_api_keys()
```


# 3. Example Data and Prompt Template

We use the built-in writing samples and a single trait (`overall_quality`).

```{r}
data("example_writing_samples", package = "pairwiseLLM")

td <- trait_description("overall_quality")
td
```

Default prompt template:

```{r}
tmpl <- set_prompt_template()
cat(substr(tmpl, 1, 400), "...\n")
```

Construct a modest number of pairs to keep the example light:

```{r}
set.seed(123)

pairs_all <- example_writing_samples |>
  make_pairs()

n_pairs <- min(40L, nrow(pairs_all))

pairs_forward <- pairs_all |>
  sample_pairs(n_pairs = n_pairs, seed = 123) |>
  randomize_pair_order(seed = 456)

pairs_reverse <- sample_reverse_pairs(
  pairs_forward,
  reverse_pct = 1.0,
  seed        = 789
)

get_pairs_for_direction <- function(direction = c("forward", "reverse")) {
  direction <- match.arg(direction)
  if (identical(direction, "forward")) {
    pairs_forward
  } else {
    pairs_reverse
  }
}
```


# 4. Designing the Batch Grid

Suppose we want to test several prompt templates across:

- Anthropic models (with/without “thinking”)
- OpenAI models (with/without “thinking” for specific models)
- Gemini models (with “thinking” enabled)

Here we define a small grid:

```{r}
anthropic_models <- c(
  "claude-sonnet-4-5",
  "claude-haiku-4-5",
  "claude-opus-4-5"
)

gemini_models <- c(
  "gemini-3-pro-preview"
)

openai_models <- c(
  "gpt-4.1",
  "gpt-4o",
  "gpt-5.1"
)

thinking_levels <- c("no_thinking", "with_thinking")
directions <- c("forward", "reverse")

anthropic_grid <- tidyr::expand_grid(
  provider  = "anthropic",
  model     = anthropic_models,
  thinking  = thinking_levels,
  direction = directions
)

gemini_grid <- tidyr::expand_grid(
  provider  = "gemini",
  model     = gemini_models,
  thinking  = "with_thinking",
  direction = directions
)

openai_grid <- tidyr::expand_grid(
  provider  = "openai",
  model     = openai_models,
  thinking  = thinking_levels,
  direction = directions
) %>%
  # For example, only allow "with_thinking" for gpt-5.1
  dplyr::filter(model == "gpt-5.1" | thinking == "no_thinking")

batch_grid <- dplyr::bind_rows(
  anthropic_grid,
  gemini_grid,
  openai_grid
)

batch_grid
```

We will also imagine multiple prompt templates have been registered. For simplicity, we use the same `tmpl` string, but in practice you would substitute different text:

```{r}
templates_tbl <- tibble::tibble(
  template_id     = c("test1", "test2", "test3", "test4", "test5"),
  prompt_template = list(tmpl, tmpl, tmpl, tmpl, tmpl)
)

templates_tbl
```


# 5. Phase 1: Submitting Many Batches (No Polling Yet)

We will:

1. Loop over all `(template_id, provider, model, thinking, direction)` combinations.
2. Submit a batch for each combination.
3. Record metadata (including `batch_id`) in an in-memory `jobs` list.
4. Write a **batch index CSV** to disk.

Create an output directory:

```{r, eval=FALSE}
out_dir <- "dev-output/advanced-multi-batch"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

Submit all batches:

```{r, eval=FALSE}
jobs <- list()

for (t_row in seq_len(nrow(templates_tbl))) {
  template_id <- templates_tbl$template_id[t_row]
  tmpl_string <- templates_tbl$prompt_template[[t_row]]

  for (i in seq_len(nrow(batch_grid))) {
    row <- batch_grid[i, ]

    provider <- row$provider
    model <- row$model
    thinking <- row$thinking
    direction <- row$direction

    message(
      "Submitting batch: template=", template_id,
      " | ", provider, " / ", model,
      " / ", thinking, " / ", direction
    )

    pairs_use <- get_pairs_for_direction(direction)
    is_thinking <- identical(thinking, "with_thinking")

    prefix <- paste(provider, template_id, model, thinking, direction, sep = "_")
    prefix <- gsub("[^A-Za-z0-9_.-]", "-", prefix)

    batch_input_path <- file.path(out_dir, paste0(prefix, "_input.jsonl"))
    batch_output_path <- file.path(out_dir, paste0(prefix, "_output.jsonl"))
    csv_path <- file.path(out_dir, paste0(prefix, ".csv"))

    if (identical(provider, "openai")) {
      # OpenAI: use the helpers from the dev scripts
      include_thoughts <- is_thinking && grepl("^gpt-5\\.1", model)

      pipeline <- run_openai_batch_pipeline(
        pairs             = pairs_use,
        model             = model,
        trait_name        = td$name,
        trait_description = td$description,
        prompt_template   = tmpl_string,
        include_thoughts  = include_thoughts,
        include_raw       = TRUE,
        batch_input_path  = batch_input_path,
        batch_output_path = batch_output_path,
        poll              = FALSE
      )

      jobs[[length(jobs) + 1L]] <- list(
        template_id       = template_id,
        provider          = provider,
        model             = model,
        thinking          = thinking,
        direction         = direction,
        prefix            = prefix,
        batch_type        = "openai",
        batch_id          = pipeline$batch$id,
        batch_input_path  = pipeline$batch_input_path,
        batch_output_path = batch_output_path,
        csv_path          = csv_path,
        done              = FALSE,
        results           = NULL
      )
    } else if (identical(provider, "anthropic")) {
      # Anthropic: use run_anthropic_batch_pipeline()
      reasoning <- if (is_thinking) "enabled" else "none"
      temperature_arg <- if (!is_thinking) 0 else NULL

      pipeline <- run_anthropic_batch_pipeline(
        pairs             = pairs_use,
        model             = model,
        trait_name        = td$name,
        trait_description = td$description,
        prompt_template   = tmpl_string,
        reasoning         = reasoning,
        include_thoughts  = is_thinking,
        batch_input_path  = batch_input_path,
        batch_output_path = batch_output_path,
        poll              = FALSE,
        temperature       = temperature_arg,
        include_raw       = TRUE
      )

      jobs[[length(jobs) + 1L]] <- list(
        template_id       = template_id,
        provider          = provider,
        model             = model,
        thinking          = thinking,
        direction         = direction,
        prefix            = prefix,
        batch_type        = "anthropic",
        batch_id          = pipeline$batch$id,
        batch_input_path  = pipeline$batch_input_path,
        batch_output_path = batch_output_path,
        csv_path          = csv_path,
        done              = FALSE,
        results           = NULL
      )
    } else if (identical(provider, "gemini")) {
      # Gemini: typically use low-level helpers, as in the dev scripts
      req_tbl <- build_gemini_batch_requests(
        pairs             = pairs_use,
        model             = model,
        trait_name        = td$name,
        trait_description = td$description,
        prompt_template   = tmpl_string,
        thinking_level    = "low", # example
        include_thoughts  = TRUE
      )

      batch <- gemini_create_batch(
        requests    = req_tbl$request,
        model       = model,
        api_key     = Sys.getenv("GEMINI_API_KEY"),
        api_version = "v1beta"
      )

      batch_name <- batch$name %||% stop(
        "Gemini batch did not return a `name` field.",
        call. = FALSE
      )

      jobs[[length(jobs) + 1L]] <- list(
        template_id       = template_id,
        provider          = provider,
        model             = model,
        thinking          = thinking,
        direction         = direction,
        prefix            = prefix,
        batch_type        = "gemini",
        batch_id          = batch_name,
        batch_input_path  = batch_input_path,
        batch_output_path = batch_output_path,
        csv_path          = csv_path,
        done              = FALSE,
        results           = NULL
      )
    }
  }
}
```

## 5.1 Writing a Batch Registry CSV (Important!)

To avoid losing batch IDs if your session dies, write a compact index of all jobs to disk:

```{r, eval=FALSE}
jobs_tbl <- tibble::tibble(
  idx = seq_along(jobs),
  template_id = vapply(jobs, `[[`, character(1), "template_id"),
  provider = vapply(jobs, `[[`, character(1), "provider"),
  model = vapply(jobs, `[[`, character(1), "model"),
  thinking = vapply(jobs, `[[`, character(1), "thinking"),
  direction = vapply(jobs, `[[`, character(1), "direction"),
  prefix = vapply(jobs, `[[`, character(1), "prefix"),
  batch_type = vapply(jobs, `[[`, character(1), "batch_type"),
  batch_id = vapply(jobs, `[[`, character(1), "batch_id"),
  batch_input_path = vapply(jobs, `[[`, character(1), "batch_input_path"),
  batch_output_path = vapply(jobs, `[[`, character(1), "batch_output_path"),
  csv_path = vapply(jobs, `[[`, character(1), "csv_path")
)

jobs_index_path <- file.path(out_dir, "batch_jobs_index.csv")
readr::write_csv(jobs_tbl, jobs_index_path)

jobs_index_path
```

You can now **stop R** or close RStudio safely — all critical details are in `batch_jobs_index.csv`.


# 6. Phase 2: Polling, Downloading, and Parsing

In a *new* session, you can:

1. Reload the batch index CSV  
2. Rebuild the `jobs` list  
3. Poll providers for batch status  
4. Download and parse results when complete  
5. Save per-job CSVs of parsed results  

First, helper functions for terminal states:

```{r, eval=FALSE}
is_terminal_openai <- function(status) {
  status %in% c("completed", "failed", "cancelled", "expired")
}

is_terminal_anthropic <- function(status) {
  status %in% c("ended", "errored", "canceled", "expired")
}

is_terminal_gemini <- function(state) {
  state %in% c("SUCCEEDED", "FAILED", "CANCELLED", "EXPIRED")
}
```

Now the polling loop, with a small delay between jobs to reduce 429 (rate limit) risks:

```{r, eval=FALSE}
interval_seconds <- 60
per_job_delay <- 2 # seconds between polling calls

# Reload batch index
jobs_index_path <- file.path(out_dir, "batch_jobs_index.csv")
jobs_tbl <- readr::read_csv(jobs_index_path, show_col_types = FALSE)

# Rebuild jobs list skeleton
jobs <- purrr::pmap(
  jobs_tbl,
  function(idx, template_id, provider, model, thinking, direction,
           prefix, batch_type, batch_id,
           batch_input_path, batch_output_path, csv_path, ...) {
    list(
      template_id       = template_id,
      provider          = provider,
      model             = model,
      thinking          = thinking,
      direction         = direction,
      prefix            = prefix,
      batch_type        = batch_type,
      batch_id          = batch_id,
      batch_input_path  = batch_input_path,
      batch_output_path = batch_output_path,
      csv_path          = csv_path,
      done              = FALSE,
      results           = NULL
    )
  }
)

unfinished <- which(!vapply(jobs, `[[`, logical(1), "done"))

while (length(unfinished) > 0L) {
  message("Polling ", length(unfinished), " unfinished batch(es)...")

  for (j in unfinished) {
    job <- jobs[[j]]
    if (job$done) next

    batch_type <- job$batch_type

    if (identical(batch_type, "openai")) {
      batch <- openai_get_batch(job$batch_id)
      status <- batch$status %||% "unknown"
      message("  [OpenAI] ", job$prefix, " status: ", status)

      if (is_terminal_openai(status)) {
        if (identical(status, "completed")) {
          openai_download_batch_output(
            batch_id = job$batch_id,
            path     = job$batch_output_path
          )

          res <- parse_openai_batch_output(job$batch_output_path)
          jobs[[j]]$results <- res
          readr::write_csv(res, job$csv_path)
          message("    -> Results written to: ", job$csv_path)
        }
        jobs[[j]]$done <- TRUE
      }
    } else if (identical(batch_type, "anthropic")) {
      batch <- anthropic_get_batch(job$batch_id)
      status <- batch$processing_status %||% "unknown"
      message("  [Anthropic] ", job$prefix, " status: ", status)

      if (is_terminal_anthropic(status)) {
        if (identical(status, "ended")) {
          output_path <- anthropic_download_batch_results(
            batch_id    = job$batch_id,
            output_path = job$batch_output_path
          )

          res <- parse_anthropic_batch_output(
            jsonl_path  = output_path,
            tag_prefix  = "<BETTER_SAMPLE>",
            tag_suffix  = "</BETTER_SAMPLE>"
          )

          jobs[[j]]$results <- res
          readr::write_csv(res, job$csv_path)
          message("    -> Results written to: ", job$csv_path)
        }
        jobs[[j]]$done <- TRUE
      }
    } else if (identical(batch_type, "gemini")) {
      batch <- gemini_get_batch(job$batch_id)
      state <- batch$state %||% "STATE_UNSPECIFIED"
      message("  [Gemini] ", job$prefix, " state: ", state)

      if (is_terminal_gemini(state)) {
        if (identical(state, "SUCCEEDED")) {
          raw_res <- gemini_download_batch_results(job$batch_id)

          res <- parse_gemini_batch_output(
            raw_results = raw_res,
            tag_prefix  = "<BETTER_SAMPLE>",
            tag_suffix  = "</BETTER_SAMPLE>"
          )

          jobs[[j]]$results <- res
          readr::write_csv(res, job$csv_path)
          message("    -> Results written to: ", job$csv_path)
        }
        jobs[[j]]$done <- TRUE
      }
    }

    Sys.sleep(per_job_delay)
  }

  unfinished <- which(!vapply(jobs, `[[`, logical(1), "done"))

  if (length(unfinished) > 0L) {
    message("Sleeping ", interval_seconds, " seconds before next poll...")
    Sys.sleep(interval_seconds)
  }
}

message("All batches have reached a terminal state.")
```

At the end of this loop:

- All completed batches have their outputs downloaded.
- Each job has a parsed results CSV (`csv_path`).
- You can now perform consistency and positional-bias analyses, or fit BT/Elo models.


# 7. Resuming After Interruption

If the polling loop is interrupted:

1. Restart R.
2. Reload `batch_jobs_index.csv`.
3. Rebuild `jobs` as above.
4. Recompute `unfinished` and re-enter the polling loop.

Because all of the essential metadata (provider, model, template, direction, batch IDs, file paths) is stored in the registry CSV, you can safely recover and continue.

For example:

```r
jobs_index_path <- file.path(out_dir, "batch_jobs_index.csv")
jobs_tbl <- readr::read_csv(jobs_index_path, show_col_types = FALSE)

# Rebuild jobs list as before...
# Then:
unfinished <- which(!vapply(jobs, `[[`, logical(1), "done"))

if (length(unfinished) > 0L) {
  message("Resuming polling for ", length(unfinished), " unfinished batch(es).")
  # ... re-enter the polling loop ...
} else {
  message("All jobs are already complete.")
}
```


# 8. Next Steps

Once you have per-job results CSVs (e.g., one per template × model × thinking × direction), you can:

- Compute **reverse consistency** with `compute_reverse_consistency()`
- Analyze **positional bias** with `check_positional_bias()`
- Aggregate results by provider/model/template using standard `dplyr` pipelines
- Fit **Bradley–Terry** models with `build_bt_data()` + `fit_bt_model()`
- Fit **Elo** models with `fit_elo_model()` (when `EloChoice` is installed)


# 9. Citation

> Mercer, S. H. (2025). pairwiseLLM: Pairwise writing quality comparisons with large language models (Version 1.0.0) [R package; Computer software]. https://github.com/shmercer/pairwiseLLM
