---
title: "Standalone Bayesian BTL with CmdStan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standalone Bayesian BTL with CmdStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
cmdstan_available <- identical(Sys.getenv("PAIRWISELLM_RUN_CMDSTAN_VIGNETTES"), "true") &&
  requireNamespace("cmdstanr", quietly = TRUE) &&
  tryCatch(!is.null(cmdstanr::cmdstan_version()), error = function(e) FALSE)
```

`fit_bayes_btl_mcmc()` fits a Bayesian Bradley--Terry--Luce (BTL) model to an existing fixed set of
pairwise outcomes. It reuses the adaptive fit and summary contracts but does not perform adaptive
pair selection.

## Prerequisites

The fit requires the suggested `cmdstanr` package, CmdStan, and a working C++ toolchain. Installation
is a machine-level setup and does not require provider credentials. Executable fitting chunks are
disabled during ordinary package builds. Set `PAIRWISELLM_RUN_CMDSTAN_VIGNETTES=true` to opt in
when rendering this source locally.

```{r install, eval=FALSE}
install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
)
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

```{r availability}
cmdstan_available
```

## Prepare canonical input

The public builder accepts `ID1`, `ID2`, and `better_id`, validates that each winner belongs to its
pair, and adds deterministic keys, iteration values, timestamps, and provenance columns.

```{r input}
library(pairwiseLLM)

data("example_writing_pairs", package = "pairwiseLLM")
observed <- example_writing_pairs[seq_len(min(40L, nrow(example_writing_pairs))), ]
results_tbl <- build_btl_results_data(
  observed,
  backend = "offline_fixture",
  model = "deterministic_observations"
)
ids <- sort(unique(c(results_tbl$A_id, results_tbl$B_id)))

results_tbl[, c("pair_uid", "A_id", "B_id", "better_id", "winner_pos", "phase")]
```

The four likelihood variants are:

- `"btl"`: item parameters only;
- `"btl_e"`: item parameters plus a lapse/error component;
- `"btl_b"`: item parameters plus a presentation-position effect;
- `"btl_e_b"`: both lapse and position components (the default).

Choose the variant before inspecting results, based on the study design and estimands. A more
complex variant is not automatically preferable, especially with sparse data.

## Fit and summarize

The following deterministic example is not evaluated during documentation builds. Run it after the
availability check returns `TRUE`. The small iteration count keeps the walkthrough practical and is
for workflow demonstration, not production inference.

```{r fit, eval=FALSE, purl=TRUE}
fit <- fit_bayes_btl_mcmc(
  results = results_tbl,
  ids = ids,
  model_variant = "btl_e_b",
  cmdstan = list(
    chains = 2L,
    parallel_chains = 2L,
    iter_warmup = 250L,
    iter_sampling = 250L,
    seed = 7007L,
    core_fraction = 1
  )
)

refits <- summarize_refits(fit)
items <- summarize_items(fit)

refits[, c(
  "round_id", "total_pairs", "diagnostics_pass",
  "divergences", "max_rhat", "min_ess_bulk"
)]
items[, c("ID", "theta_mean", "theta_sd", "rank_mean", "deg")]
```

The item scale is relative and identified by the model constraints; its absolute origin is not an
external score. Posterior SDs and intervals are conditional on the selected model and observed
judgments. Check divergences, R-hat, effective sample size, and sampling warnings before interpreting
ranks. A failed diagnostic is not repaired by hiding the warning or reporting only posterior means.

## Cumulative refits

`pair_counts` fits increasing subsets of the same data and records one round/item-log view per
subset. `subset_method = "first"` follows row order. `"sample"` creates one seeded permutation and
uses nested prefixes, so provide `seed` for reproducibility.

```{r cumulative, eval=FALSE, purl=TRUE}
cumulative <- fit_bayes_btl_mcmc(
  results_tbl,
  ids = ids,
  pair_counts = c(20L, 40L),
  subset_method = "sample",
  seed = 7007L,
  cmdstan = list(chains = 2L, parallel_chains = 2L, seed = 7007L)
)

summarize_refits(cumulative)
summarize_items(cumulative, refit_id = 2L)
```

## Files, failures, and reproducibility

CmdStan writes generated C++/executables and sampling CSV files to its cache or configured
`cmdstan$output_dir`. Preserve the seed, model variant, CmdStan configuration, package version, and
diagnostics with reported results. Missing CmdStan, invalid schemas, unknown IDs, invalid winners,
empty data, invalid subset sizes, and unsupported variants abort explicitly. Sampling failures and
poor diagnostics should be investigated rather than converted to partial rankings.

For adaptive selection and stopping, use [Guide: Adaptive Pairing](adaptive-pairing.html). For
fixed-pair frequentist alternatives, see `fit_bt_model()` and `fit_elo_model()`.

## Citation

> Mercer, S. H. (2026). *Standalone Bayesian BTL with CmdStan* [R package vignette]. Comprehensive
> R Archive Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
