---
title: "Bootstrap bias correction for BT scores"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Bootstrap bias correction for BT scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     eval = requireNamespace("withr", quietly = TRUE))
library(pairwiseLLM)
```

## What the correction estimates

The [integrated CJ workflow](adaptive-cj-workflow.html) explains where bootstrap
bias correction fits alongside pair selection, internal reliability, score
recovery, and held-out prediction. This guide focuses on the bootstrap interface.
The examples require the suggested `withr` package and are displayed without
execution when it is absent.

A Bradley–Terry (BT) score describes relative strength on a log-odds scale.
With limited comparisons, an estimator may systematically place scores too far
apart or too close together. A **parametric bootstrap** estimates this bias by
simulating new judgments from the fitted scores and fitting the same estimator
to each simulated dataset. It makes no provider calls.

Pair selection matters. For a random schedule chosen independently of results,
the bootstrap can keep the pairs fixed. For an adaptive schedule, later pairs
depend on earlier winners. Each replicate must repeat that selection process
using its own simulated winners. Reusing only the observed adaptive pairs would
answer a different question.

The function aligns item IDs and centers every score vector to sum to zero.
For each item it calculates:

```
estimated bias = average bootstrap estimate - initial estimate
corrected score = initial estimate - estimated bias
```

Zero remains a relative reference, not a passing score. The correction can change
rankings, and it does not guarantee improvement for every item or application.

## A fixed random schedule

This synthetic example uses a connected starting chain and additional random
pairs, all chosen before simulating any winners. Alpha adjustment is used here
so the example needs no optional fitting engine; the penalty is specified in
advance. Firth (`engine = "brglm2"`) is also supported for nonadaptive schedules
when that optional package is installed.

```{r fixed-bootstrap}
ids <- letters[1:4]
truth <- setNames(c(-0.8, -0.2, 0.2, 0.8), ids)
dat <- withr::with_seed(304, {
  extra <- replicate(21, sample(ids, 2))
  pairs <- data.frame(object1 = c(ids[1:3], extra[1, ]),
                      object2 = c(ids[2:4], extra[2, ]))
  p <- plogis(truth[pairs$object1] - truth[pairs$object2])
  pairs$result <- as.integer(runif(nrow(pairs)) < p)
  pairs
})
fit <- fit_bt_model(dat, engine = "alpha", alpha = 0.5, verbose = FALSE)
fixed <- bootstrap_bt_model(fit, mode = "fixed", n_rep = 20, seed = 1304)
fixed$theta
fixed$provenance[c("mode", "seed", "pairing_strategy", "refit", "versions")]
```

The fitted object supplies both the generating scores and the estimator
settings. Each replicate keeps the original ordered pair rows, including any
repetitions, and draws fresh winners. Twenty replicates keep this example quick;
that number is not a recommendation for a study.

## An adaptive p50 schedule

Save the initial state **before collecting any outcomes**. Here the p50 strategy
chooses partners with current TrueSkill win probabilities near one half.
The synthetic judge below supplies winners in place of a provider.

```{r adaptive-bootstrap}
ids <- letters[1:6]
truth <- setNames(seq(-1, 1, length.out = 6), ids)
initial <- adaptive_rank_start(ids, seed = 17,
  adaptive_config = list(pairing_strategy = "trueskill_p50"))
judge <- function(A, B, state, ...) {
  p <- plogis(truth[[A$item_id]] - truth[[B$item_id]])
  list(is_valid = TRUE, Y = as.integer(runif(1) < p))
}
# This demonstration design has no Bayesian refit within its 12-comparison budget.
schedule_config <- list(refit_pairs_target = 100L)
observed <- withr::with_seed(7304, adaptive_rank_run_live(initial, judge,
  n_steps = 12, btl_config = schedule_config, progress = "none"))
dat <- adaptive_results_history(observed)
fit <- fit_bt_model(dat, engine = "alpha", alpha = 0.5, verbose = FALSE)
adaptive <- bootstrap_bt_model(fit, mode = "adaptive", n_rep = 4, seed = 2304,
  initial_state = initial, budget = 12, btl_config = schedule_config)
adaptive$theta
adaptive$replicates[, c("replicate", "success", "n_comparisons", "schedule_digest")]
adaptive$provenance[c("mode", "seed", "pairing_strategy", "refit", "versions")]
```

The initial connected pairs and their presentations stay fixed across
replicates. After initialization, each replicate has a new deterministic
selector seed and its own simulated outcomes. The selector and state updates
are the same ones used by the ordinary adaptive runner. Four replicates here
illustrate the workflow only; their Monte Carlo error can be substantial.

The same interface supports `trueskill_pollitt`, `hybrid`, and `random` states.
Keep the original scheduling-refit settings when reproducing a design. Hybrid
routing can change after Bayesian refits establish a sufficiently identified
global scale. Those refits are retained and may require CmdStan. The final
bootstrap `estimator` is a separate choice from the scheduling fitter.
An outcome-independent predictive prior may be included in the initial state.
Sparse reservoirs restrict selection to their legal pairs; their old winners
are discarded and new binary outcomes are simulated.

The budget counts successful comparisons, including the initial connected
structure. Statistical stopping is recorded but does not truncate this
fixed-budget experiment. If legal selection is exhausted, the replicate fails.
Small panels can exhaust hybrid exposure constraints before using every possible
pair; the bootstrap does not relax these constraints to fill the budget.

## Reading the results and managing cost

`bias` estimates systematic error under the fitted generating model.
`bootstrap_sd` describes how much the replicate estimates vary. `mcse_bias`
describes simulation error in the estimated bias and usually decreases as the
replicate count increases. Neither quantity is automatically a standard error
for the corrected score. The function does not provide confidence intervals or
calculate corrected scale-separation reliability (SSR).

Inspect `n_success`, `n_failed`, and `replicates` before using a correction.
By default every requested replicate must succeed. You may choose a lower
`min_success` explicitly, but excluded failures can change the interpretation:
the average then describes successful refits. An unmet threshold raises an
error whose `result` field retains diagnostics, with missing corrected scores.
No failed estimates are silently included and no other estimator is substituted.

Each replicate costs a schedule simulation and final refit; hybrid can also
require multiple Bayesian fits. Use `workers = 2` for parallel execution with
the optional `future` and `future.apply` packages. Reproducible seeds and ordered
aggregation give the same scientific results as serial execution. Avoid
oversubscribing CPUs with both bootstrap workers and many sampler chains.
Custom callbacks must honor supplied seeds and avoid mutable external state.

The default `keep = "summary"` discards large replicate states. Use `"theta"`
to retain score draws, or `"full"` when individual schedules or detailed numerical
failure diagnostics need inspection. Detailed failures can contain large matrices.
Save the returned object with `saveRDS()` to retain inputs, settings, seeds,
software versions and diagnostics. There is no checkpoint/resume interface;
restarting with the same inputs and compatible software repeats the run.

## Citation

> Mercer, S. H. (2026). *Bootstrap bias correction for BT scores* [R package vignette]. Comprehensive R
> Archive Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
