---
title: "Getting Started with pairwiseLLM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pairwiseLLM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# 1. Introduction

`pairwiseLLM` provides a unified workflow for generating and analyzing **pairwise comparisons of writing quality** using modern LLM APIs (OpenAI, Anthropic, Gemini, and soon local models).

A typical workflow:

1. Select writing samples  
2. Construct pairwise comparison sets  
3. Submit comparisons to an LLM (live or batch API)  
4. Parse model outputs  
5. Fit Bradley–Terry or Elo models to obtain latent writing-quality scores  

For prompt evaluation and positional-bias diagnostics, see the companion vignette:

👉 `vignette("prompt-template-positional-bias")`

---

# 2. Setting API Keys

`pairwiseLLM` reads provider keys **only from environment variables**, never from R options or global variables.

| Provider | Environment Variable |
|----------|----------------------|
| OpenAI   | `OPENAI_API_KEY`     |
| Anthropic | `ANTHROPIC_API_KEY` |
| Gemini    | `GEMINI_API_KEY`    |

You should put these in your `~/.Renviron`:

```
OPENAI_API_KEY=sk-...
ANTHROPIC_API_KEY=...
GEMINI_API_KEY=...
```

Check which keys are available:

```{r}
library(pairwiseLLM)
check_llm_api_keys()
```

---

# 3. Example Writing Data

The package ships with 20 authentic student writing samples:

```{r}
data("example_writing_samples", package = "pairwiseLLM")
dplyr::slice_head(example_writing_samples, n = 3)
```

Each sample has:

- `id`  
- `text`  

---

# 4. Constructing Pairwise Comparisons

Create all unordered pairs:

```{r}
pairs <- example_writing_samples |>
  make_pairs()

dplyr::slice_head(pairs, n = 5)
```

Sample a subset of pairs:

```{r}
pairs_small <- sample_pairs(pairs, n_pairs = 10, seed = 123)
```

Randomize SAMPLE_1 / SAMPLE_2 order:

```{r}
pairs_small <- randomize_pair_order(pairs_small, seed = 99)
```

---

# 5. Traits and Prompt Templates

## 5.1 Using a built-in trait

```{r}
td <- trait_description("overall_quality")
td
```

Or define your own:

```{r}
td_custom <- trait_description(
  custom_name = "Clarity",
  custom_description = "How clearly and effectively ideas are expressed."
)
```

## 5.2 Using or customizing prompt templates

Load default prompt:

```{r}
tmpl <- set_prompt_template()
cat(substr(tmpl, 1, 300))
```

Placeholders required:

- `{TRAIT_NAME}`
- `{TRAIT_DESCRIPTION}`
- `{SAMPLE_1}`
- `{SAMPLE_2}`

Load a template from file:

```{r, eval=FALSE}
set_prompt_template(file = "my_template.txt")
```

---

# 6. Live Pairwise Comparisons

The unified wrapper works for **OpenAI, Anthropic, and Gemini**.

```{r, eval=FALSE}
res_live <- submit_llm_pairs(
  pairs             = pairs_small,
  backend           = "openai",   # also "anthropic" or "gemini"
  model             = "gpt-4o",
  trait_name        = td$name,
  trait_description = td$description,
  prompt_template   = tmpl
)
```

Preview results:

```{r, eval=FALSE}
dplyr::slice_head(res_live, 5)
```

Each row includes:

- `pair_id`
- `sample1_id`, `sample2_id`
- parsed `<BETTER_SAMPLE>` tag → `better_sample` and `better_id`
- (optionally) raw model output

---

# 7. Preparing Data for BT or Elo Modeling

Convert LLM output to a 3-column BT dataset:

```{r, eval=FALSE}
bt_data <- build_bt_data(res_live)
dplyr::slice_head(bt_data, 5)
```

---

# 8. Bradley–Terry Modeling

Fit model:

```{r, eval=FALSE}
bt_fit <- fit_bt_model(bt_data)
```

Summarize results:

```{r, eval=FALSE}
summarize_bt_fit(bt_fit)
```

The output includes:

- latent θ ability scores  
- SEs  
- reliability (sirt engine)  

---

# 9. Elo Modeling (Optional)

Requires **EloChoice** (in Suggests):

```{r, eval=FALSE}
elo_fit <- fit_elo_model(bt_data, runs = 5)
elo_fit
```

Outputs:

- Elo ratings for each sample  
- unweighted and weighted reliability  
- trial counts  

---

# 10. Batch APIs (Large Jobs)

## 10.1 Submit a batch

```{r, eval=FALSE}
batch <- llm_submit_pairs_batch(
  provider           = "openai",
  model              = "gpt-4o",
  pairs              = pairs_small,
  trait_name         = td$name,
  trait_description  = td$description,
  prompt_template    = tmpl
)
```

## 10.2 Download results

```{r, eval=FALSE}
res_batch <- llm_download_batch_results(batch)
head(res_batch)
```

---

# 11. Backend-Specific Tools

Most users use the unified interface, but backend helpers are available.

### 11.1 OpenAI
- `submit_openai_pairs_live()`
- `build_openai_batch_requests()`
- `run_openai_batch_pipeline()`
- `parse_openai_batch_output()`

### 11.2 Anthropic
- `submit_anthropic_pairs_live()`
- `build_anthropic_batch_requests()`
- `run_anthropic_batch_pipeline()`
- `parse_anthropic_batch_output()`

### 11.3 Google Gemini
- `submit_gemini_pairs_live()`
- `build_gemini_batch_requests()`
- `run_gemini_batch_pipeline()`
- `parse_gemini_batch_output()`

---

# 12. Troubleshooting

### Missing API keys  
```{r}
check_llm_api_keys()
```

### Getting chain-of-thought leakage  
Use the default template or set `include_thoughts = FALSE`.

### Timeouts  
Use batch APIs for >40 pairs.

### Positional bias  
Use `test_positional_bias()` or see the bias-testing vignette.

---

# 13. Further Reading

To evaluate positional bias across models and templates:

```r
vignette("prompt-template-positional-bias")
```

To explore modeling:

```r
?fit_bt_model
?fit_elo_model
```

---

# 14. Citation

> Mercer, S. H. (2025). *pairwiseLLM: Pairwise comparisons of writing quality with LLMs* (R package).  
> GitHub: https://github.com/shmercer/pairwiseLLM
