---
title: "Getting Started with pairwiseLLM"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting Started with pairwiseLLM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(pairwiseLLM)
```

## What you will do

`pairwiseLLM` turns judgments about **which of two writing samples is better**
into relative writing-quality scores. The aspect being judged, such as overall
quality or organization, is called a **trait**. Comparing samples in this way is
also called **comparative judgment (CJ)**.

This guide assumes you can run R code and work with a data frame. First, you
will analyze bundled synthetic comparisons without an API key or provider
charges. Then you will prepare samples and see how to collect new comparisons
using an LLM. All code through “Prepare a judging prompt” runs locally; cloud
requests later in the guide are shown but are not executed when this page builds.

## Install the tools for this example

Use **R 4.4 or later**. Install the package and the optional `sirt` modeling
engine once, then load pairwiseLLM in each new R session:

```{r installation, eval=FALSE}
install.packages(c("pairwiseLLM", "sirt"))
library(pairwiseLLM)
```

You do not need CmdStan, Python, or an LLM account for this first example.
Bayesian and adaptive analyses have additional setup described in their guides.

## Your first result: scores from bundled comparisons

The package includes 190 comparison outcomes covering every unordered pair of
20 synthetic writing samples. These outcomes were generated for demonstration;
they are not new LLM judgments or evidence of assessment validity.

```{r offline-data}
data("example_writing_pairs", package = "pairwiseLLM")
head(example_writing_pairs, 4)
```

`ID1` and `ID2` identify the two samples. `better_id` identifies the winner.
`build_bt_data()` converts these winners into the format needed by a
Bradley–Terry (BT) model. The model estimates relative scores from the pattern
of wins and losses.

```{r offline-fit, eval=requireNamespace("sirt", quietly=TRUE)}
bt_data <- build_bt_data(example_writing_pairs)
fit <- fit_bt_model(bt_data, engine = "sirt", verbose = FALSE)
scores <- fit$theta |> dplyr::arrange(dplyr::desc(theta))
head(scores, 5)
```

If `sirt` is missing, run the installation command above before this block.
Choosing the engine explicitly makes this example consistent across machines;
`fit_bt_model()` also supports an automatic choice and `BradleyTerry2`.

### Read the scores

| Column | Meaning | How to use it |
| --- | --- | --- |
| `ID` | The sample identifier | Match the result to your original text. |
| `theta` | Estimated relative writing quality | Higher values indicate stronger writing on the assessed trait. |
| `se` | Standard error of the estimate | Larger values mean greater model uncertainty. |

The displayed rows are ordered by score. The full `scores` table has 20 rows.
A difference between two estimates is not automatically a meaningful difference:
consider their uncertainty and the quality of the judgments.

Zero is a location on a relative scale, not a pass mark. These values are not
percentages or rubric grades, and independently fitted cohorts do not share a
common scale. For rubric categories, use the
[rubric calibration guide](rubric-calibration.html), which requires completed
**Bayesian** CJ results; the frequentist fit above is not an accepted input.

### Save an analysis

```{r save-offline-results, eval=FALSE}
write.csv(scores, "writing_scores.csv", row.names = FALSE)
saveRDS(fit, "writing_bt_fit.rds")
# In a later session: fit <- readRDS("writing_bt_fit.rds")
```

The CSV is convenient for viewing scores in a spreadsheet. The RDS preserves
the fitted R object. These paths are relative to your current working directory.

## Prepare your own writing samples

Start with one row per sample, a unique, non-missing `ID`, and a `text` column.
Keep IDs stable so that saved results continue to refer to the same samples.
Do not include names or other identifying information merely to label rows.

```{r personal-data}
my_samples <- data.frame(
  ID = c("essay_01", "essay_02", "essay_03"),
  text = c(
    "Writing helps us explain our ideas to others.",
    "We write to communicate, remember, and develop an argument.",
    "Writing is useful. I use it at school."
  )
)
samples <- read_samples_df(my_samples)
```

For a spreadsheet saved as CSV, map its columns to the package's names:

```{r csv-data, eval=FALSE}
my_data <- read.csv("writing.csv", colClasses = "character")
samples <- read_samples_df(my_data, id_col = "student_code", text_col = "response")
```

These three short texts illustrate the data format only. For the collection
example below, use the bundled 20 samples so every object is defined:

```{r collection-data}
data("example_writing_samples", package = "pairwiseLLM")
samples <- read_samples_df(example_writing_samples[, c("ID", "text")])
all_pairs <- make_pairs(samples)
nrow(all_pairs)

pairs_small <- all_pairs |>
  sample_pairs(n_pairs = 10, seed = 123) |>
  randomize_pair_order(seed = 99)
```

Pair construction makes no provider calls. Twenty samples produce
`choose(20, 2) = 190` unordered pairs; the total grows quickly as you add samples.
Randomizing presentation helps distribute which sample appears first.

**The ten-pair subset is a collection demonstration, not a complete ranking
study.** Random sampling does not guarantee that every sample appears or that
all samples are connected through comparisons. The offline fit above uses the
full bundled comparison set. Plan coverage and diagnostics before fitting your
own collected data; [adaptive pairing](adaptive-pairing.html) is an alternative
that selects comparisons as evidence accumulates.

## Prepare a judging prompt

A prompt combines the trait, the judging instructions, and the two texts.
Start with a built-in trait and template:

```{r prompt-setup}
td <- trait_description("overall_quality")
tmpl <- get_prompt_template("default")
td$name
```

You can inspect exactly what the model will receive before paying for a call:

```{r preview-prompt}
prompt <- build_prompt(
  template = tmpl,
  trait_name = td$name,
  trait_desc = td$description,
  text1 = pairs_small$text1[[1]],
  text2 = pairs_small$text2[[1]]
)
cat(substr(prompt, 1, 350))
```

This shows only the beginning; use `cat(prompt)` to read the full prompt locally.
See [Data Schemas and Prompt Management](data-and-prompts.html) for custom traits,
file templates, named registration, and required placeholders.

## Collect a small set of new judgments

The remaining submission examples make **cloud requests that may incur charges**
when you run them. They send the prompt and sample text to your chosen provider.
Review its privacy and retention terms before submitting sensitive data.

### Configure one provider

A **backend** selects the service; `model` selects a model offered by that
service. This example uses OpenAI. Use an identifier available to your account;
example identifiers are not a guarantee of current availability. See
[Backends and Tested Model Configurations](model-compatibility.html) for the
provider/key table and official catalogs.

For the current R session:

```{r api-key, eval=FALSE}
Sys.setenv(OPENAI_API_KEY = "your-key-here")
check_llm_api_keys()
```

`has_key = TRUE` means R can see a nonempty value; it does not test whether the
key is valid or the account can access a model. Missing keys for other providers
are expected. Never put a real key in a shared script or report.

For persistent local setup, put `OPENAI_API_KEY="your-key-here"` in your
`~/.Renviron` file and restart R. If you use `usethis`,
`usethis::edit_r_environ()` opens that file. This is a file you choose to store;
the package does not save credentials for you. Ollama uses a local server and
does not require a provider API key.

### Submit the pairs

```{r live-submission, eval=FALSE}
res_list <- submit_llm_pairs(
  pairs = pairs_small,
  backend = "openai",
  model = "gpt-4o",
  trait_name = td$name,
  trait_description = td$description,
  prompt_template = tmpl,
  parallel = FALSE,
  save_path = "live_results.csv"
)
```

The run processes pairs sequentially and saves completed results incrementally.
Use a distinct file for each study/model/trait configuration. To continue the
same interrupted run, keep its inputs and saved file and repeat the call;
matching saved pairs are skipped. Parallel execution is optional and requires
`future` and `future.apply`; leave it off for a first run.

### Check what succeeded

```{r inspect-live, eval=FALSE}
head(res_list$results)
res_list$failed_pairs
res_list$failed_attempts
```

| Component | What it tells you |
| --- | --- |
| `results` | Comparisons with a valid winner. `better_id` identifies the winning sample. |
| `failed_pairs` | Scheduled pairs with no valid outcome. |
| `failed_attempts` | Attempt-level problems, including failures followed by successful retries. |

An HTTP success alone is not a valid judgment. Inspect unresolved pairs before
analysis; missing winners are not ties. The
[recovery guide](provider-controls-and-recovery.html) explains retrying failures
and keeping successful work.

Once you have collected an adequate comparison design and inspected failures,
prepare its successful outcomes with `build_bt_data(res_list$results)` and use
the same fit-and-inspect steps as the offline example. Do not interpret the
small submission demonstration as an adequate design for all 20 samples.

## Choose the next workflow

| Your goal | Guide or function |
| --- | --- |
| More control over data and judging instructions | [Data and prompts](data-and-prompts.html) |
| Estimate token costs with a paid pilot; configure or troubleshoot providers | [Provider controls and recovery](provider-controls-and-recovery.html) |
| Submit fixed pairs for later retrieval, including resumable large jobs | [Batch workflows](advanced-batch-workflows.html) |
| Fit Elo ratings instead of BT scores | `build_elo_data()` and `fit_elo_model()` (optional `EloChoice`) |
| Estimate Bayesian scores and inspect convergence | [Bayesian BTL](bayesian-btl.html) (requires CmdStan) |
| Let the next pair depend on earlier judgments | [Adaptive pairing](adaptive-pairing.html) (requires CmdStan) |
| Convert completed Bayesian scores to ordered rubric levels | [Rubric calibration](rubric-calibration.html) |
| Compare separately evaluated cohorts on a shared scale | [Adaptive linking](adaptive-linking.html) |
| Check whether presentation order affects judgments | [Prompt bias testing](prompt-template-bias.html) |

Live processing is useful for interactive work. Batch processing can suit larger
fixed-pair jobs when you can wait for completion; support and pricing depend on
the provider. There is no universal pair-count threshold for choosing between them.

## Common first-run problems

| Problem | First check |
| --- | --- |
| The offline model says `sirt` is missing | Run `install.packages("sirt")` once, then rerun the fit. |
| R cannot see my key | Restart R after editing `.Renviron`, then use `check_llm_api_keys()`. |
| The provider rejects the model or a setting | Check account access and the provider's current catalog; start with minimal controls. |
| Some comparisons are missing | Inspect `failed_pairs` and `failed_attempts`; keep successful results. |
| Scores look unlike familiar grades | BT scores are relative estimates; rubric scoring is a separate Bayesian workflow. |
| A batch is still pending | Poll or resume the existing job; do not submit it again merely because it is unfinished. |

For a reproducible issue report, include your code with credentials removed, the
error, provider/model, operating system, and `sessionInfo()` output.

## Citation

> Mercer, S. H. (2026). *Getting started with pairwiseLLM* [R package vignette]. Comprehensive R
> Archive Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
