---
title: "Explicit-evidence linking sessions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explicit-evidence linking sessions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

E1--E3 estimate a separately centered hub shape, a separately centered spoke
shape, and a spoke offset. Choose an estimator explicitly: `fixed_shape_offset`
(E1), `gaussian_posterior_bridge` (E2), or `joint_offset` (E3). These APIs accept
explicit evidence independently of adaptive pair selection or stopping.

```{r}
library(pairwiseLLM)
hub <- list(set_id = "hub", items = data.frame(item_id = c("h1", "h2")))
spoke <- list(set_id = "spoke", items = data.frame(item_id = c("s1", "s2")))
cross <- data.frame(observation_id = c("cross-1", "cross-2"),
  A_set = "hub", A_item = c("h1", "h2"),
  B_set = "spoke", B_item = c("s1", "s2"), y_A = c(0L, 1L))
phase_a <- list(hub = list(points = c(h1 = -1, h2 = 1)),
  spoke = list(points = c(s1 = -.5, s2 = .5)))
judge <- list(beta = 0, epsilon = 0, model_variant = "btl",
  link = "logit", source = "frozen Phase A")
input <- prepare_link_input("fixed_shape_offset", hub, spoke, phase_a,
  cross[1, ], judge)
state <- start_link_session(input)
print(state)
summarize_items(state)
```

E1 consumes only point summaries; its uncertainty is conditional on those fixed
shapes. E2 consumes full posterior item draws, and E3 consumes exact raw
within-set outcomes. Both include shape and offset uncertainty. Historical
Phase A artifacts can be supplied with `list(artifact = artifact)` if they
contain the method's required payload. Summaries cannot substitute for E2 draws
or E3 raw rows. The artifact's exact hash is retained, while irrelevant evidence
is excluded from the normalized statistical input.

`theta_link_mean` is a posterior mean for E1 quadrature and E3-MCMC and a MAP
location for E2/E3 Laplace. `theta_link_eap` remains an identical compatibility
alias. Every item carries estimator ID, uncertainty scope, SD, interval and rank;
unavailable uncertainty is `NA`.

```{r}
path <- tempfile(fileext = ".rds")
save_link_session(state, path)
restored <- load_link_session(path, input = input)
stopifnot(identical(state, restored))
stopifnot(identical(resume_link_session(restored, input), restored))
next_input <- prepare_link_input("fixed_shape_offset", hub, spoke, phase_a,
  cross, judge)
continued <- resume_link_session(restored, next_input)
summary(continued)[c("estimator_id", "delta_spoke_mean", "delta_spoke_sd",
  "phase_b_active_edges_used", "uncertainty_scope", "fit_valid", "status")]
unlink(path)
```

Resume preserves exact estimator/version, canonical item order, configuration,
Phase A evidence/artifact identity and the old cross-evidence prefix. Only new
cross observations can be appended. Changing the estimator, numerical controls,
judge or Phase A requires a fresh session. Continuation restores numerical
warm-start data, never a posterior-as-prior update. Passing the identical input
is a no-op, including for explicitly selected E3-MCMC. Saving MCMC sessions
preserves draws and prediction data without requiring the original CmdStan files.

`save_adaptive_session()` and `load_adaptive_session()` also accept these sessions
using a dedicated `link-session.rds` file. Session directories cannot mix the two
formats. `adaptive_get_logs()` returns the common link-stage log. Source counts
are distinct from raw likelihood rows consumed (zero Phase A rows for E1/E2).
Unknown external source hashes/counts remain missing. Computed payload and
configuration hashes are always retained.

For multiple spokes, pass a list of prepared inputs to `start_link_session()`.
They must share the exact hub evidence, hub identities, judge and estimator.
Each result retains its own hub posterior; `summarize_items()` includes
`link_spoke_id` to keep those distinct. No cross-spoke covariance or pooled hub
posterior is invented. Status (`active`, `probe`, `frozen`) is externally supplied
reporting metadata, not a new stopping rule. Held-out probe outcomes must never
be included in the estimator's active `cross` table.

For rubric transport, fit `fit_rubric_calibration()` with the labeled Phase A hub
artifact and `calibration_design = "linked_anchors"`. Prepare linking inputs with
that exact hub artifact, then pass the common result or session to `predict()`.
Global item IDs are required. Only spoke scores are returned. The frozen hub
calibration is reused, with the original hub mean restoring the reference origin;
no target labels, cohort rescaling or updated hub posterior refit the calibration.
Invalid or unidentified links fail. The prediction's `linking` attribute retains
estimator identity, uncertainty scope, evidence provenance and diagnostics.

Saved anchored-joint Phase B sessions are unsupported and fail with a restart
message. Compatible historical Phase A artifacts remain reusable. Adaptive
selection integration and removal of the remaining old execution code are
separate epic work; use the explicit-evidence APIs above for E1--E3 sessions.

## Adaptive selection restriction

Adaptive Phase B D-optimal execution is unavailable for all E1--E3 estimators,
including the E3-MCMC audit engine, until the separate selector study validates
an information criterion. Legacy D-optimal aliases also fail explicitly. The
availability of posterior covariance does not validate a selector, and there
is no fallback to anchored-joint. Phase A and ordinary within-set ranking are
still available. Use the explicit-evidence workflow above for fixed evidence
prefixes, without candidate generation, adaptive utility, or automatic stopping.

Keep held-out outcomes outside `input$cross`. A session's `status = "probe"`
is a reporting label, not an evidence partition: it does not remove rows from
the likelihood. Probe observation IDs and both orientations of each base pair
must remain disjoint from active evidence. Original source-data partitions
remain the caller's responsibility when Phase A supplies only summaries/draws.

Internal orchestration adapters now resolve a common result for one explicit
spoke. They supply linked means, `predict_link()` probabilities, named
free-coordinate covariance, and the gradient of the ordered latent contrast
`theta_A - theta_B`. This gradient is not the derivative of a posterior-average
probability. E1 exposes only the offset coordinate; E2/E3 expose centered shape
coordinates and offset. No covariance is converted into a selection criterion.
Adaptive item adapters require explicit unique `global_item_id` values.

Probe and stopping assessments retain estimator ID/version, spoke and hub IDs,
uncertainty scope, and active/probe evidence hashes. Probe metrics use the common
prediction API in the actual presentation orientation. Reversing presentation
is not complementary when position bias is nonzero. Lagged comparisons require
the same spoke, frozen inputs and an unchanged cumulative evidence prefix.

The internal stopping adapter reports **per-refit criteria**, not a final
stopping decision or automatic session transition. Existing thresholds and
probe quality diagnostics and the active-item reliability domain (all spoke
items plus directly evidenced hub items) are preserved; the existing rolling-window controller
remains separate. Invalid or prior-only fits, missing lag history, and missing
required uncertainty block the criteria. E1's conditional uncertainty is never
reported as full Phase A uncertainty or used to pass full-uncertainty reliability.
Multiple spokes retain distinct hub posteriors and histories; they are never
averaged into one implicit hub fit. These interfaces do not constitute scientific
validation of adaptive selection or stopping for the new estimators.

## Citation

Mercer, S. H. (2026). Explicit-evidence linking sessions. In *pairwiseLLM*
[R package vignette]. Comprehensive R Archive Network.
<https://doi.org/10.32614/CRAN.package.pairwiseLLM>
