---
title: "Explicit-evidence linking sessions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explicit-evidence linking sessions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

E1--E3 estimate a separately centered hub shape, a separately centered spoke
shape, and a spoke offset. Choose an estimator explicitly: `fixed_shape_offset`
(E1), `gaussian_posterior_bridge` (E2), or `joint_offset` (E3). These APIs accept
explicit evidence independently of adaptive pair selection or stopping.

```{r}
library(pairwiseLLM)
hub <- list(set_id = "hub", items = data.frame(item_id = c("h1", "h2")))
spoke <- list(set_id = "spoke", items = data.frame(item_id = c("s1", "s2")))
cross <- data.frame(observation_id = c("cross-1", "cross-2"),
  A_set = "hub", A_item = c("h1", "h2"),
  B_set = "spoke", B_item = c("s1", "s2"), y_A = c(0L, 1L))
phase_a <- list(hub = list(points = c(h1 = -1, h2 = 1)),
  spoke = list(points = c(s1 = -.5, s2 = .5)))
judge <- list(beta = 0, epsilon = 0, model_variant = "btl",
  link = "logit", source = "frozen Phase A")
input <- prepare_link_input("fixed_shape_offset", hub, spoke, phase_a,
  cross[1, ], judge)
state <- start_link_session(input)
print(state)
summarize_items(state)
```

E1 consumes only point summaries; its uncertainty is conditional on those fixed
shapes. E2 consumes full posterior item draws, and E3 consumes exact raw
within-set outcomes. Both include shape and offset uncertainty. Historical
Phase A artifacts can be supplied with `list(artifact = artifact)` if they
contain the method's required payload. Summaries cannot substitute for E2 draws
or E3 raw rows. The artifact's exact hash is retained, while irrelevant evidence
is excluded from the normalized statistical input.

`theta_link_mean` is a posterior mean for E1 quadrature and E3-MCMC and a MAP
location for E2/E3 Laplace. `theta_link_eap` remains an identical compatibility
alias. Every item carries estimator ID, uncertainty scope, SD, interval and rank;
unavailable uncertainty is `NA`.

```{r}
path <- tempfile(fileext = ".rds")
save_link_session(state, path)
restored <- load_link_session(path, input = input)
stopifnot(identical(state, restored))
stopifnot(identical(resume_link_session(restored, input), restored))
next_input <- prepare_link_input("fixed_shape_offset", hub, spoke, phase_a,
  cross, judge)
continued <- resume_link_session(restored, next_input)
summary(continued)[c("estimator_id", "delta_spoke_mean", "delta_spoke_sd",
  "phase_b_active_edges_used", "uncertainty_scope", "fit_valid", "status")]
unlink(path)
```

Resume preserves exact estimator/version, canonical item order, configuration,
Phase A evidence/artifact identity and the old cross-evidence prefix. Only new
cross observations can be appended. Changing the estimator, numerical controls,
judge or Phase A requires a fresh session. Continuation restores numerical
warm-start data, never a posterior-as-prior update. Passing the identical input
is a no-op, including for explicitly selected E3-MCMC. Saving MCMC sessions
preserves draws and prediction data without requiring the original CmdStan files.

`save_adaptive_session()` and `load_adaptive_session()` also accept these sessions
using a dedicated `link-session.rds` file. Session directories cannot mix the two
formats. `adaptive_get_logs()` returns the common link-stage log. Source counts
are distinct from raw likelihood rows consumed (zero Phase A rows for E1/E2).
Unknown external source hashes/counts remain missing. Computed payload and
configuration hashes are always retained.

For multiple spokes, pass a list of prepared inputs to `start_link_session()`.
They must share the exact hub evidence, hub identities, judge and estimator.
Each result retains its own hub posterior; `summarize_items()` includes
`link_spoke_id` to keep those distinct. No cross-spoke covariance or pooled hub
posterior is invented. Status (`active`, `probe`, `frozen`) is externally supplied
reporting metadata, not a new stopping rule. Held-out probe outcomes must never
be included in the estimator's active `cross` table.

For rubric transport, fit `fit_rubric_calibration()` with the labeled Phase A hub
artifact and `calibration_design = "linked_anchors"`. Prepare linking inputs with
that exact hub artifact, then pass the common result or session to `predict()`.
Global item IDs are required. Only spoke scores are returned. The frozen hub
calibration is reused, with the original hub mean restoring the reference origin;
no target labels, cohort rescaling or updated hub posterior refit the calibration.
Invalid or unidentified links fail. The prediction's `linking` attribute retains
estimator identity, uncertainty scope, evidence provenance and diagnostics.

Saved anchored-joint Phase B sessions are unsupported and fail with a restart
message. Compatible historical Phase A artifacts remain reusable. Adaptive
selection integration and removal of the remaining old execution code are
separate epic work; use the explicit-evidence APIs above for E1--E3 sessions.

## Citation

Mercer, S. H. (2026). Explicit-evidence linking sessions. In *pairwiseLLM*
[R package vignette]. Comprehensive R Archive Network.
<https://doi.org/10.32614/CRAN.package.pairwiseLLM>
