---
title: "Backends and Tested Model Configurations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backends and Tested Model Configurations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What the package supports

`pairwiseLLM` implements request and response handling for six backends. The
package generally forwards `model` identifiers to the selected provider; it
does not maintain an exhaustive model allowlist.

| Backend | Provider surface | Live | Batch | Credential |
|---|---|:---:|:---:|---|
| `openai` | OpenAI | yes | yes | `OPENAI_API_KEY` |
| `anthropic` | Anthropic | yes | yes | `ANTHROPIC_API_KEY` |
| `gemini` | Gemini Developer API | yes | yes | `GEMINI_API_KEY` |
| `vertex` | Vertex AI Gemini API | yes | no | `VERTEX_API_KEY` |
| `together` | Together AI | yes | no | `TOGETHER_API_KEY` |
| `ollama` | Local Ollama server | yes | no | none |

Backend implementation is not the same as compatibility with every model.
Endpoint compatibility depends on the request shape a model accepts;
maintainer testing covers only the exact configurations below; and current
availability is controlled by the provider. Preview identifiers and reasoning
controls can change independently of this package.

Check current availability in the official catalogs for
[OpenAI](https://developers.openai.com/api/docs/models),
[Anthropic](https://platform.claude.com/docs/en/about-claude/models/overview),
[Gemini Developer API](https://ai.google.dev/gemini-api/docs/models),
[Vertex AI](https://cloud.google.com/vertex-ai/generative-ai/docs/learn/models),
and [Together AI](https://docs.together.ai/docs/serverless-models).
Ollama tags depend on what is installed on the local server.

## Dated compatibility record

The machine-readable source is installed at
`inst/extdata/model_compatibility.csv`. Absence from this registry does not
imply incompatibility. A `true` value records a successful maintained smoke
test for that exact mode; a live success is not evidence of batch success.
The dated detailed result artifacts are
`inst/extdata/model_smoke_results_2026-09-03.csv` and
`inst/extdata/model_batch_smoke_results_2026-09-03.csv`. Standard configurations
omit `temperature` and `top_p`, so their model/provider defaults apply.

```{r registry}
registry_path <- system.file(
  "extdata",
  "model_compatibility.csv",
  package = "pairwiseLLM"
)
if (!nzchar(registry_path)) {
  source_paths <- c(
    file.path("inst", "extdata", "model_compatibility.csv"),
    file.path("..", "inst", "extdata", "model_compatibility.csv")
  )
  registry_path <- source_paths[file.exists(source_paths)][1]
}
registry <- utils::read.csv(registry_path, check.names = FALSE)
registry[c(
  "backend", "model_id", "endpoint", "live_tested", "batch_tested",
  "reasoning_mode", "package_version", "test_date", "status"
)]
```

Registry statuses mean:

- `tested-current`: the recorded configuration passed on the stated date and
  was current in the provider catalog when checked.
- `tested-legacy`: a successful dated test whose identifier is now legacy.
- `retired`: the provider has retired the identifier; it is retained only as
  historical evidence.
- `unverified`: availability or a partial response was observed, but no
  successful parsed package result was established.

The opt-in `inst/scripts/smoke_model_compatibility.R` harness can re-test
selected live or batch configurations and write a dated result artifact. It
makes billable network calls and is intentionally excluded from the package
test suite. For example:

```text
PAIRWISELLM_RUN_PROVIDER_SMOKE=true \
  Rscript inst/scripts/smoke_model_compatibility.R \
  --mode=live --providers=openai,anthropic
```

Selected providers must have their recorded API-key environment variables set;
otherwise the harness records `skipped-no-key` and exits unsuccessfully. Use
`--allow-missing-keys=true` only when those skipped rows are intentional.

Claude 5 reasoning is not advertised here: that family uses adaptive-thinking
semantics that this package does not currently implement and test. Likewise,
Ollama names such as `qwen3:32b`, `gemma3:27b`, and
`mistral-small3.2:24b` are only dated, environment-dependent local examples,
not claims about what another Ollama installation provides.
