---
title: "Guide: Rubric Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide: Rubric Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Rubric calibration converts completed comparative-judgment (CJ) results into
ordered performance levels. The ordinal methods learn the relationship between
CJ quality and human rubric scores. Percentile scoring matches a chosen marginal
distribution without learning rubric criteria.

## Choose the workflow

| Available evidence | Method and design | Interpretation |
|---|---|---|
| Completed CJ, no human rubric labels | `percentile`, `same_set` | Norm-referenced, distribution-matched levels |
| Human labels for a subset of one completed CJ analysis | `ordinal_linear`, `same_set` | Default two-stage calibration on that analysis's scale |
| A human-scored historical reference and new target responses | `ordinal_linear`, `linked_anchors` | Reusable calibration after existing Phase B linking |
| Evidence of curvature in the CJ-to-rubric relationship | Explicit `ordinal_monotone`, either ordinal design | Nondecreasing nonlinear calibration, assessed with held-out labels |

`same_set` means that labeled and unlabeled items participated in **one CJ fit**.
It is a two-stage procedure, not a joint CJ/rubric likelihood. Both ordinal
methods retain the common cumulative-effect assumption; a curved fit alone does
not resolve a proportional-odds violation. There is no automatic method selector.

## Start from completed CJ results

Use a completed `fit_bayes_btl_mcmc()` result, a completed within-set
`adaptive_rank()` result (or its state), or an import-ready Phase A artifact.
Raw score tables and intermediate adaptive states are rejected. Fixed fits with
multiple refits use the last refit. All four Bayesian variants, `btl`, `btl_e`,
`btl_b`, and `btl_e_b`, use the same calibration interface. For fixed-pair results,
live versus batch collection is provenance only.

Inspect CJ reliability, connectivity, and estimation diagnostics before
calibration. Upstream failures remain visible; calibration does not repair weak
CJ measurement. Import readiness follows existing Phase A quality rules and
does not by itself certify that the originating run stopped successfully.

For constructing these inputs, see [Standalone Bayesian BTL](bayesian-btl.html),
[Guide: Adaptive Pairing](adaptive-pairing.html), and
[Guide: Adaptive Linking](adaptive-linking.html). Rubric functions do not generate
pairs, collect judgments, or control adaptive stopping.

The code below starts from your completed objects and label tables. Those objects
are not bundled example data, so dependent chunks are not evaluated during normal
documentation builds. No provider calls or CmdStan sampling occur in the rubric
examples. The downstream snippets are tested with deterministic completed-result
fixtures in the package's test suite.

Version 1.5.0 requires **R >= 4.4**. Percentile scoring needs no optional modeling
backend. Linear fitting requires `ordinal`; saved linear prediction does not.
Monotone fitting requires `mgcv >= 1.9-4` and `withr`; saved monotone prediction
requires `mgcv`. Install the optional packages only when you use those methods:

```{r install, eval=FALSE}
install.packages("ordinal")
install.packages(c("mgcv", "withr"))
```

## Define the ordered rubric and labels

```{r rubric-levels}
rubric_levels <- c("developing", "proficient", "advanced")
```

`completed_cj` below is the completed organization-specific CJ result.
`training_labels` and `evaluation_labels` are disjoint data frames with columns
`item_id` and `rubric_score`, optionally `trait`. Each ID must occur in that CJ
result. Use global IDs when the adaptive result or artifact provides them;
otherwise use the original fixed-fit IDs. Alignment is by ID, never row position.
Omitted items or `NA` rubric scores are unlabeled.

Declare levels from lowest to highest quality. Numeric scores can infer ascending
levels, and ordered factors retain their declared levels. Character scores need
explicit `levels`. One fit has one category structure (`K >= 3`, commonly 3--6).
Every intended category must occur among training labels; missing categories
fail rather than being collapsed. Higher CJ and rubric scores must both mean
better performance.

Choose calibration responses that cover the CJ range, including its extremes,
and check category counts. There is no universal sufficient number of labels.
Keep evaluation labels separate before fitting: evaluation on training labels
measures apparent performance, not held-out accuracy.

## Same-set linear calibration

```{r same-set, eval=FALSE}
if (requireNamespace("ordinal", quietly = TRUE)) {
  calibration <- fit_rubric_calibration(
    completed_cj, training_labels,
    trait = "organization", levels = rubric_levels
  )
  scores <- predict(calibration)
  scores[, c("item_id", "rubric_score", "extrapolated")]
  calibration$category_counts
  calibration$calibration_range
  calibration$transformation
}
```

Only labeled training items determine the stored mean and sample SD used in
`z = (theta - center) / scale`. Linear calibration estimates
`logit P(Y <= k) = tau[k] - beta * z`. A positive slope means that higher CJ
quality favors higher rubric categories. Nonpositive slopes warn; the software
does not reverse coding. Sparse categories, separation, and numerical or
identification problems require review even if a fit is returned.

`predict(calibration)` scores every original item, including unlabeled items.
Explicit same-set `newdata` may only reuse the original completed result with
unchanged items, accepted scores/uncertainty, and fit evidence; reordering is
allowed. Independent cohorts and refits require a different deployment analysis.

## Read probabilities and hard scores

```{r probabilities, eval=FALSE}
if (requireNamespace("ordinal", quietly = TRUE)) {
  probability_matrix <- do.call(rbind, scores$probabilities)
  head(probability_matrix)
  scores[, c("median_category", "modal_category", "expected_level")]
  modal_scores <- predict(calibration, hard_score = "mode")
}
```

Each `probabilities` entry is a full K-vector named by the original ordered
labels. The default hard score is the **median category**: the first category
whose cumulative probability reaches 0.5. Equality selects the lower category.
The mode selects the largest probability; ties select the lowest category.
Changing `hard_score` changes `category` and `rubric_score`, while retaining all
probabilities and both decision summaries.

`expected_level` averages internal indices `1:K`, not arbitrary numeric label
values. It is a continuous summary, not a replacement for the ordinal outcome.
`extrapolated` marks scores outside the labeled training range, even among
original same-set items. Endpoints are inside the range.

## Evaluate held-out labels and inspect diagnostics

```{r evaluation, eval=FALSE}
if (requireNamespace("ordinal", quietly = TRUE)) {
  assessment <- evaluate_rubric_predictions(
    calibration, evaluation_labels, diagnostics = TRUE
  )
  assessment$metrics
  assessment$calibration
  assessment$metadata$n_training_label_overlap
  assessment$diagnostics$common_effect
}
```

Normalized ranked probability score (`rps`) is the primary probabilistic metric;
lower is better. It averages squared cumulative-probability errors over the
`K - 1` boundaries. Log loss is secondary. Hard-score summaries include exact and
within-one accuracy, mean absolute category error, and quadratic weighted kappa.
Distances use internal category indices. Undefined kappa and unavailable
probability metrics have explicit unavailable values/reasons.

Evaluation aligns IDs, excludes and counts missing labels, and reports training
overlap and extrapolation. Evaluation categories may be absent even though all
categories were required for training. `assessment$calibration` compares observed
and predicted cumulative frequencies in probability and theta bins; ties stay
together. These summaries can reveal where predictions disagree with labels.

Diagnostics are off by default. When requested, linear fits use a
threshold-varying alternative for a proportional-odds likelihood-ratio diagnostic.
Monotone fits provide descriptive boundary checks, not an omnibus test. Failed
diagnostic refits return an unavailable status and preserve the production fit.
Formal and descriptive checks use training data; held-out evaluation labels do
not refit the calibration or its training diagnostics.

For repeated label-only validation, keep completed CJ evidence fixed, hide
held-out rubric labels, and refit all calibration scaling and parameters in each
training fold. Ensure every fold's training data retain all intended categories.
The package's CV machinery is internal; the exported evaluator supports explicit
holdout workflows. Model/tuning selection using validation results needs a
separate outer assessment to avoid optimistic reported performance.

## Explicit monotone alternative

```{r monotone, eval=FALSE}
if (requireNamespace("mgcv", quietly = TRUE) &&
    packageVersion("mgcv") >= "1.9.4" && requireNamespace("withr", quietly = TRUE)) {
  monotone_calibration <- fit_rubric_calibration(
    completed_cj, training_labels, method = "ordinal_monotone",
    trait = "organization", levels = rubric_levels, k = 6
  )
  monotone_scores <- predict(monotone_calibration)
  monotone_assessment <- evaluate_rubric_predictions(
    monotone_calibration, evaluation_labels
  )
  monotone_assessment$metrics
}
```

This fits a nondecreasing penalized cubic spline using `mgcv::scasm()` and
`mgcv::ocat()`. The default basis dimension is `k = 6`; it must be at least five
and is reduced with a warning if there are fewer unique labeled scores. Fewer
than five unique scores fail. This is a backend feasibility constraint, not a
sample-size recommendation. Smoothing is estimated unless a positive fixed `sp`
is supplied. No unconstrained smoother replaces a failed monotone fit.

Inspect convergence, probability behavior, and held-out performance before
preferring the more flexible model. Numerical cutpoints can be unavailable for
flat or unresolved crossings; scoring still uses probabilities. Predictions
outside the labeled range retain spline extrapolation and are flagged, without
establishing rubric calibration there.

## Percentile levels without human labels

```{r percentile, eval=FALSE}
percentile_calibration <- fit_rubric_calibration(
  completed_cj, method = "percentile", trait = "organization",
  levels = rubric_levels, target_distribution = c(0.2, 0.5, 0.3)
)
percentile_scores <- predict(percentile_calibration)
percentile_calibration$backend$requested_proportions
percentile_calibration$backend$achieved_proportions
percentile_calibration$backend$cutpoints
percentile_calibration$backend$cutpoint_tie_counts
```

Omit `target_distribution` for equal proportions; `K` may replace explicit level
names. Quantiles use fixed R type 8. A score equal to an internal cutpoint enters
the higher category. All ties stay together, so requested proportions can be
unattainable and some categories can be empty. If all scores are identical,
all items enter the highest category under this boundary convention.

These are **norm-referenced, distribution-matched CJ-derived performance levels**.
Using a historical rubric distribution does not show individual agreement with
human rubric raters. Percentile fits support original-result reuse only, not
independent cohorts or `linked_anchors`. They return deterministic categories,
not category probabilities; evaluation supplies hard metrics only.

## Reuse a historical rubric reference through Phase B

The `rubric_reference_set` is the human-scored set whose CJ metric supports the
calibration. A Phase B **hub anchor** is a hub item selected for routing recurring
comparisons, not a known true rubric score.

1. Run trait-specific Phase A for the reference, retaining its ready artifact.
2. Fit and save its ordinal calibration and retain the original reference artifact.
3. Run compatible Phase A for each new target set.
4. Use the existing Phase B linking workflow with the reference as the fixed hub.
5. Apply the saved calibration to the completed accepted Phase B result.

In the [linking guide](adaptive-linking.html), ready artifacts are available in
`reference_run$phase_a$manifest[["1"]]` and
`target_run$phase_a$manifest[["2"]]` for reference set 1 and target set 2.
Extract the artifact list for fitting; the rubric fitter does not take a path or
the whole Phase A manifest. The following assumes `reference_phase_a` contains
that original artifact and `reference_labels` contains its human scores.

```{r linked-calibration, eval=FALSE}
if (requireNamespace("ordinal", quietly = TRUE)) {
  linked_calibration <- fit_rubric_calibration(
    reference_phase_a, reference_labels,
    calibration_design = "linked_anchors",
    trait = "organization", levels = rubric_levels
  )
  reference_scores <- predict(linked_calibration)
}
```

The reference and target Phase A contracts must match the selected BTL variant
and compatible settings. To collect the links, use `adaptive_rank()` as shown
below with your combined sample table, compatible `btl_config`, local or configured
judge, and planned comparison budget. `linking_samples` includes unique `ID` and
`global_item_id`, `text`, and `set_id` values 1 and 2. This upstream call is never
run during documentation builds. Keep production quality gates and resume the
session until completion; exhausting a budget alone does not establish precision.

```{r phase-b-workflow, eval=FALSE}
completed_phase_b <- adaptive_rank(
  data = linking_samples, id_col = "ID", text_col = "text",
  judge = judge, n_steps = comparison_budget, btl_config = btl_config,
  adaptive_config = list(
    run_mode = "link_one_spoke", hub_id = 1L, phase_a_mode = "import",
    phase_a_artifacts = list(`1` = reference_phase_a, `2` = target_phase_a)
  ),
  session_dir = tempfile("rubric-linking-"), seed = 201L, progress = "none"
)
```

For several targets, use the existing `link_multi_spoke` workflow. Phase B
hard-locks hub locations and retains its shared judge parameters, D-optimal
selection, probes, and stopping. Rubric labels are not required during linking.

```{r linked-prediction, eval=FALSE}
if (requireNamespace("ordinal", quietly = TRUE)) {
  target_scores <- predict(linked_calibration, newdata = completed_phase_b)
  target_scores[, c("global_item_id", "set_id", "rubric_score", "theta_sd", "extrapolated")]
  attr(target_scores, "linking")$diagnostics
}
```

Explicit linked prediction returns targets only; `newdata = NULL` predicts the
original reference items. Every spoke needs an accepted refit and committed
active linking evidence. The accepted public `theta_link_eap` field retains its
historical name: Phase B locations are MAP estimates with Laplace/Hessian
uncertainty, not MCMC posterior means. Validated per-spoke provenance remains
available, including accepted-state reuse.

Prediction validates the original reference IDs, locations/uncertainty, evidence,
trait, orientation, and canonical fit contract. A matching configuration hash
alone does not establish a shared metric. Target-only Phase A scores, changed
reference scales, and incompatible contracts fail explicitly. **Never independently
re-standardize the target cohort**: prediction uses the stored labeled-reference
mean and SD on accepted Phase B common-scale locations.

Reference holdout performance assesses calibration on the reference scale. To
assess transport, use external human labels for linked targets:

```{r linked-evaluation, eval=FALSE}
if (requireNamespace("ordinal", quietly = TRUE)) {
  target_assessment <- evaluate_rubric_predictions(
    linked_calibration, target_labels, newdata = completed_phase_b
  )
  target_assessment$metrics
  target_assessment$metadata$linking
}
```

Phase B diagnostics alone do not demonstrate rubric accuracy, and reference
validation alone does not measure the additional error from linking.

## Separate analytic traits

Organization and evidence require separate trait-specific comparisons, CJ fits,
labels, and calibrations even when item IDs overlap. Each may have different
levels or a different Bayesian variant. Compatibility must hold within each
trait's reference/target linking workflow. One holistic CJ scale cannot supply
multiple analytic trait calibrations.

```{r analytic-traits, eval=FALSE}
if (requireNamespace("ordinal", quietly = TRUE)) {
  evidence_calibration <- fit_rubric_calibration(
    evidence_cj, evidence_training_labels, trait = "evidence",
    levels = c("limited", "adequate", "strong")
  )
  evidence_scores <- predict(evidence_calibration)
}
```

## Uncertainty and reporting

All three methods condition on accepted **CJ point locations**. Retained CJ SDs,
posterior draws, or linking uncertainty are not propagated into rubric category
probabilities. Ordinal parameter uncertainty is conditional on those estimated
predictors; monotone threshold uncertainty is not supplied by the wrapper.
There is no posterior/bootstrap propagation engine or joint likelihood here.

Report the trait, BTL variant, method/design, ordered levels, labeled category
counts and range, held-out metrics, extrapolation, and CJ/calibration/linking
diagnostics. Distinguish numerical feasibility from adequate measurement and
external rubric validity. Small deterministic examples and package recovery
tests do not establish a universal calibration sample size or validity for a new
population.

## Citation

> Mercer, S. H. (2026). *Guide: Rubric Calibration* [R package vignette]. Comprehensive R Archive Network. https://doi.org/10.32614/CRAN.package.pairwiseLLM
